---
title: "EGFR Load Impact"
format: html
---

```{r setup}
library(tidyverse)
library(patchwork)
library(glue)
library(ComplexHeatmap)
library(ggh4x)
library(ggthemes)
library(here)

devtools::load_all()
```


## Introduction

EGFR IMPACT plotting. 

TODO:

* Grab TMB, MSI, and other sample level metadata to plot in a facet
* Metastatic sites
* Mutation spectra or signatures?
* Additional non snRNA-seq samples

## Load metadata

We keep `dmp_pid`, `dmp_tumor`, and `dmp_sample` consistent across all tables for easy lookups and merging

```{r}
# metadata <- load_metadata(nucseq_excel = here("metadata/Nucseq_Shah_Lab_2023_5_24.xlsx"))
metadata <- read.table(here("metadata", "metadata_update_mar_2025.txt"), header = T, sep = "\t", quote = "\"")
markers <- load_markers()


# This is to be run on Juno where a facets directory exists
# facets_data <- load_facets(facets_path = "/rtsess01/compute/juno/cmo/juno/work/ccs/shared/resources/impact/facets/",
#                            dmp_ids = metadata$closest_impact_to_sample,
#                            here("results", "facets_data.txt"))

impact_data <- load_impact(
  impact_dir = "/data1/shahs3/users/zatzmanm/work/repos/shared/msk-impact/",
  facets_data = here("results", "facets_data_dec_2024.txt"),
  dmp_ids = metadata$closest_impact_to_sample,
  save_to = here("results", "impact_data.rds")
)

# oncokb_dat <- load_oncokb(oncokb_dir = "/data1/shahs3/users/zatzmanm/work/repos/shared/oncokb-annotated-msk-impact", dmp_ids = metadata$closest_impact_to_sample)
```


# ONCOPRINT

Metadata

```{r}
cmap <- load_colors()

onco_meta <- impact_data$clin_dat_sample %>%
  inner_join(metadata, by = c("dmp_sample" = "closest_impact_to_sample")) %>%
  left_join(impact_data$facets_data)

rownames(onco_meta) <- onco_meta$sample_id_old

onco_meta$time_point <- factor(onco_meta$time_point, levels = c("TN", "MRD", "PD"))

onco_dat <- inner_join(impact_data$onco_dat_full, onco_meta)
```


Save out table for supplemental

```{r}
g_order <- names(table(onco_dat$Gene) %>% sort(decreasing = T))
onco_df_table <- onco_dat %>%
  select(sample_id_new, time_point, source:mutation_type_simple) %>%
  dplyr::relocate(mutation_type_simple, .after = mutation_type) %>%
  janitor::clean_names() %>%
  mutate(
    sample_id_new = fct_relevel(sample_id_new, names(colors$sample_id_new)),
    time_point = fct_relevel(time_point, names(colors$time_point)),
    gene = fct_relevel(gene, g_order)
  ) %>%
  dplyr::rename(sample_id = sample_id_new) %>%
  arrange((time_point), (sample_id), gene)
onco_df_table %>%
  write.table(file = here("tables", "impact_mutation_data.txt"), quote = F, sep = "\t", row.names = F, col.names = T)
```


Prepare for plotting

```{r, fig.width=11, fig.height=10}
# Add sensible patient IDs
# Add in patients that are missing (keep empty)
# Fix colors and rect shapes
# Sample and patient metadata
# Sample metadata

library(ComplexHeatmap)

comut_mat <- onco_dat %>%
  filter(dmp_sample %in% onco_meta$dmp_sample) %>%
  pivot_wider(id_cols = sample_id_new, names_from = Gene, values_from = mutation_type_simple, values_fn = function(x) paste0(x, collapse = ";"), values_fill = "") %>%
  column_to_rownames("sample_id_new") %>%
  as.matrix() %>%
  t()

#### Add the empty samples
# samps_add <- setdiff(onco_meta$dmp_tumor, colnames(comut_mat))
# samps_add_mat <- matrix(data = "", nrow = nrow(comut_mat), ncol = length(samps_add), dimnames = list(rownames(comut_mat), samps_add))
# comut_mat_final <- cbind(comut_mat, samps_add_mat)
#####

comut_mat_final <- comut_mat

# Reorder metadata
rownames(onco_meta) <- onco_meta$sample_id_new
onco_meta <- onco_meta[colnames(comut_mat_final), ]

mut_pal <- c(
  "Missense Mutation" = "#1C91D4",
  "Nonsense Mutation" = "#D5C711",
  "Deep Deletion" = "blue",
  "Amplification" = "red",
  "Frameshift Indel" = "#D55E00",
  "Inframe Indel" = "#CC79A7",
  "RNA Fusion" = "#F0E442",
  "Structural Variant" = "#56B4E9",
  "Other" = "#007756"
)

cli::cli_alert_danger("Missing {unique(onco_dat$mutation_type_simple)[!unique(onco_dat$mutation_type_simple) %in% names(mut_pal)]} mutations")
```


```{r, fig.width=11, fig.height=10}
alter_fun <- list(
  background = alter_graphic("rect", fill = "#CCCCCC"),
  "Amplification" = alter_graphic("rect", fill = mut_pal["Amplification"]),
  "Deep Deletion" = alter_graphic("rect", fill = mut_pal["Deep Deletion"]),
  "Missense Mutation" = alter_graphic("rect", height = 0.33, fill = mut_pal["Missense Mutation"]),
  "Nonsense Mutation" = alter_graphic("rect", height = 0.33, fill = mut_pal["Nonsense Mutation"]),
  "Frameshift Indel" = alter_graphic("rect", height = 0.33, fill = mut_pal["Frameshift Indel"]),
  "Inframe Indel" = alter_graphic("rect", height = 0.33, fill = mut_pal["Inframe Indel"]),
  "RNA Fusion" = alter_graphic("rect", height = 1, width = 0.33, fill = mut_pal["RNA Fusion"]),
  "Structural Variant" = alter_graphic("rect", height = 0.33, fill = mut_pal["Structural Variant"]),
  "Other" = alter_graphic("rect", height = 0.33, fill = mut_pal["Other"])
)

# tp_names <- levels(factor(onco_meta$timepoint))
# tp_cols <- set_names(circlize::colorRamp2(c(0, 5), c("lightblue", "darkblue"))(1:length(tp_names)), tp_names)
#
# pt_cols <- set_names(ggthemes::ptol_pal()(length(unique(onco_meta$patient))), sort(unique(onco_meta$patient)))


anno_cols <- list(
  "timepoint" = cmap$time_point,
  "progressed" = cmap$tf,
  "WGD" = cmap$tf,
  "resistance" = cmap$mech_resistance,
  "FGA" = circlize::colorRamp2(range(onco_meta$fga, na.rm = T), hcl_palette = "SunsetDark", reverse = TRUE),
  "age" = circlize::colorRamp2(range(onco_meta$age_at_intial_diagnosis, na.rm = T), hcl_palette = "SunsetDark", reverse = TRUE)
)

# anno_labels <- c("Mutation Type", names(anno_cols))
# anno_labels[anno_labels == "Age"] <- "Age (months)"
# anno_labels[anno_labels == "WBC"] <- "WBC (10^9/l)"

tmb_breaks <- c(1, 10, 100)


top_anno <- HeatmapAnnotation(
  na_col = "white",
  annotation_name_side = "left",
  annotation_label = c(
    "TMB",
    "FGA",
    "WGD",
    # "Ever progressed?",
    "Age",
    # "Panel type",
    # "Mech. of resistance",
    "Time point"
  ),
  # cbar = anno_oncoprint_barplot(),
  TMB = anno_barplot(
    x = log10(onco_meta$CVR_TMB_SCORE + 1),
    axis_param = list(
      at = log10(tmb_breaks + 1),
      labels = tmb_breaks,
      side = "left"
    ),
    ylim = c(log10(1), log10(100)),
    border = FALSE,
    baseline = 0
  ),
  FGA = onco_meta$fga,
  WGD = onco_meta$wgd,
  # timepoint = onco_meta$time_point,
  # progressed = onco_meta$progressed,
  age = onco_meta$age_at_intial_diagnosis,
  # assay_type = onco_meta$GENE_PANEL,
  # resistance = onco_meta$mech_resistance,
  timepoint = anno_block(
    gp = gpar(fill = cmap$time_point),
    labels = c("Treatment Naive", "MRD", "Progressive Disease"),
    labels_gp = gpar(col = "white", fontsize = 11)
  ),

  # timepoint = onco_meta$timepoint,
  # response = onco_meta$response_category,
  # disease_status = onco_meta$disease_status,
  # chemo = onco_meta$chemo,
  # transplant = onco_meta$transplant,
  # t8_21 = onco_meta$t8_21,
  col = anno_cols
)
```


Plot the plot

```{r}
min_mut_freq <- 0.05

# Get mutation frequencies
mut_freq <- (apply(comut_mat_final, MARGIN = 1, function(x) sum(x != "")) / ncol(comut_mat_final)) %>%
  sort(decreasing = T)

plot_genes <- names(which(mut_freq >= min_mut_freq))
```

Plot and save

```{r, fig.width=15, fig.height=12}
ht <- oncoPrint(comut_mat_final[plot_genes, ],
  right_annotation = rowAnnotation(
    row_barplot = anno_oncoprint_barplot(show_fraction = T, border = F)
  ),
  top_annotation = top_anno,
  column_split = onco_meta$time_point,
  # row_names_side = "left",
  column_title = NULL,
  alter_fun = alter_fun,
  col = mut_pal,
  show_column_names = TRUE,
  remove_empty_columns = FALSE,
  # pct_side = "right",
  column_title_rot = 0,
  border = TRUE
)

pdf(file = here("plots", "oncoprint.pdf"), width = 15, height = 12)
draw(ht)
dev.off()

suppressWarnings(ht)
```

Average TMB per group

```{r, fig.width = 11, fig.height = 3}
my_comparisons <- list(c("TN", "MRD"), c("TN", "PD"), c("MRD", "PD"))

# stat_comp <- ggpubr::stat_compare_means(method = "t.test", comparisons = my_comparisons)
#
mean_comp <- ggpubr::stat_compare_means(comparisons = my_comparisons, label = "p.signif", tip.length = 0, vjust = 0, size = 3)


p1 <- onco_meta %>%
  ggplot(aes(x = time_point, y = CVR_TMB_SCORE)) +
  ggbeeswarm::geom_quasirandom(aes(color = time_point)) +
  geom_boxplot(fill = NA, outlier.shape = NA, aes(color = time_point)) +
  scale_y_log10(breaks = 10^(0:3)) +
  theme_classic() +
  mean_comp +
  # stat_comp +
  labs(x = "Timepoint", y = "TMB (mut/Mb)") +
  scale_color_manual(values = anno_cols$timepoint) +
  coord_cartesian(clip = "off", ylim = c(0.5, 100)) +
  theme(legend.position = "none", plot.margin = margin(t = 2, unit = "lines"))

p2 <- onco_meta %>%
  ggplot(aes(x = time_point, y = fga)) +
  ggbeeswarm::geom_quasirandom(aes(color = time_point)) +
  geom_boxplot(fill = NA, outlier.shape = NA, aes(color = time_point)) +
  # scale_y_log10(breaks = 10^(0:3), limits = c(min(onco_meta$CVR_TMB_SCORE), 100)) +
  theme_classic() +
  mean_comp +
  labs(x = "Timepoint", y = "Fraction genome altered") +
  scale_color_manual(values = anno_cols$timepoint) +
  coord_cartesian(clip = "off", ylim = c(0, 1)) +
  theme(legend.position = "none")

p3 <- onco_meta %>%
  ggplot(aes(x = time_point, y = ploidy)) +
  ggbeeswarm::geom_quasirandom(aes(color = time_point)) +
  geom_boxplot(fill = NA, outlier.shape = NA, aes(color = time_point)) +
  # scale_y_continuous(trans = "log2") +
  theme_classic() +
  mean_comp +
  labs(x = "Timepoint", y = "Ploidy") +
  scale_color_manual(values = anno_cols$timepoint) +
  coord_cartesian(clip = "off", ylim = c(1, 3.7)) +
  theme(legend.position = "none")

p4 <- onco_meta %>%
  ggplot(aes(x = time_point, y = frac_loh)) +
  ggbeeswarm::geom_quasirandom(aes(color = time_point)) +
  geom_boxplot(fill = NA, outlier.shape = NA, aes(color = time_point)) +
  # scale_y_log10(breaks = 10^(0:3), limits = c(min(onco_meta$CVR_TMB_SCORE), 100)) +
  theme_classic() +
  mean_comp +
  labs(x = "Timepoint", y = "Fraction genome LOH") +
  coord_cartesian(clip = "off", ylim = c(0, 1)) +
  scale_color_manual(values = anno_cols$timepoint) +
  theme(legend.position = "none")


p5 <- onco_meta %>%
  dplyr::count(time_point, wgd) %>%
  dplyr::filter(!is.na(wgd)) %>%
  group_by(time_point) %>%
  add_tally(n) %>%
  mutate(wgd_frac = n / nn) %>%
  filter(wgd == T) %>%
  ggplot(aes(x = time_point, y = wgd_frac)) +
  geom_col(aes(fill = time_point)) +
  labs(y = "Proportion of samples with WGD", fill = "Timepoint") +
  scale_fill_manual(values = anno_cols$timepoint) +
  theme_classic()

pcomb <- (p1 | p2 | p3 | p4 | p5) + plot_layout(guides = "collect") & guides(x = guide_axis(angle = 0)) & labs(x = NULL) &
  theme(plot.margin = margin(t = 1.5, 0.25, 0.25, 0.25, unit = "lines"))
ggsave(pcomb, filename = here("plots", "impact_wgd.pdf"), width = 11, height = 3)
pcomb
```


## Enrichment in progressed vs pre

```{r, fig.width=2.5, fig.height=1.5}
x <- (comut_mat_final != "") %>%
  as.data.frame() %>%
  rownames_to_column(var = "Gene") %>%
  pivot_longer(cols = !Gene, names_to = "sample_id_new") %>%
  left_join(metadata[, c("sample_id_new", "time_point")]) %>%
  group_by(Gene, time_point) %>%
  mutate(
    n_gene = sum(value),
    n_tp = n_distinct(sample_id_new)
  ) %>%
  distinct(Gene, time_point, n_gene, n_tp) %>%
  mutate(n_absent = n_tp - n_gene) %>%
  filter(time_point != "MRD") %>%
  pivot_wider(id_cols = "Gene", names_from = c(time_point), values_from = c("n_gene", "n_absent"))

enrich_res <- map(c("TP53", "RB1"), .f = function(gene) {
  vals <- x[x$Gene == gene, 2:5]

  mat <- matrix(c(vals[[1]], vals[[3]], vals[[2]], vals[[4]]),
    nrow = 2,
    dimnames = list(
      PD = c("Mut", "Wt"),
      Pre = c("Mut", "Wt")
    )
  )

  res <- stats::fisher.test(mat) %>%
    broom::tidy() %>%
    mutate(Gene = gene)
  return(res)
}) %>%
  list_rbind() %>%
  arrange(p.value) %>%
  rstatix::adjust_pvalue(method = "BH")


p <- enrich_res %>%
  ggplot(aes(x = estimate, y = Gene)) +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high)) +
  geom_text(aes(label = glue("{scales::label_pvalue(accuracy = 0.0001, add_p=T)(p.value)}")),
    nudge_y = 0.35, size = 3.5
  ) +
  theme_classic(base_size = 16) +
  scale_x_continuous(trans = "log2") +
  labs(x = "Odds Ratio") +
  geom_vline(xintercept = 1, linetype = "22") +
  theme(panel.border = element_rect(fill = NA), axis.line = element_blank()) +
  theme(plot.margin = margin(r = 1, 0.25, 0.25, 0.25, "lines"))

ggsave(p, file = here("plots", "egfr_prog_enrich.pdf"), width = 2.5, height = 1.5)


p
```


## Reduced oncoprint for main

```{r}
onco_signals <- c("Cell Cycle", "HIPPO", "MYC", "NOTCH", "NRF2", "PI3K", "TGF-Beta", "RTK RAS", "TP53", "WNT")
res <- map(onco_signals, .f = function(x) {
  sr <- 1
  if (x == "Cell Cycle") {
    sr <- 3
  }
  df <- openxlsx::read.xlsx(xlsxFile = here("impact", "mmc3.xlsx"), sheet = x, check.names = F, colNames = T, rowNames = F, startRow = sr) %>%
    dplyr::mutate("Pathway" = x, .before = 1) %>%
    janitor::clean_names()
  return(df)
})

genelist <- map2(res, onco_signals, \(x, y) {
  data.frame(Pathway = y, Gene = x$gene)
}) %>%
  list_rbind()

resistance_genes <- data.frame("Pathway_resistance" = "Resistance Gene", Gene = markers$resistance_genes)

impact_groupings <- read.table(file = here("impact", "impact_pathway_groupings.txt"), header = T, sep = "\t")

impact_2 <- read.table(file = here("impact", "impact_annos_sarcoma.txt"), header = T, sep = "\t")

# Merge them all

impact_list <- genelist %>%
  left_join(y = impact_groupings, by = c("Gene" = "Hugo_symbol"), suffix = c("_cell", "_list1")) %>%
  left_join(y = impact_2, by = c("Gene" = "Hugo.symbol")) %>%
  dplyr::rename("Pathway_sarcoma" = "Pathway") %>%
  dplyr::relocate("Gene") %>%
  mutate(
    in_egfr = Gene %in% impact_data$onco_dat_full$Gene,
    Pathway_cell = janitor::make_clean_names(Pathway_cell, allow_dupes = T),
    Pathway_list1 = janitor::make_clean_names(Pathway_list1, allow_dupes = T),
    Pathway_sarcoma = janitor::make_clean_names(Pathway_sarcoma, allow_dupes = T),
  ) %>%
  add_count(Gene)

impact_list$pathway <- impact_list$Pathway_sarcoma
impact_list$pathway[impact_list$pathway == "na" & impact_list$Pathway_cell != "na"] <- impact_list$Pathway_cell[impact_list$pathway == "na" & impact_list$Pathway_cell != "na"]

impact_list$pathway[impact_list$pathway == "na" & impact_list$Pathway_list1 != "na"] <- impact_list$Pathway_list1[impact_list$pathway == "na" & impact_list$Pathway_list1 != "na"]

# if n = 2 take pathway_cell or
impact_list$pathway[impact_list$n == 2] <- impact_list$Pathway_cell[impact_list$n == 2]

impact_df <- impact_list %>%
  distinct(Gene, pathway, in_egfr) %>%
  filter(!(Gene == "SCRIB" & pathway == "rtk_ras"))

rownames(impact_df) <- impact_df$Gene

impact_df$resistance_gene <- impact_df$Gene %in% resistance_genes$Gene

write.table(x = impact_df, file = here("impact", "impact_mapping.txt"), quote = F, sep = "\t", row.names = F, col.names = T)
```

# OLD
### EGFR hits


```{r}
memoSort <- function(M) {
  geneOrder <- sort(rowSums(M), decreasing = TRUE, index.return = TRUE)$ix
  scoreCol <- function(x) {
    score <- 0
    for (i in 1:length(x)) {
      if (x[i]) {
        score <- score + 2^(length(x) - i)
      }
    }
    return(score)
  }
  scores <- apply(M[geneOrder, ], 2, scoreCol)
  sampleOrder <- sort(scores, decreasing = TRUE, index.return = TRUE)$ix
  return(M[geneOrder, sampleOrder])
}
```

Determine denovo vs acquired alterations

```{r, fig.width = 9, fig.height = 5}
egfr_hits <- oncokb_dat$oncokb_summary_table %>%
  filter(gene == "EGFR") %>%
  left_join(metadata, by = c("dmp_sample" = "closest_impact_to_sample")) %>%
  mutate(time_point = factor(time_point, levels = c("Pre", "On", "PD")))

egfr_hits <- egfr_hits %>%
  mutate(
    variant = gsub("^p\\.", "", variant),
    egfr_cat = case_when(
      grepl("*74.*[ins|del]|exon 19 del|T751_I759", variant) ~ "Exon 19",
      # grepl("S768_V", variant) ~ "Exon 20",
      # grepl("*74.*del|exon 19 del", variant) ~ "Exon 19 Deletion",
      grepl("G719*", variant) ~ "G719X",
      variant == "L858R" ~ "L858R",
      variant == "T790M" ~ "T790M",
      # variant == "L861Q" ~ "L861Q",
      # variant == "S768I" ~ "S768I",
      variant == "C797S" ~ "C797S",
      variant == "Amplification" ~ "Amplification",
      .default = "Other"
    ), .after = variant_type
  )

egfr_hits$denovo <- map2(egfr_hits$variant, egfr_hits$de_novo_mutation_egfr_specifics, .f = function(x, y) {
  ifelse(str_detect(pattern = x, string = y), "De novo", "Acquired")
}) %>%
  unlist() %>%
  factor(levels = c("De novo", "Acquired"))
# Pre should all be de novo
egfr_hits$denovo[egfr_hits$time_point == "Pre"] <- "De novo"
```


```{r, fig.width = 9, fig.height = 5}
egfr_mat <- egfr_hits %>%
  mutate(value = 1) %>%
  pivot_wider(id_cols = c("sample_id_old"), names_from = "variant", values_from = value, values_fill = 0) %>%
  column_to_rownames("sample_id_old") %>%
  t() %>%
  memoSort()

# egfr_hits %>%
#   ggplot(aes(x = fct_relevel(sample_id_old, colnames(egfr_mat)), y = fct_rev(fct_relevel(variant, rownames(egfr_mat))))) +
#   geom_tile() +
#   facet_grid(.~time_point, space = "free", scales = "free") +
#   guides(x = guide_axis(angle = 90)) +
#   labs(x = "Sample ID", y = "EGFR Variant") +
#   ggthemes::theme_pander() +
#   theme(panel.border = element_rect(fill = NA, color = "black"))

egfr_hits %>%
  ggplot(aes(x = fct_relevel(sample_id_old, colnames(egfr_mat)), y = fct_rev(fct_relevel(variant, rownames(egfr_mat))))) +
  geom_tile() +
  facet_grid(denovo ~ time_point, space = "free", scales = "free") +
  guides(x = guide_axis(angle = 90)) +
  labs(x = "Sample ID", y = "EGFR Variant") +
  theme_pander() +
  theme(panel.border = element_rect(fill = NA, color = "black"), strip.clip = "off")
```

Final EGFR plot

```{r, fig.width=9, fig.height = 3}
egfr_comb_mat <- egfr_hits %>%
  mutate(value = 1) %>%
  pivot_wider(id_cols = c("sample_id_old"), names_from = c("denovo", "egfr_cat"), values_from = value, values_fill = 0, values_fn = sum) %>%
  column_to_rownames("sample_id_old") %>%
  t() %>%
  memoSort()

egfr_hits <- egfr_hits %>%
  mutate(type_name = paste(denovo, egfr_cat, sep = "_"))


p1 <- egfr_hits %>%
  ggplot(aes(x = fct_relevel(sample_id_old, colnames(egfr_comb_mat)), y = fct_rev(fct_relevel(type_name, rownames(egfr_comb_mat))))) +
  geom_tile() +
  facet_nested(denovo ~ time_point, space = "free", scales = "free", switch = "y") +
  scale_y_discrete(breaks = rownames(egfr_comb_mat), labels = gsub("De novo_|Acquired_", "", rownames(egfr_comb_mat))) +
  guides(x = guide_axis(angle = 90)) +
  labs(x = "Sample ID", y = "EGFR Variant") +
  theme_pander(base_size = 10) +
  theme(panel.border = element_rect(fill = NA, color = "black"), strip.clip = "off", strip.text.y.left = element_text(angle = 90, hjust = 0.5, size = 8), strip.placement = "outside", axis.text.x = element_text(size = 6), panel.grid = element_line(linetype = "22", linewidth = 0.25, color = "grey85"), strip.background.y = element_part_rect(side = "r", color = "black", fill = NA))
# p1


p2 <- egfr_hits %>%
  # add_count(type_name) %>%
  # distinct(type_name, type, n, time_point) %>%
  ggplot(aes(y = fct_rev(fct_relevel(type_name, rownames(egfr_comb_mat))))) +
  # theme_classic() +
  geom_bar(aes(fill = time_point)) +
  scale_y_discrete(breaks = rownames(egfr_comb_mat), labels = gsub("De novo_|Acquired_", "", rownames(egfr_comb_mat))) +
  scale_x_continuous(position = "top") +
  facet_grid(denovo ~ ., space = "free", scales = "free", switch = "y") +
  scale_fill_manual(values = cmap$time_point) +
  guides(
    fill = guide_legend(keywidth = 0.4, keyheight = 0.4, position = "inside"),
    x = guide_axis(cap = "both"),
    y = guide_axis(cap = "both")
  ) +
  theme_classic() +
  theme(plot.margin = margin(unit = "lines"), strip.text = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank(), legend.position.inside = c(1.1, 0.5), legend.background = element_blank()) +
  labs(x = "Count", y = NULL, fill = "Timepoint")

# Instead do as proportion of the timepoints
df_egfr_cat_prop <- egfr_hits %>%
  add_count(time_point, type_name, name = "n_type") %>%
  complete(time_point, type_name, fill = list(n_type = 0)) %>%
  add_count(time_point, name = "n_tp") %>%
  mutate(prop = n_type / n_tp) %>%
  distinct(time_point, type_name, time_point, n_type, n_tp, prop, denovo) %>%
  mutate(denovo = factor(ifelse(grepl("De novo", type_name), "De novo", "Acquired"),
    levels = c("De novo", "Acquired")
  ))
# %>%
# filter()


p2 <- df_egfr_cat_prop %>%
  ggplot(aes(y = fct_rev(fct_relevel(type_name, rownames(egfr_comb_mat))))) +
  geom_col(aes(fill = time_point, x = prop), position = "dodge") +
  scale_y_discrete(breaks = rownames(egfr_comb_mat), labels = gsub("De novo_|Acquired_", "", rownames(egfr_comb_mat)), position = "right") +
  scale_x_continuous(position = "top") +
  facet_grid(denovo ~ ., space = "free", scales = "free") +
  scale_fill_manual(values = cmap$time_point) +
  guides(
    fill = guide_legend(keywidth = 0.4, keyheight = 0.4, position = "inside"),
    x = guide_axis(cap = "both"),
    y = guide_axis(cap = "both")
  ) +
  theme_classic() +
  theme(plot.margin = margin(unit = "lines"), strip.text = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank(), legend.position.inside = c(0.8, 0.5), legend.background = element_blank()) +
  labs(x = "Proportion", y = NULL, fill = "Timepoint")

pegfr <- p1 + (p2 + theme(plot.margin = margin(r = 3, unit = "lines"))) + plot_layout(ncol = 2, widths = c(1, 0.125)) & theme(panel.spacing = unit(0.2, "lines"))
pegfr
ggsave(pegfr, file = here("plots", "egfr_mut_plot.pdf"), width = 9, height = 3)
```


Cross check both

```{r}
df_check <- egfr_hits %>%
  mutate(variant = gsub("^p\\.", "", variant)) %>%
  pivot_wider(id_cols = "sample_id_old", names_from = "gene", values_from = "variant", values_fn = function(x) {
    paste(x, collapse = ";")
  }) %>%
  dplyr::rename("EGFR_impact" = "EGFR") %>%
  full_join(metadata[, c("sample_id_old", "impact_for_sample", "closest_impact_to_sample", "de_novo_mutation_egfr_specifics", "secondary_egfr_genomic", "has_impact", "mech_of_resistance_to_osimertinib", "time_point")]) %>%
  arrange(time_point, has_impact, sample_id_old)
view(df_check)
```


```{r}
pid_check <- "P-0072023"
impact_data$onco_dat_full %>%
  filter(Gene == "EGFR", dmp_pid == pid_check) %>%
  select(dmp_sample, Gene, mutation_type, HGVSp_Short)

oncokb_dat$oncokb_summary_table %>%
  filter(gene == "EGFR", dmp_pid == pid_check)

oncokb_dat$oncokb_muts %>%
  filter(hugo_symbol == "EGFR", dmp_pid == pid_check) %>%
  select(hugo_symbol, hgv_sp_short)

oncokb_dat$oncokb_clindat %>%
  filter(dmp_pid == pid_check)

# Check the cbio mutation files directory
chk <- data.table::fread("/data1/shahs3/users/zatzmanm/work/repos/shared/cbio_mutations/data_clinical_sample.oncokb.txt") %>%
  as.data.frame() %>%
  janitor::clean_names() %>%
  filter(patient_id == pid_check)
```


## Full IMPACT

Patients have multiple IMPACT samples. Here we want to summarize the full level of IMPACT sample collection for each patient. considering the types of samples (primary, metastatic, sites), assays, and timepoints available. Can also plot things like TMB over time etc...

```{r}
onco_meta_full <- clin_dat_sample %>%
  left_join(metadata, by = c("dmp_pid"), suffix = c("_imp", "_egfr")) %>%
  mutate(
    dmp_tumor = dmp_tumor_imp,
    dmp_sample = dmp_sample_imp,
    dmp_egfr_match = ifelse(dmp_sample == dmp_sample_egfr, T, F)
  )

rownames(onco_meta_full) <- onco_meta_full$dmp_sample

onco_meta_full$time_point <- factor(onco_meta_full$time_point, levels = c("Pre", "On", "Progression"))

onco_dat_full <- inner_join(onco_dat_full, onco_meta_full)
```

Take a look at one patient just to see what it looks like

```{r}
meta_cols <- c(
  "CANCER_TYPE",
  # "CANCER_TYPE_DETAILED",
  # "ONCOTREE_CODE",
  "PRIMARY_SITE",
  "METASTATIC_SITE",
  "SAMPLE_TYPE",
  "SAMPLE_CLASS",
  # "GENE_PANEL",
  # "MSI_TYPE",
  # "CVR_TMB_SCORE",
  "dmp_egfr_match",
  "time_point"
)

onco_meta_full <- onco_meta_full %>%
  mutate(
    PRIMARY_SITE = str_to_title(PRIMARY_SITE),
    METASTATIC_SITE = str_to_title(METASTATIC_SITE)
  ) %>%
  add_count(dmp_pid) # %>%
filter(n > 1)

anno_cols$time_point <- anno_cols$timepoint

anno_cols$dmp_egfr_match <- anno_cols$progressed

comut_mat_full <- onco_dat_full %>%
  filter(dmp_pid %in% onco_meta_full$dmp_pid) %>%
  pivot_wider(id_cols = dmp_sample, names_from = Gene, values_from = mutation_type_simple, values_fn = function(x) paste0(x, collapse = ";"), values_fill = "") %>%
  column_to_rownames("dmp_sample") %>%
  as.matrix() %>%
  t()


# Subset
onco_meta_full <- onco_meta_full[colnames(comut_mat_full), ]

onco_meta_full$SAMPLE_TYPE <- factor(onco_meta_full$SAMPLE_TYPE, levels = c("Primary", "Local Recurrence", "Metastasis", "Unknown"))

# Reorder both
onco_meta_full <- onco_meta_full[order(onco_meta_full$time_point, onco_meta_full$SAMPLE_TYPE), ]

onco_meta_full$sample_id_old <- factor(onco_meta_full$sample_id_old, levels = unique(onco_meta_full$sample_id_old))

comut_mat_full <- comut_mat_full[, rownames(onco_meta_full)]

ha <- HeatmapAnnotation(df = onco_meta_full[, meta_cols], col = anno_cols)

ht_full <- oncoPrint(comut_mat_full[plot_genes, rownames(onco_meta_full)],
  right_annotation = rowAnnotation(
    row_barplot = anno_oncoprint_barplot(show_fraction = F, border = F)
  ),
  top_annotation = ha,
  column_split = factor(onco_meta_full$sample_id_old),
  # row_names_side = "left",
  # column_title = NULL,
  alter_fun = alter_fun,
  col = mut_pal,
  show_column_names = TRUE,
  remove_empty_columns = FALSE,
  column_order = 1:ncol(comut_mat_full[plot_genes, rownames(onco_meta_full)]),
  # pct_side = "right",
  column_title_rot = 90,
  border = TRUE
)

pdf(file = "oncoprint_full.pdf", width = 30, height = 13)
draw(ht_full)
dev.off()
```

