---
title: "EGFR Load Impact"
format: html
---

```{r setup}
library(tidyverse)
library(patchwork)
library(glue)
library(ComplexHeatmap)
library(here)

devtools::load_all()
```


```{r}
impact_data <- readRDS(here("results", "impact_data.rds"))
# metadata <- load_metadata()
metadata <- read.table(file = here("metadata", "metadata_update_dec_2024.txt"), header = T, sep = "\t", quote = "\"")
metadata$has_impact <- !is.na(metadata$impact)
markers <- load_markers()
```


Use the oncokb mutations to define secondary MAPK alterations across samples

```{r}
library(gt)
oncokb_samp <- impact_data$oncokb_dat$oncokb_clindat[, c("dmp_sample", "oncogenic_mutations", "resistance_mutations")] %>%
  left_join(metadata[, c("sample_id", "closest_impact_to_sample", "de_novo_mutation_egfr_specifics", "secondary_egfr_genomic", "secondary_other_mapk_genomic")], by = c("dmp_sample" = "closest_impact_to_sample"))
oncokb_samp$has_impact <- T

# The five without impact
other_samp <- metadata[is.na(metadata$closest_impact_to_sample), c("sample_id", "closest_impact_to_sample", "de_novo_mutation_egfr_specifics", "secondary_egfr_genomic", "secondary_other_mapk_genomic")]
other_samp$has_impact <- F

full_set <- plyr::rbind.fill(oncokb_samp, other_samp) %>%
  left_join(metadata[, c("sample_id", "time_point", "procedure", "impact", "osi_start")]) %>%
  mutate(
    impact_tp_diff = impact - procedure,
    impact_osi_diff = impact - osi_start
  ) %>%
  mutate(
    impact_before_procedure = impact_tp_diff < 0,
    impact_before_osi = impact_osi_diff < 0
  ) %>%
  dplyr::relocate(sample_id, dmp_sample, has_impact, time_point, impact, procedure) %>%
  mutate(
    time_point = fct_relevel(time_point, "Pre", "On", "PD"),
    sample_id = fct_relevel(sample_id, gtools::mixedsort(sample_id))
  ) %>%
  arrange(time_point, sample_id)

write.table(full_set, file = here("metadata", "supplemental_mapk_mutation_table_raw.txt"), col.names = T, quote = F, sep = "\t", row.names = F)


tab <- full_set %>%
  mutate(sample_id = factor(sample_id, levels = gtools::mixedsort(unique(metadata$sample_id)))) %>%
  select("Sample" = sample_id, time_point, "IMPACT ID" = closest_impact_to_sample, "Primary EGFR Mutation" = de_novo_mutation_egfr_specifics, "Secondary EGFR Mutation" = secondary_egfr_genomic, "Secondary MAPK Alteration" = secondary_other_mapk_genomic, oncogenic_mutations, resistance_mutations, impact, procedure, impact_before_procedure, impact_before_osi) %>%
  arrange(time_point, Sample) %>%
  group_by(time_point) %>%
  gt(rowname_col = "Sample") %>%
  tab_spanner(
    label = "EGFR Mutations",
    columns = c("Primary EGFR Mutation", "Secondary EGFR Mutation")
  ) %>%
  # cols_width(
  #   # This isn't quite working
  #   Sample ~ px(200),
  #   starts_with("IMPACT") ~ px(500)
  # ) %>%
  opt_stylize(style = 4) # %>%
# tab_style(
#   style = list(
#     cell_fill(color = "blue")
#   ),
#   locations = cells_body(
#     columns = "Histological Transformation",
#     rows = "Histological Transformation" == "sclc"
#   )
# )

# gtsave(tab, filename = here("plots", "summary_table.html"))
tab
```



Format a nice table with the relevant information per sample as a reference

```{r, fig.width=10, fig.height=8}
library(gt)
tab <- metadata %>%
  mutate(sample_id = factor(sample_id, levels = gtools::mixedsort(unique(metadata$sample_id)))) %>%
  select("Sample" = sample_id, time_point, "IMPACT ID" = closest_impact_to_sample, "Primary EGFR Mutation" = de_novo_mutation_egfr_specifics, "Secondary EGFR Mutation" = secondary_egfr_genomic, "Secondary MAPK Alteration" = secondary_other_mapk_genomic, ) %>%
  arrange(time_point, Sample) %>%
  group_by(time_point) %>%
  gt(rowname_col = "Sample") %>%
  tab_spanner(
    label = "EGFR Mutations",
    columns = c("Primary EGFR Mutation", "Secondary EGFR Mutation")
  ) %>%
  # cols_width(
  #   # This isn't quite working
  #   Sample ~ px(200),
  #   starts_with("IMPACT") ~ px(500)
  # ) %>%
  opt_stylize(style = 4) # %>%
# tab_style(
#   style = list(
#     cell_fill(color = "blue")
#   ),
#   locations = cells_body(
#     columns = "Histological Transformation",
#     rows = "Histological Transformation" == "sclc"
#   )
# )

# gtsave(tab, filename = here("plots", "summary_table.html"))
tab
```



```{r}
```



```{r}
onco_signals <- c("Cell Cycle", "HIPPO", "MYC", "NOTCH", "NRF2", "PI3K", "TGF-Beta", "RTK RAS", "TP53", "WNT")
res <- map(onco_signals, .f = function(x) {
  sr <- 1
  if (x == "Cell Cycle") {
    sr <- 3
  }
  df <- openxlsx::read.xlsx(xlsxFile = here("impact", "mmc3.xlsx"), sheet = x, check.names = F, colNames = T, rowNames = F, startRow = sr) %>%
    dplyr::mutate("Pathway" = x, .before = 1) %>%
    janitor::clean_names()
  return(df)
})

genelist <- map2(res, onco_signals, \(x, y) {
  data.frame(Pathway = y, Gene = x$gene)
}) %>%
  list_rbind()

resistance_genes <- data.frame("Pathway_resistance" = "Resistance Gene", Gene = markers$resistance_genes)
```


```{r}
impact_groupings <- read.table(file = here("impact", "impact_pathway_groupings.txt"), header = T, sep = "\t")

impact_2 <- read.table(file = here("impact", "impact_annos_sarcoma.txt"), header = T, sep = "\t")
```


```{r}
impact_groupings$Hugo_symbol %in% genelist$Gene
```



```{r}
table(unique(impact_data$onco_dat_full$Gene) %in% genelist$Gene)
```


```{r}
table(unique(impact_data$onco_dat_full$Gene) %in% impact_groupings$Hugo_symbol)
```


```{r}
table(unique(impact_data$onco_dat_full$Gene) %in% impact_2$Hugo.symbol)
```

```{r}
table(unique(impact_data$onco_dat_full$Gene) %in% resistance_genes$Gene)
```


Merge

```{r}
impact_list <- genelist %>%
  left_join(y = impact_groupings, by = c("Gene" = "Hugo_symbol"), suffix = c("_cell", "_list1")) %>%
  left_join(y = impact_2, by = c("Gene" = "Hugo.symbol")) %>%
  dplyr::rename("Pathway_sarcoma" = "Pathway") %>%
  dplyr::relocate("Gene") %>%
  mutate(
    in_egfr = Gene %in% impact_data$onco_dat_full$Gene,
    Pathway_cell = janitor::make_clean_names(Pathway_cell, allow_dupes = T),
    Pathway_list1 = janitor::make_clean_names(Pathway_list1, allow_dupes = T),
    Pathway_sarcoma = janitor::make_clean_names(Pathway_sarcoma, allow_dupes = T),
  ) %>%
  add_count(Gene)

impact_list$pathway <- impact_list$Pathway_sarcoma
impact_list$pathway[impact_list$pathway == "na" & impact_list$Pathway_cell != "na"] <- impact_list$Pathway_cell[impact_list$pathway == "na" & impact_list$Pathway_cell != "na"]

impact_list$pathway[impact_list$pathway == "na" & impact_list$Pathway_list1 != "na"] <- impact_list$Pathway_list1[impact_list$pathway == "na" & impact_list$Pathway_list1 != "na"]

# if n = 2 take pathway_cell or
impact_list$pathway[impact_list$n == 2] <- impact_list$Pathway_cell[impact_list$n == 2]

impact_df <- impact_list %>%
  distinct(Gene, pathway, in_egfr) %>%
  filter(!(Gene == "SCRIB" & pathway == "rtk_ras"))

rownames(impact_df) <- impact_df$Gene

impact_df$resistance_gene <- impact_df$Gene %in% resistance_genes$Gene

write.table(x = impact_df, file = here("impact", "impact_mapping.txt"), quote = F, sep = "\t", row.names = F, col.names = T)
```

# ONCOPRINT

Metadata

```{r}
cmap <- load_colors()

onco_meta <- impact_data$clin_dat_sample %>%
  inner_join(metadata, by = c("dmp_sample" = "closest_impact_to_sample")) %>%
  left_join(impact_data$facets_data)

rownames(onco_meta) <- onco_meta$sample_id

onco_meta$time_point <- factor(onco_meta$time_point, levels = c("Pre", "On", "Progression"))

onco_dat <- inner_join(impact_data$onco_dat_full, onco_meta)
```


```{r}
# filter(onco_dat, Gene %in% markers$resistance_genes) %>% view
```


Prepare for plotting

```{r, fig.width=11, fig.height=10}
# Add sensible patient IDs
# Add in patients that are missing (keep empty)
# Fix colors and rect shapes
# Sample and patient metadata
# Sample metadata

library(ComplexHeatmap)

comut_mat <- onco_dat %>%
  filter(Gene %in% impact_df$Gene) %>%
  filter(dmp_tumor %in% onco_meta$dmp_tumor) %>%
  pivot_wider(id_cols = sample_id, names_from = Gene, values_from = mutation_type_simple, values_fn = function(x) paste0(x, collapse = ";"), values_fill = "") %>%
  column_to_rownames("sample_id") %>%
  as.matrix() %>%
  t()

### Add the empty samples
samps_add <- setdiff(onco_meta$dmp_tumor, colnames(comut_mat))
samps_add_mat <- matrix(data = "", nrow = nrow(comut_mat), ncol = length(samps_add), dimnames = list(rownames(comut_mat), samps_add))
comut_mat_final <- cbind(comut_mat, samps_add_mat)
####

comut_mat_final <- comut_mat

# Reorder metadata
onco_meta <- onco_meta[colnames(comut_mat_final), ]

mut_pal <- c(
  "Missense Mutation" = "#1C91D4",
  "Nonsense Mutation" = "#D5C711",
  "Deep Deletion" = "blue",
  "Amplification" = "red",
  "Frameshift Indel" = "#D55E00",
  "Inframe Indel" = "#CC79A7",
  "RNA Fusion" = "#F0E442",
  "Structural Variant" = "#56B4E9",
  "Other" = "#007756"
)


gene_groups <- impact_df[rownames(comut_mat_final), "pathway", drop = F]

gene_groups$pathway <- fct_infreq(gene_groups$pathway)

gene_groups <- gene_groups %>%
  arrange(pathway)


comut_mat_final <- comut_mat_final[rownames(gene_groups), ]

# cli::cli_alert_danger("Missing {unique(onco_dat$mutation_type_simple)[!unique(onco_dat$mutation_type_simple) %in% names(mut_pal)]} mutations")
```


```{r, fig.width=11, fig.height=10}
alter_fun <- list(
  background = alter_graphic("rect", fill = "#CCCCCC"),
  "Amplification" = alter_graphic("rect", fill = mut_pal["Amplification"]),
  "Deep Deletion" = alter_graphic("rect", fill = mut_pal["Deep Deletion"]),
  "Missense Mutation" = alter_graphic("rect", height = 0.33, fill = mut_pal["Missense Mutation"]),
  "Nonsense Mutation" = alter_graphic("rect", height = 0.33, fill = mut_pal["Nonsense Mutation"]),
  "Frameshift Indel" = alter_graphic("rect", height = 0.33, fill = mut_pal["Frameshift Indel"]),
  "Inframe Indel" = alter_graphic("rect", height = 0.33, fill = mut_pal["Inframe Indel"]),
  "RNA Fusion" = alter_graphic("rect", height = 1, width = 0.33, fill = mut_pal["RNA Fusion"]),
  "Structural Variant" = alter_graphic("rect", height = 0.33, fill = mut_pal["Structural Variant"]),
  "Other" = alter_graphic("rect", height = 0.33, fill = mut_pal["Other"])
)

# tp_names <- levels(factor(onco_meta$timepoint))
# tp_cols <- set_names(circlize::colorRamp2(c(0, 5), c("lightblue", "darkblue"))(1:length(tp_names)), tp_names)
#
# pt_cols <- set_names(ggthemes::ptol_pal()(length(unique(onco_meta$patient))), sort(unique(onco_meta$patient)))


anno_cols <- list(
  "timepoint" = cmap$time_point,
  "progressed" = cmap$tf,
  "mapk_sustained_genome" = cmap$tf,
  "histological_transformation_clinical" = cmap$tf,
  # "WGD" = cmap$tf,
  "resistance" = cmap$mech_resistance # ,
  # "age" = circlize::colorRamp2(range(onco_meta$AGE_AT_SEQ_REPORTED_YEARS, na.rm = T), hcl_palette = "SunsetDark", reverse = TRUE)
)

# anno_labels <- c("Mutation Type", names(anno_cols))
# anno_labels[anno_labels == "Age"] <- "Age (months)"
# anno_labels[anno_labels == "WBC"] <- "WBC (10^9/l)"

tmb_breaks <- c(1, 10, 100)


top_anno <- HeatmapAnnotation(
  annotation_name_side = "left",
  annotation_label = c(
    "TMB",
    # "FGA",
    # "WGD",
    "Ever progressed?",
    # "Age",
    # "Panel type",
    "mapk_sustained_genome",
    "histological_transformation_clinical",
    "Time point"
  ),
  # cbar = anno_oncoprint_barplot(),
  TMB = anno_barplot(
    x = log10(onco_meta$CVR_TMB_SCORE + 1),
    axis_param = list(
      at = log10(tmb_breaks + 1),
      labels = tmb_breaks,
      side = "left"
    ),
    ylim = c(log10(1), log10(100)),
    border = FALSE,
    baseline = 0
  ),
  # FGA = onco_meta$facets_fga,
  # WGD = onco_meta$facets_wgd,
  # timepoint = onco_meta$time_point,
  progressed = onco_meta$progressed,
  # age = onco_meta$AGE_AT_SEQ_REPORTED_YEARS,
  # assay_type = onco_meta$GENE_PANEL,
  mapk_sustained_genome = onco_meta$mapk_sustained_genomic,
  histological_transformation_clinical = onco_meta$histological_transformation_clinical_simple,
  timepoint = anno_block(
    gp = gpar(fill = cmap$time_point),
    labels = c("Pre-treatment", "On\ntreatment", "Progression"),
    labels_gp = gpar(col = "white", fontsize = 11)
  ),

  # timepoint = onco_meta$timepoint,
  # response = onco_meta$response_category,
  # disease_status = onco_meta$disease_status,
  # chemo = onco_meta$chemo,
  # transplant = onco_meta$transplant,
  # t8_21 = onco_meta$t8_21,
  col = anno_cols
)


gene_anno <- HeatmapAnnotation(df = gene_groups, which = "row")
```


Plot the plot

```{r}
min_mut_freq <- 0.00

# Get mutation frequencies
mut_freq <- (apply(comut_mat_final, MARGIN = 1, function(x) sum(x != "")) / ncol(comut_mat_final)) %>%
  sort(decreasing = T)

plot_genes <- names(which(mut_freq >= min_mut_freq))
```

Plot and save

```{r, fig.width=15, fig.height=12}
ht <- oncoPrint(comut_mat_final,
  right_annotation = rowAnnotation(
    row_barplot = anno_oncoprint_barplot(show_fraction = F, border = F)
  ),
  top_annotation = top_anno,
  column_split = onco_meta$time_point,
  row_split = gene_groups$pathway,
  row_title_rot = 0,
  # row_names_side = "left",
  column_title = NULL,
  alter_fun = alter_fun,
  col = mut_pal,
  show_column_names = TRUE,
  remove_empty_columns = FALSE,
  # pct_side = "right",
  column_title_rot = 0,
  border = TRUE
)

# pdf(file = "oncoprint.pdf", width = 15, height = 13)
draw(ht)
# dev.off()

# suppressWarnings(ht)
```

Binarize each sample has having an alteration or not. Plus secondary EGFR or not

```{r}
egfr_hits <- onco_dat %>%
  filter(Gene == "EGFR") %>%
  left_join(impact_df) %>%
  add_count(sample_id, Gene) %>%
  distinct(sample_id, Gene, n, mutation_type_simple, HGVSp_Short, mech_resistance, secondary_egfr_genomic, de_novo_mutation_egfr_specifics, time_point, pathway, mech_of_resistance_to_osimertinib, mech_of_resistance_classification, notes) %>%
  arrange(time_point, sample_id, n)

egfr_hits$mut_type <- egfr_hits$HGVSp_Short

egfr_hits$mut_type[is.na(egfr_hits$mut_type)] <- egfr_hits$mutation_type_simple[is.na(egfr_hits$mut_type)]


p1 <- egfr_hits %>%
  distinct(sample_id, time_point, n) %>%
  ggplot(aes(x = fct_reorder(sample_id, -n, mean), y = n)) +
  geom_col() +
  facet_grid(. ~ time_point, scales = "free", space = "free") +
  guides(x = guide_axis(angle = 90)) +
  labs(x = "Sample", y = "")



p2 <- ggplot(egfr_hits, aes(x = fct_reorder(sample_id, -n, mean), y = fct_rev(fct_infreq(mut_type)))) +
  geom_tile() +
  facet_grid(. ~ time_point, scales = "free", space = "free") +
  guides(x = guide_axis(angle = 90)) +
  labs(x = "Sample", y = "Mutation")

p1 + p2 + plot_layout(axes = "collect", ncol = 1, heights = c(1, 5))
```

Reannotate samples using the following

```{r}
egfr_hits %>%
  filter(n > 1 | mech_of_resistance_classification == "On Target") %>%
  pivot_wider(id_cols = c(sample_id, time_point, mech_resistance, secondary_egfr_genomic, de_novo_mutation_egfr_specifics, mech_of_resistance_classification, mech_of_resistance_to_osimertinib), names_from = Gene, values_from = mut_type, values_fn = function(x) paste0(x, collapse = "; "), values_fill = "") %>%
  view()
```


Then check for other secondary hits that can be labelled according to known resistance mechanisms

```{r}
all_resist <- onco_dat %>%
  # filter(Gene =="EGFR") %>%
  left_join(impact_df) %>%
  add_count(sample_id, Gene) %>%
  distinct(sample_id, Gene, n, mutation_type, mutation_type_simple, HGVSp_Short, secondary_egfr_genomic, secondary_other_mapk_genomic, de_novo_mutation_egfr_specifics, time_point, mech_resistance, pathway, mech_of_resistance_to_osimertinib, mech_of_resistance_classification, notes) %>%
  arrange(time_point, sample_id, n)

all_resist$mut_type <- all_resist$HGVSp_Short

all_resist$mut_type[is.na(all_resist$mut_type)] <- all_resist$mutation_type_simple[is.na(all_resist$mut_type)]
all_resist$mut_type[is.na(all_resist$mut_type) | all_resist$mut_type == ""] <- all_resist$mutation_type[is.na(all_resist$mut_type) | all_resist$mut_type == ""]


other_muts <- all_resist %>%
  filter(!is.na(pathway), Gene != "EGFR") %>%
  mutate("gene_mut" = paste(Gene, mut_type, sep = ": ")) %>%
  # filter(n > 1 | mech_of_resistance_classification == "On Target") %>%
  pivot_wider(id_cols = c(sample_id, time_point, mech_resistance, secondary_egfr_genomic, de_novo_mutation_egfr_specifics, secondary_other_mapk_genomic, mech_of_resistance_classification, mech_of_resistance_to_osimertinib), names_from = pathway, values_from = gene_mut, values_fn = function(x) paste0(x, collapse = "; "), values_fill = "")

openxlsx::write.xlsx(other_muts, file = here("impact", "other_muts.xlsx"))
```


```{r}
other_muts_binary <- other_muts %>%
  select(any_of(unique(impact_df$pathway))) %>%
  as.matrix()

other_muts_binary[other_muts_binary != ""] <- 1
other_muts_binary[other_muts_binary == ""] <- 0
other_muts_binary <- apply(other_muts_binary, MARGIN = 2, as.numeric)

rownames(other_muts_binary) <- other_muts$sample_id

Heatmap(t(other_muts_binary), cluster_rows = F, cluster_columns = F)
```



Format a nice table with the relevant information per sample as a reference

```{r, fig.width=10, fig.height=8}
library(gt)
tab <- metadata %>%
  mutate(sample_id = factor(sample_id, levels = gtools::mixedsort(unique(metadata$sample_id)))) %>%
  select("Sample" = sample_id, time_point, "IMPACT ID" = closest_impact_to_sample, "Primary EGFR Mutation" = de_novo_mutation_egfr_specifics, "Secondary EGFR Mutation" = secondary_egfr_genomic, "Secondary MAPK Alteration" = secondary_other_mapk_genomic) %>%
  arrange(time_point, Sample) %>%
  group_by(time_point) %>%
  gt(rowname_col = "Sample") %>%
  tab_spanner(
    label = "EGFR Mutations",
    columns = c("Primary EGFR Mutation", "Secondary EGFR Mutation")
  ) %>%
  # cols_width(
  #   # This isn't quite working
  #   Sample ~ px(200),
  #   starts_with("IMPACT") ~ px(500)
  # ) %>%
  opt_stylize(style = 4) # %>%
# tab_style(
#   style = list(
#     cell_fill(color = "blue")
#   ),
#   locations = cells_body(
#     columns = "Histological Transformation",
#     rows = "Histological Transformation" == "sclc"
#   )
# )

# gtsave(tab, filename = here("plots", "summary_table.html"))
tab
```



