---
title: "TCGA MAPK analysis"
author: "Matthew Zatzman"
---

# Introduction

```{r setup}
#| output: false
library(tidyverse)
library(ggthemes)
library(ggsci)
library(ggrepel)
library(patchwork)
library(glue)
library(ggh4x)
library(patchwork)
library(Seurat)
library(ggbeeswarm)
library(tidyverse)
library(readxl)
library(easystats)
library(here)
theme_set(theme_classic())
devtools::load_all()
```


```{r}
cmap <- load_colors()
markers <- load_markers()
```


# TCGA

## Load data

```{r}
tcga_alts <- read_excel(here("data", "pancan_alteration_matrix.xlsx"), sheet = 2)

tcga_subtypes <- read_excel(here("data", "tcga_subtypes.xlsx"), skip = 2)
# For clinical outcome endpoints, we recommend the use of PFI for progression-free interval, and OS for overall survival. Both endpoints are relatively accurate. Given the relatively short follow-up time, PFI is preferred over OS. Detailed recommendations please refer to Table 3 in the accompanying paper.
tcga_clin <- read_excel(here("data/TCGA-CDR-SupplementalTableS1.xlsx"),
  sheet = "TCGA-CDR",
)


tcga_purplo <- read.table(here("data", "TCGA_mastercalls.abs_tables_JSedit.fixed.txt"), header = T, sep = "\t")

oncopaths <- c("Cell Cycle", "HIPPO", "MYC", "NOTCH", "NRF2", "PI3K", "TGF-Beta", "RTK RAS", "TP53", "WNT")

tcga_oncopaths_1 <- read_excel(here("data", "tcga_oncopaths.xlsx"), sheet = 1, skip = 2) %>%
  janitor::clean_names() %>%
  select(gene, og_tsg) %>%
  mutate(pathway = oncopaths[1])

tcga_oncopaths <- map(2:10, .f = function(i) {
  x <- read_excel(here("data", "tcga_oncopaths.xlsx"), sheet = oncopaths[[i]]) %>%
    janitor::clean_names() %>%
    select(gene, og_tsg) %>%
    mutate(pathway = oncopaths[[i]])
}) %>%
  list_rbind()

tcga_oncopaths <- rbind(tcga_oncopaths, tcga_oncopaths_1)

impact_oncopaths <- read.table(file = here("data", "impact_pathway_groups.txt"), header = T, sep = "\t")

tcga_clin <- tcga_clin %>%
  mutate(tumor_stage_simple = case_when(
    grepl("Stage IV", ajcc_pathologic_tumor_stage) ~ "Stage IV",
    grepl("Stage III", ajcc_pathologic_tumor_stage) ~ "Stage III",
    grepl("Stage II", ajcc_pathologic_tumor_stage) ~ "Stage II",
    grepl("Stage I", ajcc_pathologic_tumor_stage) ~ "Stage I",
    .default = "Unknown"
  )) %>%
  mutate(tumor_stage_simple = fct_relevel(tumor_stage_simple, "Stage I"))

gexp_data <- data.table::fread(file = here("data", "EBPlusPlusAdjustPANCAN_IlluminaHiSeq_RNASeqV2.geneExp.tsv"))
gexp_data$gene <- str_split(gexp_data$gene_id, pattern = "\\|", simplify = T)[, 1]
gexp_data$gene_id <- NULL
gexp_data <- gexp_data[gexp_data$gene != "?" & gexp_data$gene != "SLC35E2", ]
genes <- gexp_data$gene
gexp_data$gene <- NULL
gexp_data <- as.matrix(gexp_data)
rownames(gexp_data) <- genes
```



```{r}
pathways <- read.table(here("metadata", "pathways", "egfr_pathways.txt"), header = T, sep = "\t")

pathways <- split(pathways$Gene, pathways$pathway)
pathway_names <- gsub("_GIRARD|_CCLE", "", names(pathways))
```


## Score Pathways

```{r}
library(UCell)

# x <- apply(gexp_data, MARGIN = 2, FUN = function(x) {sum(x>100)})
# ranks_tcga <- UCell::StoreRankings_UCell(matrix = gexp_data, maxRank = 10000, BPPARAM = BiocParallel::MulticoreParam(workers=16, progressbar = T), chunk.size = 100)

# scores_tcga <- UCell::ScoreSignatures_UCell(precalc.ranks = ranks_tcga, BPPARAM = BiocParallel::MulticoreParam(workers=16, progressbar = T), features = pathways, name = "")

# write.table(scores_tcga %>% as.data.frame() %>% rownames_to_column("sample_id"), file = gzfile(here("results", "tcga_pathway_scores.txt.gz")), quote = F, sep = "\t", row.names = F, col.names = T)
scores_tcga <- read.table(here("results", "tcga_pathway_scores.txt.gz"), header = T, sep = "\t")
```


```{r}
# scores_tcga <- scores_tcga %>% as.data.frame() %>% rownames_to_column("sample_id")

scores_tcga$SAMPLE_BARCODE <- substr(scores_tcga$sample_id, 1, 15)
```


```{r}
tcga_df <- inner_join(scores_tcga, tcga_subtypes) %>%
  dplyr::rename(
    "LUAD" = "LUAD_GIRARD",
    "LUSC" = "LUSC_GIRARD",
    "SCLC" = "SCLC_CCLE"
  ) %>%
  inner_join(tcga_alts) %>%
  left_join(tcga_clin[, -1], by = c("PATIENT_BARCODE" = "bcr_patient_barcode"))

tcga_df$CDKN2A <- as.numeric(tcga_df$CDKN2A)
mut_genes_tcga <- tcga_df %>%
  select(CCND1:WWTR1) %>%
  colnames()

# Add binary pathway mutation indicator
tcga_pathway_df <- tcga_df %>%
  select(sample_id, mut_genes_tcga) %>%
  pivot_longer(cols = mut_genes_tcga, names_to = "gene", values_to = "mut") %>%
  left_join(tcga_oncopaths) %>%
  group_by(sample_id, pathway) %>%
  summarize(mut = sum(mut) > 0) %>%
  pivot_wider(id_cols = "sample_id", names_from = "pathway", values_from = mut, values_fn = function(x) sum(x, na.rm = T), values_fill = 0, names_prefix = "path_") %>%
  janitor::clean_names()


impact_pathway_df <- tcga_df %>%
  select(sample_id, mut_genes_tcga) %>%
  pivot_longer(cols = mut_genes_tcga, names_to = "gene", values_to = "mut") %>%
  filter(gene %in% impact_oncopaths$Hugo_symbol) %>%
  left_join(impact_oncopaths, by = c("gene" = "Hugo_symbol")) %>%
  group_by(sample_id, Pathway) %>%
  summarize(mut = sum(mut) > 0) %>%
  pivot_wider(id_cols = "sample_id", names_from = "Pathway", values_from = mut, values_fn = function(x) sum(x, na.rm = T), values_fill = 0, names_prefix = "pathimpact_") %>%
  janitor::clean_names()



tcga_df <- left_join(tcga_df, tcga_pathway_df) %>%
  left_join(impact_pathway_df) %>%
  mutate(array = substr(sample_id, 1, 15)) %>%
  left_join(tcga_purplo)

resist_muts <- markers$resistance_genes[markers$resistance_genes %in% colnames(tcga_df)]
```


## Plotting


```{r}
p1 <- tcga_df %>%
  ggplot(aes(x = fct_reorder(DISEASE, LUAD, median), y = LUAD)) +
  geom_boxplot() +
  guides(x = guide_axis(angle = 90))

p2 <- tcga_df %>%
  filter(SUBTYPE != "Not_Applicable") %>%
  ggplot(aes(x = fct_reorder(SUBTYPE, LUAD, median), y = LUAD)) +
  geom_boxplot() +
  facet_grid(. ~ DISEASE, scales = "free", space = "free") +
  guides(x = guide_axis(angle = 90))

p1 / p2
```

```{r}
p1 <- tcga_df %>%
  ggplot(aes(x = fct_reorder(DISEASE, LUSC, median), y = LUSC)) +
  geom_boxplot() +
  guides(x = guide_axis(angle = 90))

p2 <- tcga_df %>%
  filter(SUBTYPE != "Not_Applicable") %>%
  ggplot(aes(x = fct_reorder(SUBTYPE, LUSC, median), y = LUSC)) +
  geom_boxplot() +
  facet_grid(. ~ DISEASE, scales = "free", space = "free") +
  guides(x = guide_axis(angle = 90))
p1 / p2
```


```{r}
p1 <- tcga_df %>%
  ggplot(aes(x = fct_reorder(DISEASE, SCLC, median), y = SCLC)) +
  geom_boxplot() +
  guides(x = guide_axis(angle = 90))

p2 <- tcga_df %>%
  filter(SUBTYPE != "Not_Applicable") %>%
  ggplot(aes(x = fct_reorder(SUBTYPE, SCLC, median), y = SCLC)) +
  geom_boxplot() +
  facet_grid(. ~ DISEASE, scales = "free", space = "free") +
  guides(x = guide_axis(angle = 90))
p1 / p2
```


```{r, fig.width = 5, fig.height = 5}
p1 <- tcga_df %>%
  filter(DISEASE %in% c("LUAD")) %>%
  ggplot(aes(x = REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES, y = LUAD)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggpubr::stat_cor() +
  theme(aspect.ratio = 1) +
  labs(x = "MAPK")
p1
```



```{r, fig.width = 8, fig.height = 12}
p1 <- tcga_df %>%
  filter(DISEASE %in% c("LUAD", "LUSC")) %>%
  ggplot(aes(x = REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES, y = LUAD)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggpubr::stat_cor() +
  facet_grid(. ~ DISEASE)

p2 <- tcga_df %>%
  filter(DISEASE %in% c("LUAD", "LUSC")) %>%
  ggplot(aes(x = REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES, y = LUSC)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggpubr::stat_cor() +
  facet_grid(. ~ DISEASE)


p3 <- tcga_df %>%
  filter(DISEASE %in% c("LUAD", "LUSC")) %>%
  ggplot(aes(x = REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES, y = SCLC)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggpubr::stat_cor() +
  facet_grid(. ~ DISEASE)


p1 + p2 + p3 + plot_layout(ncol = 1) & theme(aspect.ratio = 1) & labs(x = "MAPK Pathway")
```


```{r}
tcga_cor_df <- tcga_df %>%
  group_by(DISEASE) %>%
  correlation(select = c("LUAD", "LUSC", "SCLC"), select2 = grep("MAPK", names(pathways), value = T), method = "spearman", p_adjust = "none")
tcga_cor_df$sig <- ifelse(tcga_cor_df$p < 0.05, "*", "")
```


```{r, fig.width = 10, fig.height = 10}
p <- tcga_cor_df %>%
  # filter(Group %in% c("LUAD")) %>%
  ggplot(aes(x = Parameter1, y = Parameter2)) +
  geom_point(aes(fill = rho, size = -log10(p)), pch = 21) +
  facet_wrap(~Group) +
  geom_text(aes(label = sig), vjust = 0.8) +
  scale_fill_gradient2(low = scales::muted("blue"), high = scales::muted("red")) +
  # scale_fill_gradientn(colors = palette_bluebrown(palette = "contrast")(5)[c(2,3,4)],
  #                      rescaler = ~ scales::rescale_mid(.x, mid = 0)) +
  guides(x = guide_axis(angle = 90))
p
```


```{r}
p3 <- tcga_df %>%
  # filter(DISEASE %in% c("LUAD", "LUSC")) %>%
  ggplot(aes(x = REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES, y = LUAD)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggpubr::stat_cor() +
  facet_wrap(~DISEASE)
```


```{r}
# pl_dat <- x
p <- tcga_cor_df %>%
  arrange(desc(p)) %>%
  ggplot(aes(x = Parameter1, y = rho)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  gghalves::geom_half_boxplot(outliers = F, aes(fill = Parameter1), errorbar.draw = T) +
  gghalves::geom_half_point_panel(aes(size = -log10(p), fill = Parameter1), pch = 21) +
  geom_text_repel(data = tcga_cor_df[tcga_cor_df$Group %in% c("LUAD", "LUSC"), ], aes(label = Group)) +
  # gghalves::geom_half_point_panel(data = pl_dat[pl_dat$p < 0.05,], aes(size = -log10(p), fill = Parameter1), pch = 21) +
  # gghalves::geom_half_point_panel(data = pl_dat[pl_dat$p >= 0.05,], aes(size = -log10(p)), fill = "grey", pch = 21) +
  # geom_jitter(aes(size = -log10(p), fill = Parameter1), pch = 21) +
  facet_grid(. ~ Parameter2) +
  guides(fill = "none", size = guide_legend(override.aes = list(pch = 19))) +
  scale_fill_manual(values = cmap$histology_predominant_short) +
  ggpubr::geom_pwc(tip.length = 0, label = "p.adj.signif", step.increase = 0.05, vjust = 0.5) +
  theme_few() +
  theme(
    strip.clip = "off",
    plot.title = element_text(hjust = 0.5, size = 12), plot.title.position = "panel",
    axis.text = element_text(color = "black")
  ) +
  labs(title = "Per sample MAPK correlations", x = "Lineage Signature", y = bquote("Spearman's" ~ rho), size = bquote("-log10"[Padj]))
# ggsave(p, file = here("plots", "per_sample_mapk_cor.pdf"), width = 3.5, height = 3.5)
p
```


```{r, fig.width = 15, fig.height = 10}
tcga_cor_df %>%
  mutate(dir = sign(rho)) %>%
  ggplot(aes(y = -log10(p), x = rho)) +
  geom_vline(xintercept = 0) +
  geom_point() +
  geom_text_repel(aes(label = Group), max.overlaps = Inf, min.segment.length = 0) +
  facet_grid(Parameter1 ~ Parameter2) +
  theme_few()
```


What is specifically anticorrelate with either MAPK or LUAD loass
```{r}
tcga_cor_df <- tcga_df %>%
  filter(DISEASE == "LUAD") %>%
  # group_by(DISEASE) %>%
  correlation(select = c("LUAD"), select2 = pathway_names[pathway_names != "LUAD"], method = "spearman", p_adjust = "none")
tcga_cor_df$sig <- ifelse(tcga_cor_df$p < 0.05, "*", "")
```



Look at the mutation frequencies in LUAD and LUSC as a function of their SCLC, LUSC or LUAD signature


```{r}
p1 <- tcga_df %>%
  filter(DISEASE %in% c("LUAD"), EGFR == 1) %>%
  ggplot(aes(x = REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES, y = LUAD)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggpubr::stat_cor() +
  theme(aspect.ratio = 1) +
  labs(x = "MAPK")
p1
```


```{r}
## OLD VERSION
# tcga_df$DISEASE <- factor(tcga_df$DISEASE)
#
# tests <- expand.grid(c("LUAD", "LUSC"), c("LUAD", "SCLC", "LUSC"))
#
# res <- map(1:nrow(tests), .f = function(i) {
#   disease = as.character(tests[i,1])
#   path = as.character(tests[i,2])
#
#
#   dat <- tcga_df[tcga_df$DISEASE == disease,]
#
#   p <- (colSums(dat[,mut_genes_tcga], na.rm = T) / nrow(dat)) %>%
#     enframe() %>%
#     ggplot(aes(x = fct_reorder(name, value, median), y = value)) +
#     geom_col() +
#     guides(x = guide_axis(angle = 90))
#
#   test_genes <- p$data$name[p$data$value >0.05]
#
#   mods <- map(test_genes, .f = function(g) {
#     mod <- glm(formula = as.formula(glue("{g} ~ {path}")), family = binomial(), data = dat)
#     return(mod)
#   })
#   names(mods) <- test_genes
#
#   param_df <- map(names(mods), .f = function(g) {
#     parameters(mods[[g]], exponentiate = T) %>%
#       mutate(Gene = g)
#   }) %>%
#     list_rbind() %>%
#     mutate(DISEASE = disease,
#            pathway = path)
#   return(param_df)
# }) %>%
#   list_rbind()
```

## Model

Go other way? Explain the pathway expression using the mutations...

```{r}
tcga_df$DISEASE <- factor(tcga_df$DISEASE)

tests <- expand.grid(c("LUAD", "LUSC"), c("LUAD", "SCLC", "LUSC"))

res_tcga <- map(1:nrow(tests), .f = function(i) {
  disease <- as.character(tests[i, 1])
  path <- as.character(tests[i, 2])


  dat <- tcga_df[tcga_df$DISEASE == disease, ]

  p <- (colSums(dat[, mut_genes_tcga], na.rm = T) / nrow(dat)) %>%
    enframe() %>%
    ggplot(aes(x = fct_reorder(name, value, median), y = value)) +
    geom_col() +
    guides(x = guide_axis(angle = 90))

  test_genes <- p$data$name[p$data$value > 0.05]

  mod <- glm(formula = as.formula(glue("{path} ~ {paste(test_genes, collapse = '+')} + tumor_stage_simple + purity + ploidy + Subclonal.genome.fraction + REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES")), family = gaussian(), data = dat)


  param_df <- parameters(mod)
  param_df$pretty_names <- attributes(param_df)$pretty_names
  param_df <- param_df %>%
    mutate(
      DISEASE = disease,
      pathway = path
    )

  return(param_df)
}) %>%
  list_rbind()

res_tcga$label <- gsub("tumor_stage_simple", "", res_tcga$Parameter)
res_tcga$param_type <- res_tcga %>%
  mutate(param_type = case_when(
    Parameter == "(Intercept)" ~ "Intercept",
    Parameter %in% mut_genes_tcga ~ "Gene",
    grepl("stage", Parameter) ~ "Tumor Stage",
    Parameter %in% c("Subclonal.genome.fraction", "ploidy") ~ "CIN",
    .default = "Other"
  )) %>%
  pull(param_type) %>%
  factor()
```



```{r, fig.width = 9, fig.height = 12}
pl_dat <- res_tcga %>%
  filter(Parameter != "(Intercept)") %>%
  filter(DISEASE == "LUAD") %>%
  mutate(Gene = Parameter)

pl_dat$padj <- p.adjust(pl_dat$p, method = "holm")

gene_ord <- pl_dat[pl_dat$pathway == "LUAD", ] %>%
  mutate(label = fct_reorder(factor(label), Coefficient, median)) %>%
  pull(label) %>%
  levels()


p <- pl_dat %>%
  ggplot(aes(x = Coefficient, y = fct_relevel(label, gene_ord), color = pathway)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = seq(0.5, n_distinct(pl_dat$Gene) - 0.5, by = 1), linetype = "dashed", alpha = 0.5) +
  geom_pointrange(aes(xmin = CI_low, xmax = CI_high), position = position_dodge(width = 0.5)) +
  # scale_x_continuous(trans = "log2", breaks = scales::breaks_log(base = 2), labels = scales::label_log(base = 2)) +
  facet_grid(param_type ~ ., scales = "free_y", space = "free_y") +
  scale_color_manual(values = cmap$histology_predominant_short) +
  # scale_y_discrete(expand = c(0,0)) +
  theme_few() +
  theme(
    panel.border = element_part_rect(side = "tblr", fill = NA),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  ) +
  labs(y = NULL, x = "Coefficient\n(Lineage Association)", color = "Lineage Signature", title = "TCGA LUAD")
ggsave(p, file = here("plots", "tcga_pergene_lineage_assoc.pdf"), width = 5, height = 7)
p
```


As dotplot

```{r, fig.width=4, fig.height=5}
mat <- pl_dat %>%
  pivot_wider(id_cols = "pathway", names_from = "label", values_from = "Coefficient") %>%
  column_to_rownames("pathway") %>%
  as.matrix() %>%
  t()
cls_coefs <- hclust(dist(mat))

p <- pl_dat %>%
  ggplot(aes(x = pathway, y = fct_relevel(label, gene_ord))) +
  # ggplot(aes(x = pathway, y = fct_relevel(label, cls_coefs$label[cls_coefs$order]))) +
  geom_point(pch = 21, aes(fill = Coefficient, size = -log10(p))) +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"), limits = c(-0.05, 0.05),
    oob = scales::squish
  ) +
  geom_text() +
  # scale_x_continuous(trans = "log2", breaks = scales::breaks_log(base = 2), labels = scales::label_log(base = 2)) +
  facet_grid(fct_relevel(param_type, "Gene", "Tumor Stage", "CIN") ~ ., scales = "free_y", space = "free_y") +
  theme_few() +
  theme(
    panel.border = element_part_rect(side = "tblr", fill = NA),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    strip.clip = "off"
  ) +
  guides(x = guide_axis(angle = 45)) +
  labs(y = NULL, x = "Coefficient\n(Lineage Association)", color = "Lineage Signature", title = "TCGA LUAD")
ggsave(p, file = here("plots", "tcga_pergene_lineage_assoc_dotplot.pdf"), width = 4, height = 5)
p
```




## Model with Pathways

```{r}
tests <- expand.grid(c("LUAD", "LUSC"), c("LUAD", "SCLC", "LUSC"), c("path_", "pathimpact_"))

res_tcga_path <- map(1:nrow(tests), .f = function(i) {
  disease <- as.character(tests[i, 1])
  path <- as.character(tests[i, 2])
  prefix <- as.character(tests[i, 3])


  dat <- tcga_df[tcga_df$DISEASE == disease, ]

  colns <- grep(glue("^{prefix}"), colnames(dat), value = T)

  p <- (colSums(dat[, colns], na.rm = T) / nrow(dat)) %>%
    enframe() %>%
    ggplot(aes(x = fct_reorder(name, value, median), y = value)) +
    geom_col() +
    guides(x = guide_axis(angle = 90))

  test_genes <- p$data$name[p$data$value > 0]

  mod <- glm(formula = as.formula(glue("{path} ~ {paste(test_genes, collapse = '+')} + tumor_stage_simple + purity + ploidy + Subclonal.genome.fraction")), family = gaussian(), data = dat)


  param_df <- parameters(mod)
  param_df$pretty_names <- attributes(param_df)$pretty_names
  param_df <- param_df %>%
    mutate(
      DISEASE = disease,
      pathway = path
    )

  return(param_df)
}) %>%
  list_rbind()

res_tcga_path$label <- gsub("tumor_stage_simple", "", res_tcga_path$Parameter)
res_tcga_path$param_type <- res_tcga_path %>%
  mutate(param_type = case_when(
    Parameter == "(Intercept)" ~ "Intercept",
    grepl("path_", Parameter) ~ "Pathway (TCGA)",
    grepl("pathimpact_", Parameter) ~ "Pathway (IMPACT)",
    grepl("stage", Parameter) ~ "Tumor Stage",
    .default = "Other"
  )) %>%
  pull(param_type) %>%
  factor()
```


```{r, fig.width = 8, fig.height = 10}
pl_dat <- res_tcga_path %>%
  filter(Parameter != "(Intercept)") %>%
  filter(DISEASE == "LUAD") %>%
  mutate(Gene = Parameter)

pl_dat$padj <- p.adjust(pl_dat$p, method = "holm")

gene_ord <- pl_dat[pl_dat$pathway == "LUAD", ] %>%
  mutate(label = fct_reorder(factor(label), Coefficient, median)) %>%
  pull(label) %>%
  levels()


p <- pl_dat %>%
  ggplot(aes(x = Coefficient, y = fct_relevel(label, gene_ord), color = pathway)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = seq(0.5, n_distinct(pl_dat$Gene) - 0.5, by = 1), linetype = "dashed", alpha = 0.5) +
  geom_pointrange(aes(xmin = CI_low, xmax = CI_high), position = position_dodge(width = 0.5)) +
  # scale_x_continuous(trans = "log2", breaks = scales::breaks_log(base = 2), labels = scales::label_log(base = 2)) +
  facet_grid(param_type ~ ., scales = "free_y", space = "free_y") +
  scale_color_manual(values = cmap$histology_predominant_short) +
  # scale_y_discrete(expand = c(0,0)) +
  theme_few() +
  theme(panel.border = element_part_rect(side = "tblr", fill = NA)) +
  labs(y = NULL, x = "Coefficient\n(Lineage Association)", color = "Lineage Signature")
p
```


### Subet for EGFR mut and repeat

Should also subset to TP53 to show BRAF, KRAS and EGFR effects within TP53mut (which are all positive)

```{r}
tcga_df$DISEASE <- factor(tcga_df$DISEASE)

tests <- expand.grid(c("LUAD"), c("LUAD", "SCLC", "LUSC"))

res_tcga_egfr <- map(1:nrow(tests), .f = function(i) {
  disease <- as.character(tests[i, 1])
  path <- as.character(tests[i, 2])

  dat <- tcga_df[tcga_df$DISEASE == disease & tcga_df$EGFR == 1, ]
  # dat <- tcga_df[tcga_df$DISEASE == disease & tcga_df$TP53 == 1,]

  p <- (colSums(dat[, mut_genes_tcga], na.rm = T) / nrow(dat)) %>%
    enframe() %>%
    ggplot(aes(x = fct_reorder(name, value, median), y = value)) +
    geom_col() +
    guides(x = guide_axis(angle = 90))

  test_genes <- p$data$name[p$data$value > 0.05]
  test_genes <- test_genes[test_genes != "EGFR"]

  # mod <- glm(formula = as.formula(glue("{path} ~ {paste(test_genes, collapse = '+')} + tumor_stage_simple + purity")), family = gaussian(), data = dat)

  mod <- glm(formula = as.formula(glue("{path} ~ TP53 + RB1 + tumor_stage_simple + purity + REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES")), family = gaussian(), data = dat)
  # mod <- glm(formula = as.formula(glue("{path} ~ KRAS + EGFR + tumor_stage_simple + purity")), family = gaussian(), data = dat)



  param_df <- parameters(mod)
  param_df$pretty_names <- attributes(param_df)$pretty_names
  param_df <- param_df %>%
    mutate(
      DISEASE = disease,
      pathway = path
    )

  return(param_df)
}) %>%
  list_rbind()

res_tcga_egfr$label <- gsub("tumor_stage_simple", "", res_tcga_egfr$Parameter)
res_tcga_egfr$param_type <- res_tcga_egfr %>%
  mutate(param_type = case_when(
    Parameter == "(Intercept)" ~ "Intercept",
    grepl("path_", Parameter) ~ "Pathway (TCGA)",
    grepl("pathimpact_", Parameter) ~ "Pathway (IMPACT)",
    grepl("stage", Parameter) ~ "Tumor Stage",
    .default = "Other"
  )) %>%
  pull(param_type) %>%
  factor()
```



```{r, fig.width = 5, fig.height = 4}
pl_dat <- res_tcga_egfr %>%
  filter(Parameter != "(Intercept)") %>%
  filter(DISEASE == "LUAD") %>%
  mutate(Gene = Parameter)

pl_dat$padj <- p.adjust(pl_dat$p, method = "holm")

gene_ord <- pl_dat[pl_dat$pathway == "LUAD", ] %>%
  mutate(label = fct_reorder(factor(label), Coefficient, median)) %>%
  pull(label) %>%
  levels()


p <- pl_dat %>%
  ggplot(aes(x = Coefficient, y = fct_relevel(label, gene_ord), color = pathway)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = seq(0.5, n_distinct(pl_dat$Gene) - 0.5, by = 1), linetype = "dashed", alpha = 0.5) +
  geom_pointrange(aes(xmin = CI_low, xmax = CI_high), position = position_dodge(width = 0.5)) +
  # scale_x_continuous(trans = "log2", breaks = scales::breaks_log(base = 2), labels = scales::label_log(base = 2)) +
  facet_grid(param_type ~ ., scales = "free_y", space = "free_y") +
  scale_color_manual(values = cmap$histology_predominant_short) +
  # scale_y_discrete(expand = c(0,0)) +
  theme_few() +
  theme(
    panel.border = element_part_rect(side = "tblr", fill = NA),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
  ) +
  labs(y = NULL, x = "Coefficient\n(Lineage Association)", color = "Lineage Signature", title = "TCGA EGFR Mutant LUAD")
ggsave(p, file = here("plots", "tcga_pergene_lineage_assoc_egfr.pdf"), width = 5, height = 4)
p
```




Nested comparisons

```{r}
luad_dat <- tcga_df %>%
  filter(DISEASE == "LUAD")
mod <- glm(formula = as.formula(glue("LUAD ~ TP53 * EGFR + TP53 * BRAF  + tumor_stage_simple")), family = gaussian(), data = dat)
```




```{r, fig.width=5, fig.height = 8}
mut <- "RB1"
df_alts %>%
  filter(DISEASE %in% c("LUAD", "LUSC")) %>%
  filter(!is.na(.data[[mut]])) %>%
  pivot_longer(cols = c("LUAD", "LUSC", "SCLC"), names_to = "Signature", values_to = "Score") %>%
  ggplot(aes(x = as.factor(.data[[mut]]), y = Score)) +
  gghalves::geom_half_boxplot(outliers = F) +
  gghalves::geom_half_point_panel() +
  labs(x = NULL, title = mut) +
  # geom_violin(scale = "width") +
  # geom_boxplot(width = 0.2, outliers = F) +
  facet_grid(Signature ~ DISEASE) +
  ggpubr::geom_pwc(method = "t_test")
```


```{r, fig.width=15, fig.height = 9}
p <- df_alts %>%
  filter(DISEASE %in% c("LUAD")) %>%
  pivot_longer(cols = c("LUAD", "LUSC", "SCLC"), names_to = "Signature", values_to = "Score") %>%
  pivot_longer(cols = gene_ord, names_to = "Gene", values_to = "Mutated") %>%
  mutate(Mutated = factor(ifelse(Mutated == 0, "wt", "mut"), levels = c("wt", "mut"))) %>%
  filter(Gene %in% gene_ord, !is.na(Mutated)) %>%
  ggplot(aes(x = Mutated, y = Score)) +
  gghalves::geom_half_boxplot(outliers = F) +
  gghalves::geom_half_point_panel() +
  labs(x = NULL) +
  theme_few() +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  # geom_violin(scale = "width") +
  # geom_boxplot(width = 0.2, outliers = F) +
  facet_grid(Signature ~ fct_relevel(Gene, gene_ord), scales = "free_y") +
  ggpubr::geom_pwc(method = "t_test", tip.length = 0, label = "p.adj.signif")
p
```


## Survival

```{r}
library(survival)
library(survminer)

survdat <- tcga_df[tcga_df$DISEASE == "LUAD", ]


fit <- survfit(Surv(OS.time, OS) ~ TP53, data = survdat)
ggsurvplot(fit, data = survdat)
```


```{r}
survdat$LUSChi <- survdat$LUSC > 0.2
survdat$SCLChi <- survdat$SCLC > 0.2

res.cox <- coxph(Surv(OS.time, OS) ~ LUAD + LUSC + SCLC + TP53 + RB1 + EGFR + tumor_stage_simple, data = survdat)


# res.cox <- coxph(Surv(OS.time, OS) ~ LUAD*TP53 + LUSC*TP53 + SCLC*TP53 + tumor_stage_simple, data=survdat[survdat$EGFR == 1,])


res.cox <- coxph(Surv(OS.time, OS) ~ LUSC + SCLC + TP53 + RB1 + EGFR + tumor_stage_simple, data = survdat)
res.cox2 <- coxph(Surv(PFI.time, PFI) ~ LUSC + SCLC + TP53 + RB1 + EGFR + tumor_stage_simple, data = survdat)

# res.cox <- coxph(Surv(OS.time, OS) ~ LUSC*TP53*RB1*EGFR + tumor_stage_simple, data=survdat)

# summary(res.cox, conf.int = FALSE)
plot(parameters(res.cox))

ggforest(res.cox, data = survdat)
ggforest(res.cox2, data = survdat)
```


```{r}
res.cox.bin <- coxph(Surv(OS.time, OS) ~ LUSChi + SCLChi + TP53 + RB1 + EGFR + tumor_stage_simple, data = survdat)
plot(parameters(res.cox.bin))
```


# CPTAC

## Load Data

```{r}
cp_meta <- read_excel(path = here("data", "cptac_meta.xlsx"), sheet = 2) %>%
  janitor::clean_names()

cp_gexp <- read_excel(path = here("data", "cptac_lung.xlsx"), sheet = "Table S2D", skip = 2)

gex_mat <- cp_gexp[69:nrow(cp_gexp), 7:ncol(cp_gexp)] %>%
  as.matrix()
gex_mat <- apply(gex_mat, MARGIN = 2, as.numeric)
rownames(gex_mat) <- cp_gexp[69:nrow(cp_gexp), 1, drop = T]

x <- colSums(apply(gex_mat, 1, is.na)) %>%
  sort(decreasing = T)

gex_mat_delog <- 2^gex_mat
```


```{r}
library(UCell)

# x <- apply(gexp_data, MARGIN = 2, FUN = function(x) {sum(x>100)})
# ranks_cptac <- UCell::StoreRankings_UCell(matrix = gex_mat_delog, maxRank = 10000, BPPARAM = BiocParallel::MulticoreParam(workers=16, progressbar = T), chunk.size = 100)
#
# scores_cptac <- UCell::ScoreSignatures_UCell(precalc.ranks = ranks_cptac, BPPARAM = BiocParallel::MulticoreParam(workers=16, progressbar = T), features = pathways, name = "")
#
# write.table(scores_cptac %>% as.data.frame() %>% rownames_to_column("sample_id"), file = gzfile(here("results", "cptac_pathway_scores.txt.gz")), quote = F, sep = "\t", row.names = F, col.names = T)
scores_cptac <- read.table(file = here("results", "cptac_pathway_scores.txt.gz"), header = T, sep = "\t")
```


```{r}
df_cp <- inner_join(cp_meta, scores_cptac) %>%
  dplyr::rename(
    "LUAD" = "LUAD_GIRARD",
    "LUSC" = "LUSC_GIRARD",
    "SCLC" = "SCLC_CCLE"
  )

mut_genes_cp <- df_cp %>%
  select(ends_with("mutation_status")) %>%
  colnames()

new_names <- str_to_upper(gsub("_mutation_status", "", mut_genes_cp))
df_cp[, new_names] <- df_cp[, mut_genes_cp]



df_cp <- df_cp %>%
  mutate(tumor_stage_simple = case_when(
    grepl("^4", stage) ~ "Stage IV",
    grepl("^3", stage) ~ "Stage III",
    grepl("^2", stage) ~ "Stage II",
    grepl("^1", stage) ~ "Stage I",
    .default = "Unknown"
  )) %>%
  mutate(tumor_stage_simple = fct_relevel(tumor_stage_simple, "Stage I"))

mut_genes_cp <- new_names
```


```{r}
x <- df_cp %>%
  group_by(type) %>%
  correlation(select = c("LUAD", "LUSC", "SCLC"), select2 = c("BIOCARTA_MAPK_PATHWAY", "REACTOME_MAPK_FAMILY_SIGNALING_CASCADES", "REACTOME_ONCOGENIC_MAPK_SIGNALING", "REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES"), method = "spearman", p_adjust = "none")
```


```{r}
df_cp %>%
  ggplot(aes(x = type, y = LUAD)) +
  geom_boxplot()
```


```{r}
df_cp %>%
  ggplot(aes(x = REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES, y = LUAD, color = type)) +
  geom_point() +
  ggpubr::stat_cor() +
  geom_smooth(method = "lm", se = F)
```

```{r}
df_cp %>%
  filter(EGFR == 1) %>%
  ggplot(aes(x = REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES, y = LUAD, color = type)) +
  geom_point() +
  ggpubr::stat_cor() +
  geom_smooth(method = "lm", se = F) +
  labs(title = "EGFR mutants")
```



```{r}
df_cp %>%
  ggplot(aes(x = REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES, y = LUSC, color = type)) +
  geom_point() +
  ggpubr::stat_cor() +
  geom_smooth(method = "lm", se = F)
```


## Model

```{r}
tests <- expand.grid(c("LUAD"), c("LUAD", "SCLC", "LUSC"))

df_cp_tum <- df_cp[df_cp$type == "Tumor", ]

df_cp_tum$purity <- df_cp_tum$tumor_purity_by_estimate_rn_aseq

res_cp <- map(1:nrow(tests), .f = function(i) {
  path <- as.character(tests[i, 2])

  p <- (colSums(df_cp_tum[, mut_genes_cp], na.rm = T) / nrow(df_cp_tum)) %>%
    enframe() %>%
    ggplot(aes(x = fct_reorder(name, value, median), y = value)) +
    geom_col() +
    guides(x = guide_axis(angle = 90))

  test_genes <- p$data$name[p$data$value > 0.05]

  mod <- glm(formula = as.formula(glue("{path} ~ {paste(test_genes, collapse = '+')} + tumor_stage_simple + purity + REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES")), family = gaussian(), data = df_cp_tum)

  param_df <- parameters(mod)
  param_df$pretty_names <- attributes(param_df)$pretty_names
  param_df <- param_df %>%
    mutate(
      DISEASE = "LUAD",
      pathway = path
    )

  return(param_df)
}) %>%
  list_rbind()

res_cp$label <- gsub("tumor_stage_simple", "", res_cp$Parameter)
res_cp$param_type <- res_cp %>%
  mutate(param_type = case_when(
    Parameter == "(Intercept)" ~ "Intercept",
    Parameter %in% mut_genes_cp ~ "Gene",
    grepl("stage", Parameter) ~ "Tumor Stage",
    .default = "Other"
  )) %>%
  pull(param_type) %>%
  factor()
```


```{r}
pl_dat <- res_cp %>%
  filter(Parameter != "(Intercept)") %>%
  filter(DISEASE == "LUAD") %>%
  mutate(Gene = Parameter)

pl_dat$padj <- p.adjust(pl_dat$p, method = "holm")

gene_ord <- pl_dat[pl_dat$pathway == "LUAD", ] %>%
  mutate(label = fct_reorder(factor(label), Coefficient, median)) %>%
  pull(label) %>%
  levels()


p <- pl_dat %>%
  ggplot(aes(x = Coefficient, y = fct_relevel(label, gene_ord), color = pathway)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = seq(0.5, n_distinct(pl_dat$Gene) - 0.5, by = 1), linetype = "dashed", alpha = 0.5) +
  geom_pointrange(aes(xmin = CI_low, xmax = CI_high), position = position_dodge(width = 0.5)) +
  # scale_x_continuous(trans = "log2", breaks = scales::breaks_log(base = 2), labels = scales::label_log(base = 2)) +
  facet_grid(param_type ~ ., scales = "free_y", space = "free_y") +
  scale_color_manual(values = cmap$histology_predominant_short) +
  # scale_y_discrete(expand = c(0,0)) +
  theme_few() +
  theme(panel.border = element_part_rect(side = "tblr", fill = NA), plot.title = element_text(hjust = 0.5, face = "bold", size = 14)) +
  labs(y = NULL, x = "Coefficient\n(Lineage Association)", color = "Lineage Signature", title = "CPTAC LUAD")

# ggsave(p, file = here("plots", "cptac_pergene_lineage_assoc.pdf"), width = 5, height = 5)
p
```



### Subet for EGFR mut and repeat

```{r}
tests <- expand.grid(c("LUAD"), c("LUAD", "SCLC", "LUSC"))

res_cp_egfr <- map(1:nrow(tests), .f = function(i) {
  path <- as.character(tests[i, 2])

  dat <- df_cp_tum[df_cp_tum$EGFR == 1, ]

  p <- (colSums(dat[, mut_genes_cp], na.rm = T) / nrow(dat)) %>%
    enframe() %>%
    ggplot(aes(x = fct_reorder(name, value, median), y = value)) +
    geom_col() +
    guides(x = guide_axis(angle = 90))

  # test_genes <- p$data$name[p$data$value > 0.05]
  # test_genes <- test_genes[test_genes != "EGFR"]

  # mod <- glm(formula = as.formula(glue("{path} ~ {paste(test_genes, collapse = '+')} + tumor_stage_simple")), family = gaussian(), data = dat)

  mod <- glm(formula = as.formula(glue("{path} ~ TP53 + RB1 + tumor_stage_simple + purity + REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES")), family = gaussian(), data = dat)



  param_df <- parameters(mod)
  param_df$pretty_names <- attributes(param_df)$pretty_names
  param_df <- param_df %>%
    mutate(pathway = path)

  return(param_df)
}) %>%
  list_rbind()

res_cp_egfr$label <- gsub("tumor_stage_simple", "", res_cp_egfr$Parameter)
res_cp_egfr$param_type <- res_cp_egfr %>%
  mutate(param_type = case_when(
    Parameter == "(Intercept)" ~ "Intercept",
    Parameter %in% mut_genes_cp ~ "Gene",
    grepl("stage", Parameter) ~ "Tumor Stage",
    .default = "Other"
  )) %>%
  pull(param_type) %>%
  factor()
```



```{r, fig.width = 5, fig.height = 4}
pl_dat <- res_cp_egfr %>%
  filter(Parameter != "(Intercept)") %>%
  # filter(DISEASE == "LUAD") %>%
  mutate(Gene = Parameter)

pl_dat$padj <- p.adjust(pl_dat$p, method = "holm")

gene_ord <- pl_dat[pl_dat$pathway == "LUAD", ] %>%
  mutate(label = fct_reorder(factor(label), Coefficient, median)) %>%
  pull(label) %>%
  levels()


p <- pl_dat %>%
  ggplot(aes(x = Coefficient, y = fct_relevel(label, gene_ord), color = pathway)) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = seq(0.5, n_distinct(pl_dat$Gene) - 0.5, by = 1), linetype = "dashed", alpha = 0.5) +
  geom_pointrange(aes(xmin = CI_low, xmax = CI_high), position = position_dodge(width = 0.5)) +
  # scale_x_continuous(trans = "log2", breaks = scales::breaks_log(base = 2), labels = scales::label_log(base = 2)) +
  facet_grid(param_type ~ ., scales = "free_y", space = "free_y") +
  scale_color_manual(values = cmap$histology_predominant_short) +
  # scale_y_discrete(expand = c(0,0)) +
  theme_few() +
  theme(
    panel.border = element_part_rect(side = "tblr", fill = NA),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
  ) +
  labs(y = NULL, x = "Coefficient\n(Lineage Association)", color = "Lineage Signature", title = "TCGA EGFR Mutant LUAD")
ggsave(p, file = here("plots", "cptac_pergene_lineage_assoc_egfr.pdf"), width = 5, height = 4)
p
```

```{r, fig.width = 5, fig.height = 4}
pl_dat <- res %>%
  filter(Parameter != "(Intercept)")

pl_dat$padj <- p.adjust(pl_dat$p, method = "holm")

gene_ord <- pl_dat[pl_dat$Parameter == "LUAD", ] %>%
  mutate(Gene = fct_reorder(Gene, Coefficient, median)) %>%
  pull(Gene) %>%
  levels()


p <- pl_dat %>%
  ggplot(aes(x = Coefficient, y = fct_relevel(Gene, gene_ord), color = pathway)) +
  geom_vline(xintercept = 1) +
  geom_hline(yintercept = seq(1.5, n_distinct(pl_dat$Gene) - .5, by = 1), linetype = "dashed", alpha = 0.5) +
  geom_pointrange(aes(xmin = CI_low, xmax = CI_high), position = position_dodge(width = 0.5)) +
  scale_x_continuous(trans = "log2", breaks = scales::breaks_log(base = 2), labels = scales::label_log(base = 2)) +
  # facet_grid(DISEASE~., scales = "free_y", space = "free_y") +
  scale_color_manual(values = cmap$histology_predominant_short) +
  theme_few() +
  labs(y = "Mutated Gene", x = "Log Odds\n(Lineage Association)", color = "Lineage Signature", title = "CPTAC Lung") +
  theme(plot.title = element_text(hjust = 0.5))
p
```


```{r, fig.width = 9, fig.height = 7}
p <- df_cp_tum %>%
  pivot_longer(cols = c("LUAD", "LUSC", "SCLC"), names_to = "Signature", values_to = "Score") %>%
  pivot_longer(cols = gene_ord, names_to = "Gene", values_to = "Mutated") %>%
  mutate(Mutated = factor(ifelse(Mutated == 0, "wt", "mut"), levels = c("wt", "mut"))) %>%
  filter(Gene %in% gene_ord, !is.na(Mutated)) %>%
  ggplot(aes(x = Mutated, y = Score)) +
  gghalves::geom_half_boxplot(outliers = F) +
  gghalves::geom_half_point_panel() +
  labs(x = NULL, title = "CPTAC Lung") +
  theme_few() +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  # geom_violin(scale = "width") +
  # geom_boxplot(width = 0.2, outliers = F) +
  facet_grid(Signature ~ fct_relevel(Gene, gene_ord), scales = "free_y") +
  ggpubr::geom_pwc(method = "wilcox_test", tip.length = 0, label = "p.adj.signif")
p
```



# Both combined plots

## Join data

```{r}
res_tcga$egfr_sub <- res_cp$egfr_sub <- "All"
res_tcga_egfr$egfr_sub <- res_cp_egfr$egfr_sub <- "EGFRmut"

res_cp$DISEASE <- res_cp_egfr$DISEASE <- "LUAD"

res_tcga_join <- rbind(res_tcga, res_tcga_egfr) %>%
  filter(
    DISEASE == "LUAD",
    # pathway == "LUAD",
    Parameter != "(Intercept)"
  ) %>%
  mutate(cohort = "TCGA")

res_cp_join <- rbind(res_cp, res_cp_egfr) %>%
  filter(Parameter != "(Intercept)") %>%
  mutate(cohort = "CPTAC")


res_join <- rbind(res_tcga_join, res_cp_join)


res_join <- left_join(res_join, impact_oncopaths, by = c("Parameter" = "Hugo_symbol"))

res_join$label[res_join$label == "Subclonal.genome.fraction"] <- "Subclonal Fraction"
res_join$label[res_join$label == "purity"] <- "Tumor Purity"
res_join$label[res_join$label == "ploidy"] <- "Tumor Ploidy"
res_join$label[res_join$label == "REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES"] <- "Reactome Nuclear MAPK"
res_join$label[res_join$group_label == "Reactome Nuclear MAPK"] <- ""

res_join$group_label <- res_join$Pathway
res_join$group_label[is.na(res_join$group_label)] <- as.character(res_join$param_type[is.na(res_join$group_label)])

res_join$group_label[res_join$group_label == "Tumor Stage"] <- "Tumor Stage\n(vs Stage I)"
```


## Model associations

```{r, fig.width = 10, fig.height = 10}
# p <- res_join %>%
#   ggplot(aes(x = Coefficient, y = fct_rev(label), color = pathway)) +
#   geom_vline(xintercept = 0) +
#   # geom_hline(yintercept = seq(0.5, n_distinct(pl_dat$Gene)-0.5, by = 1), linetype = "dashed", alpha = 0.5) +
#   geom_pointrange(aes(xmin = CI_low, xmax = CI_high), position = position_dodge(width = 0.5)) +
#   # scale_x_continuous(trans = "log2", breaks = scales::breaks_log(base = 2), labels = scales::label_log(base = 2)) +
#   facet_nested(param_type+Pathway~fct_rev(cohort)+egfr_sub, scales = "free_y", space = "free_y", switch = "y") +
#   scale_color_manual(values = cmap$histology_predominant_short) +
#   # scale_y_discrete(expand = c(0,0)) +
#   theme_few() +
#   theme(panel.border = element_part_rect(side = "tblr", fill = NA),
#         strip.placement = "outside",
#         plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
#         strip.background.x = element_part_rect(side = "b", fill = NA, color = "grey"),
#         strip.background.y = element_part_rect(side = "r", fill = NA, color = "grey"),
#         strip.clip = "off") +
#   labs(y = NULL, x = "Coefficient\n(Lineage Association)", color = "Lineage Signature")
# ggsave(p, file = here("plots", "per_gene_lineage_assoc_full.pdf"), width = 10, height = 9)
# p
```


As dotplot

```{r, fig.width=6.5, fig.height=7}
n_cptac <- nrow(df_cp_tum)
n_tcga <- nrow(tcga_df[tcga_df$DISEASE == "LUAD", ])

p <- res_join %>%
  filter(egfr_sub == "All") %>%
  mutate(cohort = fct_rev(case_when(
    cohort == "TCGA" ~ glue("TCGA ({n_tcga})"),
    cohort == "CPTAC" ~ glue("CPTAC ({n_cptac})"),
    .default = NA
  ))) %>%
  group_by(cohort, pathway) %>%
  rstatix::adjust_pvalue(p.col = "p", method = "BH") %>%
  rstatix::add_significance(
    p.col = "p.adj",
    cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1),
    symbols = c("***", "**", "*", "^", ""), output.col = "p.signif"
  ) %>%
  ggplot(aes(x = pathway, y = fct_rev(label))) +
  # ggplot(aes(x = pathway, y = fct_relevel(label, cls_coefs$label[cls_coefs$order]))) +
  geom_point(pch = 21, aes(fill = Coefficient, size = -log10(p.adj))) +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"), limits = c(-0.05, 0.05),
    oob = scales::squish,
    breaks = scales::breaks_extended(n = 3)
  ) +
  geom_text(aes(label = p.signif), size = 4, vjust = 0.8, fontface = "bold") +
  # geom_text_repel(aes(label = p.signif), size = 4, vjust = 0.8, color = "black", fontface = "bold", force = 0, bg.colour = "white", bg.r = 0) +
  # scale_x_continuous(trans = "log2", breaks = scales::breaks_log(base = 2), labels = scales::label_log(base = 2)) +
  facet_nested(fct_relevel(group_label, sort(unique(impact_oncopaths$Pathway))) ~ fct_relevel(cohort, "TCGA"), scales = "free_y", space = "free_y", switch = "y") +
  theme_few() +
  theme(
    panel.border = element_part_rect(side = "tblr", fill = NA),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    strip.clip = "off",
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 0, hjust = 1),
    strip.background.y = element_part_rect(side = "r", fill = NA, color = "black")
  ) +
  guides(
    x = guide_axis(angle = 45),
    fill = guide_colorbar(keywidth = 0.4, keyheight = 3)
  ) +
  labs(y = NULL, x = "Lineage", fill = "Coefficient\n(Lineage Association)")
ggsave(p, file = here("plots", "combined_TCGA_CPTAC_lineage_assoc_dotplot.pdf"), width = 6.5, height = 7)
p
```


```{r}
p <- res_join %>%
  filter(egfr_sub != "All") %>%
  add_count(cohort, name = "n") %>%
  mutate(cohort = fct_rev(case_when(
    cohort == "TCGA" ~ glue("TCGA ({nrow(tcga_df[tcga_df$DISEASE == 'LUAD' & tcga_df$EGFR == 1,])})"),
    cohort == "CPTAC" ~ glue("CPTAC ({nrow(df_cp_tum[df_cp_tum$EGFR == 1,])})"),
    .default = NA
  ))) %>%
  mutate(group_label = case_when(
    grepl("TP53|CELL", group_label) ~ "Gene",
    .default = group_label
  )) %>%
  group_by(cohort, pathway) %>%
  rstatix::adjust_pvalue(p.col = "p", method = "BH") %>%
  rstatix::add_significance(
    p.col = "p.adj",
    cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1),
    symbols = c("***", "**", "*", "^", ""), output.col = "p.signif"
  ) %>%
  rstatix::add_significance(
    p.col = "p",
    cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1),
    symbols = c("***", "**", "*", "^", "")
  ) %>%
  ggplot(aes(x = pathway, y = fct_rev(label))) +
  # ggplot(aes(x = pathway, y = fct_relevel(label, cls_coefs$label[cls_coefs$order]))) +
  geom_point(pch = 21, aes(fill = Coefficient, size = -log10(p.adj))) +
  scale_fill_gradient2(
    low = scales::muted("blue"), high = scales::muted("red"), limits = c(-0.05, 0.05),
    oob = scales::squish,
    breaks = scales::breaks_extended(n = 3)
  ) +
  geom_text(aes(label = p.signif), size = 4, vjust = 0.8, fontface = "bold") +
  # scale_x_continuous(trans = "log2", breaks = scales::breaks_log(base = 2), labels = scales::label_log(base = 2)) +
  facet_nested(group_label ~ fct_relevel(cohort, "TCGA"), scales = "free_y", space = "free_y", switch = "y") +
  theme_few() +
  theme(
    panel.border = element_part_rect(side = "tblr", fill = NA),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    strip.clip = "off",
    strip.placement = "outside",
    strip.text.y.left = element_text(angle = 0, hjust = 1),
    strip.background.y = element_part_rect(side = "r", fill = NA, color = "black")
  ) +
  guides(
    x = guide_axis(angle = 45),
    fill = guide_colorbar(keywidth = 0.4, keyheight = 3)
  ) +
  labs(y = NULL, x = "Lineage", fill = "Coefficient\n(Lineage Association)", title = "EGFRmut Subset")
ggsave(p, file = here("plots", "combined_TCGA_CPTAC_lineage_assoc_dotplot_egfrmut.pdf"), width = 6.5, height = 3)
p
```


```{r}
sum(tcga_df$DISEASE == "LUAD")
tcga_df[tcga_df$EGFR == 1 & tcga_df$DISEASE == "LUAD", ]
tcga_df[tcga_df$EGFR == 1 & tcga_df$DISEASE == "LUAD" & tcga_df$RB1 == 1, ]
tcga_df[tcga_df$EGFR == 1 & tcga_df$DISEASE == "LUAD" & tcga_df$TP53 == 1, ]

nrow(df_cp[df_cp$type != "NAT", ])
df_cp[df_cp$EGFR == 1, ]
df_cp[df_cp$EGFR == 1 & df_cp$RB1 == 1, ]
df_cp[df_cp$EGFR == 1 & df_cp$TP53 == 1, ]

with(df_cp, table(EGFR, RB1, TP53))
```



## MAPK cors

```{r, fig.width = 6, fig.height = 6}
df_cp_tum <- df_cp[df_cp$type != "NAT", ]
df_cp_tum$type <- NULL
df_cp_tum$DISEASE <- "LUAD"
df_cp_tum$cohort <- "CPTAC"
df_cp_tum$purity <- df_cp_tum$tumor_purity_by_estimate_rn_aseq
tcga_df$cohort <- "TCGA"
colns_common <- intersect(colnames(tcga_df), colnames(df_cp_tum))

tcga_cptac_combined <- rbind(tcga_df[, colns_common], df_cp_tum[, colns_common])


mod <- glm(LUAD ~ REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES + purity + tumor_stage_simple + EGFR, data = tcga_cptac_combined[tcga_cptac_combined$cohort == "CPTAC", ])

mapks <- grep("MAPK", colnames(tcga_cptac_combined), value = T)

p1 <- tcga_cptac_combined %>%
  filter(DISEASE %in% c("LUAD")) %>%
  ggplot(aes(x = REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES, y = LUAD)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggpubr::stat_cor(geom = "label", method = "pearson", label.x.npc = 0.05, label.y.npc = 0.05, size = 3) +
  facet_grid("All Samples" ~ cohort) +
  theme(aspect.ratio = 1) +
  labs(x = "Reactome Nuclear MAPK Signaling") +
  theme_few() +
  theme(aspect.ratio = 1)
# p1

p2 <- tcga_cptac_combined %>%
  filter(DISEASE %in% c("LUAD"), EGFR == 1) %>%
  ggplot(aes(x = REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES, y = LUAD)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggpubr::stat_cor(geom = "label", method = "pearson", label.x.npc = 0.05, label.y.npc = 0.05, size = 3) +
  facet_grid("EGFRmut Subset" ~ cohort) +
  theme(aspect.ratio = 1) +
  labs(x = "Reactome Nuclear MAPK Signaling") +
  theme_few() +
  theme(aspect.ratio = 1)
# p2


pcomb <- p1 + p2 + plot_layout(ncol = 1, axis_titles = "collect") &
  theme(
    axis.line.x = element_blank(), axis.line.y = element_blank(), axis.line = element_blank(),
    panel.border = element_rect(fill = NA, linewidth = 1)
  ) &
  labs(y = "LUAD Signature")
pcomb
ggsave(pcomb, file = here("plots", "tcga_cptac_mapk_cors.pdf"), width = 6, height = 6)
```


```{r}
p1 <- tcga_df %>%
  filter(DISEASE %in% c("LUAD")) %>%
  ggplot(aes(x = REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES, y = LUAD)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggpubr::stat_cor() +
  theme(aspect.ratio = 1) +
  labs(x = "MAPK", title = "TCGA")
p1

p1b <- tcga_df %>%
  filter(DISEASE %in% c("LUAD"), EGFR == 1) %>%
  ggplot(aes(x = REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES, y = LUAD)) +
  geom_point() +
  geom_smooth(method = "lm") +
  ggpubr::stat_cor() +
  theme(aspect.ratio = 1) +
  labs(x = "MAPK", title = "TCGA")
p1b

p2 <- df_cp %>%
  # filter(EGFR == 1) %>%
  ggplot(aes(x = REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES, y = LUAD, color = type)) +
  geom_point() +
  ggpubr::stat_cor() +
  geom_smooth(method = "lm", se = F) +
  labs(title = "CPTAC")
p2

p1 + p2 & labs(x = "MAPK")
```

```{r}
df_cp %>%
  filter(EGFR == 1) %>%
  ggplot(aes(x = REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES, y = LUAD, color = type)) +
  geom_point() +
  ggpubr::stat_cor() +
  geom_smooth(method = "lm", se = F) +
  labs(title = "EGFR mutants")
```

