---
title: "Main plotting script"
author: "Matthew Zatzman"
---

# Setup

```{r setup}
library(tidyverse)
library(here)
library(ggbeeswarm)
library(ggsci)
library(ggthemes)
library(glue)
library(patchwork)
library(cowplot)
library(ggnewscale)
library(ggh4x)
library(ggtext)
library(ComplexHeatmap)
library(scCustomize)
library(SingleCellExperiment)
# library(BPCells)
library(Seurat)
library(dittoSeq)
# library(SeuratObject)
# library(SeuratDisk)
library(scatools)
raster.dpi <- c(1000, 1000)

devtools::load_all()

symnum.args <- list(cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, Inf), symbols = c("p<0.0001", "p<0.001", "p<0.01", "p<0.05", "ns"))
symnum_0.1 <- list(cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 0.1, Inf), symbols = c(
  "p<0.0001",
  "p<0.001", "p<0.01", "p<0.05", "p<0.1", "ns"
))
symnum_0.1_star <- list(cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 0.1, Inf), symbols = c(
  "****",
  "***", "**", "*", "^", "ns"
))
```

# Load Data

Load database object

```{r}
db <- load_project()
```


```{r}
srt <- db$srt
srt_epi <- db$srt_epi
```

## Pathways

```{r}
pathways <- read.table(here("metadata", "pathways", "egfr_pathways_annotations.txt"), header = T, sep = "\t")
mapk_paths <- filter(pathways, category == "MAPK")$pathway
tum_paths <- c("LUAD", "LUSC", "SCLC")

srt_tum$LUAD <- srt_tum$LUAD_GIRARD
srt_tum$LUSC <- srt_tum$LUSC_GIRARD
srt_tum$SCLC <- srt_tum$SCLC_CCLE
srt_tum$MAPK <- srt_tum$REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES

srt_epi$LUAD <- srt_epi$LUAD_GIRARD
srt_epi$LUSC <- srt_epi$LUSC_GIRARD
srt_epi$SCLC <- srt_epi$SCLC_CCLE
srt_epi$MAPK <- srt_epi$REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES

pathways$pathway[pathways$pathway == "LUAD_GIRARD"] <- "LUAD"
pathways$pathway[pathways$pathway == "LUSC_GIRARD"] <- "LUSC"
pathways$pathway[pathways$pathway == "SCLC_CCLE"] <- "SCLC"

pathways_v2 <- read.table(here("metadata", "pathways", "egfr_pathways_v2.txt"), header = T, sep = "\t") # Additional pathways
```



# Plotting

## Figure 1 -- Swimmers plot

### Main panel

```{r, fig.width = 7, fig.height = 15}
# SETTINGS
tri_stroke <- 0.25
tri_size <- 2
tri_nudge <- 0.275
######
swimmers_df <- db$metadata %>%
  mutate(dx_date = 0) %>%
  select(sample_id = sample_id_new, dmp_id, dx_date, procedure, osi_start, osi_end, osi_lot, other_treatment, other_treatment_start, other_treatment_end, other_treatment_lot, other_same_osi, radio_prog, met_date, impact, last_tp, last_followup_status, time_point, histology, has_impact) %>%
  mutate(
    # Category to split the timing up
    t_group = ifelse(last_tp >= 60, ">5years", "<5years"),
    last_followup_status = fct_recode(last_followup_status,
      "Alive" = "ALIVE",
      "Deceased" = "DECEASED"
    )
  )

# Pull out treatment information into single table
treatments_df <- swimmers_df %>%
  mutate(treatment = "Osimertinib") %>%
  filter(!is.na(osi_start)) %>%
  select(sample_id, dmp_id, treatment, time_point, last_tp, treatment_start = osi_start, treatment_end = osi_end, lot = osi_lot)

other_treatments_df <- swimmers_df %>%
  filter(other_same_osi == "different") %>%
  select(sample_id, dmp_id, treatment = other_treatment, time_point, last_tp, treatment_start = other_treatment_start, treatment_end = other_treatment_end, lot = other_treatment_lot) %>%
  # Clean up treatment name
  mutate(treatment = gsub(" |rechallenge|now with pemetrexed \\(for metastatic KRAS mutated LC\\)|Breast cancer treatment: |S\\/p| \\(peme maintenance\\)|\\(peme/bev maintenance\\)", "", treatment)) %>%
  separate_longer_delim(cols = treatment, delim = c("+")) %>%
  # separate_longer_delim(cols = treatment, delim = c("/")) %>%
  mutate(
    treatment_orig = treatment,
    treatment = case_when(
      grepl("pembro", treatment, ignore.case = T) ~ "Chemo+IO",
      grepl("bev", treatment, ignore.case = T) ~ "Chemo+Bev",
      grepl("carb|peme|pacli", treatment, ignore.case = T) ~ "Chemo",
      grepl("Osi", treatment, ignore.case = T) ~ "Osimertinib",
      .default = treatment
    ),
    treatment_simple = case_when(
      grepl("Chemo\\+", treatment) ~ "Chemo combo",
      .default = treatment
    )
  )

treatments_df <- plyr::rbind.fill(treatments_df, other_treatments_df)

# Bind the additional EGFR TKI data
other_treatments_egfr <- read.table(file = here("metadata", "other_tkis_update_mar_2025.txt"), header = T, sep = "\t") %>%
  dplyr::rename(
    treatment = AGENT,
    treatment_start = trt_start,
    treatment_end = trt_end
  ) %>%
  select(sample_id, treatment, treatment_start, treatment_end) %>%
  left_join(db$metadata[, c("sample_id", "sample_id_new", "time_point", "last_tp", "dmp_id")]) %>%
  select(!sample_id) %>%
  dplyr::rename(sample_id = sample_id_new) %>%
  mutate(treatment_simple = treatment)
treatments_df <- plyr::rbind.fill(treatments_df, other_treatments_egfr)

osi_df <- treatments_df %>%
  filter(treatment == "Osimertinib")
other_treatments_df <- treatments_df %>%
  filter(treatment != "Osimertinib")


# Backgrounds for each timepoint facet and texts
backgrounds <- map(levels(db$metadata$time_point), .f = function(tp) {
  element_part_rect(side = "r", fill = colors$time_point[[tp]])
})
texts <- map(levels(db$metadata$time_point), .f = function(tp) {
  element_text(face = "bold", color = "black")
})

# Facet relabeller
tp_names <- c(
  "TN" = "Treatment-Naive",
  "MRD" = "Residual Disease",
  "PD" = "Progressive Disease"
)


# backgrounds <- list(element_part_rect(side = "r", fill = colors$time_point[["Pre"]]),
# element_part_rect(side = "r", fill = colors$time_point[["On"]]),
# element_part_rect(side = "r", fill = colors$time_point[["PD"]]))

# TO NOTE: Some values for met date and impact are <0. These are set to zero to allow the x axis sqrt transform.
swimmers_df$met_date[swimmers_df$met_date < 0] <- 0
swimmers_df$impact[swimmers_df$impact < 0] <- 0

swimmers_plot <- swimmers_df %>%
  mutate(sample_id = fct_reorder(sample_id, last_tp, max)) %>%
  ggplot(aes(y = sample_id)) +
  # LAST FOLLOWUP
  geom_segment(aes(x = dx_date, xend = last_tp), linewidth = 1, color = "black") +
  # OSI INTERVAL
  geom_segment(data = osi_df, linewidth = 4, aes(x = treatment_start, xend = treatment_end, color = "Osimertinib"), position = position_nudge(y = 0)) +
  # OTHER TREATMENTS
  geom_segment(data = other_treatments_df, linewidth = 1.5, aes(x = treatment_start, xend = treatment_end, color = treatment_simple)) +
  scale_color_manual(
    na.translate = F,
    name = "Treatments",
    values = c(
      "Osimertinib" = "#a5c1f2",
      "Erlotinib" = "#0072B5FF",
      "Dacomitinib" = "#3B96E2",
      "Afatinib" = "#8B919C",
      "Chemo combo" = "#BC3C29FF",
      "Chemo" = "#FFDC91FF",
      "Crizotinib" = "#20854EFF",
      "TDM1" = "#EE4C97FF"
    ),
    limits = c("Osimertinib", "Erlotinib", "Dacomitinib", "Afatinib", "Chemo combo", "Chemo", "Crizotinib", "TDM1"),
    guide = guide_legend(order = 3)
  ) +
  # LAST STATUS
  geom_point(aes(x = last_tp, shape = last_followup_status, fill = last_followup_status), size = 2.5, stroke = 0.2) +
  scale_shape_manual(
    name = "Last followup", values = c("Alive" = 21, "Deceased" = 22),
    guide = guide_legend(order = 4)
  ) +
  scale_fill_manual(
    name = "Last followup", values = c("Alive" = "#35aaf2", "Deceased" = "#ab2230"),
    guide = guide_legend(order = 4)
  ) +
  new_scale_fill() +
  new_scale("shape") +
  # EVENTS
  ## MET TIMEPOINT
  geom_point(aes(x = met_date, fill = "Metastasis"), size = tri_size, stroke = tri_stroke, position = position_nudge(y = tri_nudge), pch = 25) +
  ## Date of radiographic progression
  geom_point(aes(x = radio_prog, fill = "Radiographic Progression"), size = tri_size, stroke = tri_stroke, position = position_nudge(y = tri_nudge), pch = 25) +
  scale_fill_manual(
    name = "Clinical Event",
    values = c(
      "Metastasis" = "red",
      "Radiographic Progression" = "orange"
    ), guide = guide_legend(override.aes = list(size = 3), order = 2),
    limits = c("Metastasis", "Radiographic Progression")
  ) +
  new_scale_fill() +
  ## IMPACT SAMPLE
  geom_point(aes(x = impact, fill = "IMPACT"), size = tri_size, stroke = tri_stroke, position = position_nudge(y = -tri_nudge), pch = 24) +
  ## Procedure sample
  geom_point(aes(x = procedure, fill = "snRNA"), size = tri_size, stroke = tri_stroke, position = position_nudge(y = 0), pch = 23) +
  scale_fill_manual(
    name = "Molecular Assay",
    values = c(
      "snRNA" = "#33589c",
      "IMPACT" = "black"
    ), guide = guide_legend(override.aes = list(size = 3), order = 1),
    limits = c("snRNA", "IMPACT")
  ) +
  # AXIS LABELS
  labs(x = "Time from initial diagnosis (Years)", y = NULL) +
  # scale_x_break(breaks = c(150, 320), scales = 0.25) +
  scale_x_sqrt(
    breaks = c(0, 1, 2, 3, 4, 5, 6, 8, 10, 15, 20, 25, 30),
    labels = c(0, 1, 2, 3, 4, 5, 6, 8, 10, 15, 20, 25, 30),
    sec.axis = sec_axis(~.,
      name = "Time from initial diagnosis (Years)",
      breaks = c(1e-4, 1, 2, 3, 4, 5, 6, 8, 10, 15, 20, 25, 30),
      labels = c(0, 1, 2, 3, 4, 5, 6, 8, 10, 15, 20, 25, 30)
    )
  ) +
  guides(
    # x = guide_axis(cap = "upper"),
    # x.sec = guide_axis(cap = "upper"),
    # y = guide_axis(cap = "both")
  ) +
  facet_grid2(time_point ~ .,
    scales = "free", space = "free_y", switch = "y",
    strip = strip_themed(
      background_y = backgrounds,
      text_y = texts
    ),
    labeller = as_labeller(tp_names)
  ) +
  theme_clean() +
  theme(
    # strip.text.y = element_blank(),
    panel.border = element_part_rect(side = "tblr", fill = NA),
    axis.line.x = element_blank(),
    axis.line.y = element_blank(),
    # axis.title.x = element_text(hjust = 0.5),
    strip.background = element_part_rect(side = "r", fill = NA, color = "black", linewidth = 1, linetype = "solid"),
    # panel.grid.major.y = element_line(),
    # panel.grid.major.x = element_line(colour = "gray", linetype = "dotted"),
    # panel.spacing = unit(0.1, "lines"),
    plot.background = element_blank(),
    # axis.line.x.top = element_blank(),
    strip.placement = "outside",
    strip.clip = "off",
    legend.position = "left"
    # axis.text.x.top = element_blank(),
    # axis.ticks.x.top = element_blank()
    # panel.grid.major.x = element_line(linetype = "dashed", color = "grey50", linewidth = 0.2)
  ) +
  coord_cartesian(clip = "off")

swimmers_plot
```


### Side annotations

```{r, fig.width = 7, fig.height = 9}
meta_plot <- db$metadata %>%
  # Add WGD annotation
  left_join(db$impact_data$facets_data[, c("dmp_pid", "wgd", "ploidy")], by = c("dmp_id" = "dmp_pid")) %>%
  mutate(sample_id = fct_reorder(sample_id_new, last_tp, max)) %>%
  ggplot(aes(y = sample_id)) +
  # geom_tile(aes(x = "Mechanism of Resistance", fill = mor), color = "black", linewidth = 0.5) +
  # scale_fill_manual(
  #   name = "Mechanism of Resistance",
  #   values = colors$mor
  # ) +
  # AGE
  geom_tile(aes(x = "Age at diagnosis", fill = age_at_intial_diagnosis), color = "black", linewidth = 0.5) +
  scale_fill_gradientn(
    name = "Age at Diagnosis", colours = viridis::mako(10),
    guide = guide_colorbar(order = 2, position = "bottom", keywidth = 5, keyheight = 0.7)
  ) +
  new_scale_fill() +
  # HISTOLOGY
  geom_tile(aes(x = "Clinical Histology", fill = histology_predominant), color = "black", linewidth = 0.5) +
  scale_fill_manual(
    name = "Clinical Histology", values = colors$histology_predominant,
    breaks = c(
      "Lung Adenocarcinoma",
      "Lung Squamous Cell Carcinoma",
      "Small Cell Lung Cancer",
      "Poorly Differentiated"
    ),
    guide = guide_legend(order = 1, position = "bottom", keywidth = 0.7, keyheight = 1, ncol = 2)
  ) +
  new_scale_fill() +
  # TUMOR STAGE
  geom_tile(aes(x = "Stage of sample", fill = stage_of_tissue_simple), color = "black", linewidth = 0.5) +
  geom_tile(aes(x = "Stage at diagnosis", fill = stage_at_initial_diagnosis_simple), color = "black", linewidth = 0.5) +
  scale_fill_viridis_d(name = "Tumor Stage", guide = guide_legend(order = 4, position = "bottom", keywidth = 0.7, keyheight = 1, ncol = 2)) +
  new_scale_fill() +
  # TISSUE SITE
  geom_tile(aes(x = "Tissue Site", fill = site_of_tissue_simple), color = "black", linewidth = 0.5) +
  scale_fill_manual(
    name = "Tissue Site", values = colors$site_of_tissue_simple, breaks = names(colors$site_of_tissue_simple),
    guide = guide_legend(order = 3, keywidth = 0.7, keyheight = 1, ncol = 2, position = "bottom")
  ) +
  new_scale_fill() +
  geom_tile(aes(x = "Sex", fill = sex), color = "black", linewidth = 0.5) +
  scale_fill_manual(
    name = "Sex", values = colors$sex, breaks = names(colors$sex), na.value = "white",
    guide = guide_legend(order = 6, keywidth = 0.7, keyheight = 1, ncol = 1, position = "bottom")
  ) +
  new_scale_fill() +
  # BINARY (IMPACT, TP53, RB1)
  geom_tile(aes(x = "IMPACT", fill = has_impact), color = "black", linewidth = 0.5, show.legend = F) +
  geom_tile(aes(x = "Smoking History", fill = smoking_history_simple), color = "black", linewidth = 0.5, show.legend = F) +
  geom_tile(aes(x = "WGD", fill = wgd), color = "black", linewidth = 0.5, show.legend = F) +
  geom_tile(aes(x = "TP53mut", fill = TP53mut), color = "black", linewidth = 0.5, show.legend = F) +
  geom_tile(aes(x = "RB1mut", fill = RB1mut), color = "black", linewidth = 0.5, show.legend = F) +
  scale_fill_manual(
    name = "IMPACT", values = colors$tf, breaks = names(colors$tf), na.value = "white",
    guide = guide_legend(order = 5)
  ) +
  new_scale_fill() +
  # scale_fill_pander(name = "Histology") +
  facet_nested(time_point ~ ., scales = "free", space = "free_y") +
  scale_y_discrete(position = "right", expand = c(0, 0)) +
  scale_x_discrete(
    expand = c(0, 0),
    limits = c("Clinical Histology", "Age at diagnosis", "Tissue Site", "Stage at diagnosis", "Stage of sample", "IMPACT", "WGD", "TP53mut", "RB1mut", "Smoking History", "Sex")
  ) +
  guides(
    x = guide_axis(angle = 55),
    x.sec = guide_axis(angle = 90)
  ) +
  labs(x = NULL, y = NULL) +
  theme_clean() +
  theme(
    strip.text.y = element_blank(),
    panel.border = element_blank(),
    plot.background = element_blank(),
    axis.line.x = element_blank(),
    axis.line.y = element_blank(),
    panel.grid.major.y = element_blank(),
    # panel.spacing = unit(0.1, "lines"),
    legend.key.height = unit(1, "lines"),
    legend.key.width = unit(0.6, "lines"),
    legend.title.position = "top", legend.title = element_text(hjust = 0.5)
    # panel.grid.major.x = element_line(linetype = "dashed", color = "grey50", linewidth = 0.2)
  )
meta_plot

# meta_leg <- ggpubr::as_ggplot(ggpubr::get_legend(swimmers_plot))

# meta_plot <- meta_plot + theme(legend.position = "none")
```


### EGFR de novo vs secondary

```{r, fig.width = 9, fig.height = 5}
egfr_hits <- read.table(file = here("results", "egfr_hits_table.txt"), header = T, sep = "\t") %>%
  left_join(db$metadata)
egfr_hits <- egfr_hits %>%
  mutate(type_name = paste(type, egfr_cat, sep = "_"))

egfr_mat <- egfr_hits %>%
  mutate(value = 1) %>%
  pivot_wider(id_cols = c("sample_id_new"), names_from = "variant", values_from = value, values_fill = 0) %>%
  column_to_rownames("sample_id_new") %>%
  t() %>%
  memoSort()

# egfr_hits %>%
#   ggplot(aes(x = fct_relevel(sample_id_old, colnames(egfr_mat)), y = fct_rev(fct_relevel(variant, rownames(egfr_mat))))) +
#   geom_tile() +
#   facet_grid(.~time_point, space = "free", scales = "free") +
#   guides(x = guide_axis(angle = 90)) +
#   labs(x = "Sample ID", y = "EGFR Variant") +
#   ggthemes::theme_pander() +
#   theme(panel.border = element_rect(fill = NA, color = "black"))

egfr_hits %>%
  ggplot(aes(x = fct_relevel(sample_id_new, colnames(egfr_mat)), y = fct_rev(fct_relevel(variant, rownames(egfr_mat))))) +
  geom_tile() +
  facet_nested(type ~ time_point, space = "free", scales = "free") +
  guides(x = guide_axis(angle = 90)) +
  labs(x = "Sample ID", y = "EGFR Mutation") +
  theme_pander() +
  theme(panel.border = element_rect(fill = NA, color = "black"), strip.clip = "off")

# library(legendry)
# egfr_hits %>%
#   ggplot(aes(x = fct_relevel(sample_id, colnames(egfr_mat)), y = interaction(variant, egfr_cat))) +
#   geom_tile() +
#   facet_nested(type ~ time_point, space = "free", scales = "free") +
#   guides(x = guide_axis(angle = 90), y = "axis_nested") +
#   labs(x = "Sample ID", y = "EGFR Mutation") +
#   theme_pander() +
#   theme(panel.border = element_rect(fill = NA, color = "black"), strip.clip = "off")
```


#### Vertical and aligned

```{r, fig.width = 5, fig.height = 9}
sample_order <- swimmers_df %>%
  mutate(sample_id = fct_reorder(sample_id, last_tp, max)) %>%
  pull(sample_id) %>%
  levels() %>%
  as.character()

var_order <- c(
  "Primary_Exon 19", "Primary_L858R", "Secondary_Amplification",
  "Primary_G719X", "Primary_T790M", "Secondary_C797S", "Primary_Other"
)

egfr_hits$sample_id_new <- fct_relevel(egfr_hits$sample_id_new, sample_order)

egfr_hit_plot <- egfr_hits %>%
  # mutate(denovo = factor(ifelse(grepl("De novo", type_name), "Pre-Osi", "Post-Osi"),
  #   levels = c("Pre-Osi", "Post-Osi")
  # )) %>%
  ggplot(aes(y = sample_id_new, x = fct_relevel(type_name, var_order))) +
  geom_tile() +
  facet_nested(time_point ~ "EGFR Mutation" + type, space = "free", scales = "free", switch = "x", strip = strip_nested(clip = "off")) +
  scale_x_discrete(breaks = var_order, labels = gsub("Primary_|Secondary_", "", var_order), expand = c(0, 0)) +
  scale_y_discrete(position = "right", expand = c(0, 0)) +
  guides(x = guide_axis(angle = 55)) +
  labs(y = "Sample ID", x = NULL) +
  theme_clean() +
  theme(axis.line.x = element_blank(), axis.line.y = element_blank(), panel.border = element_rect(fill = NA, color = "black"), plot.background = element_blank(), strip.placement = "outside", strip.background = element_part_rect(side = "l", color = "black", fill = NA, linetype = "solid"), strip.background.x = element_part_rect(side = "t", color = "black", fill = NA, linetype = "solid"))
# p1

# Instead do as proportion of the timepoints
df_egfr_cat_prop <- egfr_hits %>%
  add_count(time_point, type_name, name = "n_type") %>%
  complete(time_point, type_name, fill = list(n_type = 0)) %>%
  add_count(time_point, name = "n_tp") %>%
  mutate(prop = n_type / n_tp) %>%
  distinct(time_point, type_name, time_point, n_type, n_tp, prop, type) %>%
  filter(!is.na(type)) %>%
  mutate(time_point = fct_relevel(time_point, levels(db$metadata$time_point)))
# %>%
# filter()


egfr_hit_plot_top <- df_egfr_cat_prop %>%
  ggplot(aes(x = (fct_relevel(type_name, var_order)))) +
  geom_col(aes(fill = time_point, y = prop), position = "dodge") +
  scale_x_discrete(breaks = var_order, labels = gsub("Primary_|Secondary_", "", var_order)) +
  scale_y_continuous(position = "right") +
  facet_nested(. ~ "EGFR Mutation" + type, space = "free", scales = "free", switch = "y", strip = strip_nested(clip = "off")) +
  scale_fill_manual(values = colors$time_point) +
  guides(
    fill = guide_legend(keywidth = 0.7, keyheight = 1, position = "right"),
    x = guide_axis(cap = "both", angle = 55),
    y = guide_axis(cap = "both")
  ) +
  theme_clean() +
  theme(plot.background = element_blank(), axis.line.y.right = element_line(), strip.placement = "outside", strip.background.x = element_part_rect(linetype = "solid", side = "b", color = "black", fill = NA)) +
  labs(y = "Proportion", x = NULL, fill = "Timepoint")

pegfr <- (egfr_hit_plot_top + theme(plot.margin = margin(r = 3, unit = "lines"))) + egfr_hit_plot + plot_layout(ncol = 1, heights = c(0.125, 1)) & theme(panel.spacing = unit(0.2, "lines"))
pegfr
```



### Final Swimmers Combined

```{r, fig.width = 11, fig.height = 13}
# pcomb <- swimmers_plot + meta_plot + plot_layout(widths = c(1, 0.5))
pcomb <- plot_grid(swimmers_plot, meta_plot, align = "h", nrow = 1, ncol = 2, rel_widths = c(1, 0.9))
pcomb
# ggsave(pcomb, file = here("plots", "swimmers_plus_annos.pdf"), width = 11, height = 13)
```

### Full version

grab legend

```{r, fig.width = 5, fig.height = 13}
lgds <- ggpubr::get_legend(meta_plot)
lgd <- ggpubr::as_ggplot(lgds)


ggsave(lgd, file = here("plots", "swimmers_anno_legend.pdf"), width = 14, height = 2)
```


##### Plot

```{r, fig.width = 11, fig.height = 13}
spc <- margin(t = 0.2, r = 0.2, b = 0.2, l = 0.2, unit = "lines")
empty_panel <- plot_spacer() + theme(plot.margin = spc)
pcomb_ext <- plot_grid(
  plotlist = list(
    # empty_panel,
    # empty_panel,
    # egfr_hit_plot_top + theme(plot.margin = spc),
    swimmers_plot + theme(plot.margin = spc) + coord_cartesian(clip = "off"),
    (meta_plot + theme(
      legend.position = "none",
      plot.margin = spc,
      axis.text.y.right = element_blank(),
      axis.ticks.y.right = element_blank(),
      axis.text.x.bottom = element_text(angle = 55),
      strip.clip = "off"
    ) + coord_cartesian(clip = "off")),
    egfr_hit_plot + theme(
      plot.margin = spc, axis.ticks = element_line(color = "black"),
      strip.background.y.right = element_blank(),
      strip.text.y.right = element_blank(),
      axis.title.y.right = element_blank(),
      axis.text.x.bottom = element_text(angle = 55)
    ) + coord_cartesian(clip = "off")
  ),
  align = "hv",
  nrow = 1,
  ncol = 3,
  rel_widths = c(1, 0.32, 0.3),
  # rel_heights = c(0.25, 1),
  axis = "tb"
)


# pcomb_ext <- (plot_spacer() / swimmers_plot) | (plot_spacer() / meta_plot +   guides(x.sec = guide_axis()))

pcomb_ext

ggsave(pcomb_ext, file = here("plots", "swimmers_plus_annos_muts.pdf"), width = 10, height = 14.3)


# (egfr_hit_plot_top + theme(plot.margin = margin(r = 3, unit = "lines"))) + egfr_hit_plot + plot_layout(ncol = 1, heights = c(0.125, 1)) & theme(panel.spacing = unit(0.2, "lines"))
#
```

##### Top Proportion

```{r, fig.width = 3.5, fig.height = 2.5}
egfr_hit_plot_top_final <- egfr_hit_plot_top +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.line.x = element_blank(), strip.clip = "off", plot.margin = margin(t = 0.25, b = 1.5, l = 0.25, r = 0.25, unit = "lines"), strip.background = element_blank()) +
  coord_cartesian(clip = "off") +
  guides(x = guide_axis(angle = 35, cap = "both"))
egfr_hit_plot_top_final
ggsave(egfr_hit_plot_top_final, file = here("plots", "swimmers_mut_prop_top.pdf"), width = 3.5, height = 2.5)
```

## Figure 2 -- Cohort Overview

### Main Plots

#### Unbatch corrected

```{r, fig.width = 14, fig.height = 7}
# Patient
p1 <- DimPlot(srt, reduction = "umap_nb", group.by = "site_of_tissue_simple") +
  umap_theme() +
  scale_color_manual(values = colors$site_of_tissue_simple, breaks = names(colors$site_of_tissue_simple), labels = label_num_generator(srt$site_of_tissue_simple)) +
  NoAxes() +
  labs(title = "Tissue site") +
  theme(plot.margin = margin(1, 1, 1, 1, "mm"))

p2 <- DimPlot(srt, reduction = "umap_nb", group.by = "sample_id_new") +
  umap_theme() +
  scale_color_manual(values = colors$sample_id_new, breaks = names(colors$sample_id_new), labels = label_num_generator(srt$sample_id_new)) +
  guides(color = guide_legend(ncol = 3, override.aes = list(size = 3))) +
  NoAxes() +
  labs(title = "Patient") +
  theme(plot.margin = margin(8, 1, 1, 1, "mm"))

p3 <- DimPlot(srt, reduction = "umap_nb", group.by = "time_point", order = "MRD") +
  umap_theme() +
  NoAxes() +
  scale_color_manual(values = colors$time_point, breaks = names(colors$time_point), labels = label_num_generator(srt$time_point)) +
  labs(title = "Timepoint") +
  theme(plot.margin = margin(1, 1, 1, 1, "mm"))

# Cell types
p4 <- DimPlot(srt, group.by = "cell_type", reduction = "umap_nb") +
  umap_theme() +
  NoAxes() +
  scale_color_manual(values = colors$cell_type, breaks = names(colors$cell_type), labels = label_num_generator(srt$cell_type)) +
  labs(title = "Cell Type") +
  theme(plot.margin = margin(1, 1, 1, 1, "mm"))
# ggsave(cohort_umap, file = here("plots", "cohort_umap_cell_types_nb.pdf"), width = 6, height = 5)


pcomb <- p4 + p2 + p1 + p3 + plot_layout(ncol = 2) & theme(aspect.ratio = 1, legend.box.margin = margin(), legend.margin = margin())
# pcomb
ggsave(pcomb, file = here("plots", "cohort_umap_combined_nobatch.pdf"), width = 14, height = 8)
```

#### Batch corrected

```{r, fig.width = 14, fig.height = 7}
# Patient
p1 <- DimPlot(srt, reduction = "umap", group.by = "site_of_tissue_simple", shuffle = T) +
  umap_theme() +
  scale_color_manual(values = colors$site_of_tissue_simple, breaks = names(colors$site_of_tissue_simple), labels = label_num_generator(srt$site_of_tissue_simple)) +
  NoAxes() +
  labs(title = "Tissue site")

p2 <- DimPlot(srt, reduction = "umap", group.by = "sample_id_new") +
  umap_theme() +
  scale_color_manual(values = colors$sample_id_new, breaks = names(colors$sample_id_new), labels = label_num_generator(srt$sample_id_new)) +
  NoAxes() +
  labs(title = "Patient")

p3 <- DimPlot(srt, reduction = "umap", group.by = "time_point", order = "On") +
  umap_theme() +
  NoAxes() +
  scale_color_manual(values = colors$time_point, breaks = names(colors$time_point), labels = label_num_generator(srt$time_point)) +
  labs(title = "Timepoint")

# Cell types
p4 <- DimPlot(srt, group.by = "cell_type", reduction = "umap") +
  umap_theme() +
  NoAxes() +
  scale_color_manual(values = colors$cell_type, breaks = names(colors$cell_type), labels = label_num_generator(srt$cell_type)) +
  labs(title = "Cell Type")
# ggsave(cohort_umap, file = here("plots", "cohort_umap_cell_types_nb.pdf"), width = 6, height = 5)


pcomb <- wrap_plots(p4, p2, p1, p3) & theme(aspect.ratio = 1)
pcomb
ggsave(pcomb, file = here("plots", "cohort_umap_combined_batch.pdf"), width = 14, height = 7)

# For supplement: patient and tissue site
pcomb2 <- wrap_plots(p2, p1, byrow = F) + plot_layout(nrow = 1) & theme(aspect.ratio = 1, plot.margin = margin(1, 1, 1, 1, "mm"))
pcomb2
ggsave(pcomb2, file = here("plots", "cohort_umap_combined_batch_supp.pdf"), width = 15, height = 5)
```

Select for main

```{r, fig.width = 9, fig.height = 3}
p3 <- DimPlot(srt, reduction = "umap", group.by = "time_point", order = "On") +
  umap_theme() +
  NoAxes() +
  scale_color_manual(values = colors$time_point, breaks = names(colors$time_point)) +
  labs(title = "Timepoint")

# Cell types
p4 <- DimPlot(srt, group.by = "cell_type", reduction = "umap") +
  umap_theme() +
  NoAxes() +
  scale_color_manual(values = colors$cell_type, breaks = names(colors$cell_type)) +
  labs(title = "Cell Type")
# ggsave(cohort_umap, file = here("plots", "cohort_umap_cell_types_nb.pdf"), width = 6, height = 5)
#
cohort_umap_tumor_cells <- DimPlot(srt, group.by = "is_tumor_cell", cols = list("TRUE" = as.character(colors$cell_type["Tumor Cell"]), "FALSE" = "grey90")) +
  labs(title = NULL) +
  # theme_void() +
  theme(aspect.ratio = 1, legend.key.spacing = unit(0.1, "lines")) +
  umap_theme() +
  NoAxes() +
  NoLegend() +
  labs(title = "Tumor Cells")
# ggsave(cohort_umap_tumor_cells, file = here("plots", "cohort_umap_tumor_cells.pdf"), width = 6, height = 5)
# cohort_umap_tumor_cells



pcomb <- wrap_plots(p4, p3, byrow = F) + plot_layout(nrow = 1) & theme(aspect.ratio = 1, plot.margin = margin(1, 1, 1, 1, "mm"))
pcomb
ggsave(pcomb, file = here("plots", "cohort_umap_combined_batch_main2.pdf"), width = 8, height = 3)
```


### Cell Type Markers

```{r, fig.width = 9, fig.height = 3}
ct_markers <- markers$cell_type_markers

names(ct_markers) <- map(names(ct_markers), .f = function(ct) {
  ct_col <- colors$cell_type[ct]
  ct_colname <- glue("<b style='color:{ct_col}'>{ct}</b>")
  return(ct_colname)
}) %>%
  list_c()

library(ggtext)

srt$cell_type_col <- factor(srt$cell_type, levels = levels(srt$cell_type), labels = glue("<b style='color:{colors$cell_type[levels(srt$cell_type)]}'>{levels(srt$cell_type)}</b>"))
# srt$cell_type_col <- fct_rev(srt$cell_type_col)


pl <- DotPlotter(srt, features = ct_markers, group.by = "cell_type_col", cluster_groups = F, rotate_y_strip = F) +
  theme(
    strip.clip = "off",
    axis.line = element_blank(),
    strip.text.x = element_markdown(),
    axis.text.y = element_markdown()
  )

ggsave(pl, file = here("plots", "cell_type_markers_dotplot.pdf"), width = 9, height = 3)
pl
```


### Specificity plots

Check all four

```{r, fig.width = 6, fig.height = 6}
cls_order <- levels(fct_reorder(db$spec$cell_type$cid2, db$spec$cell_type$cid1_purity, median))

spec_pls <- map(names(db$spec), .f = function(l) {
  df <- db$spec[[l]]

  p1 <- df %>%
    # group_by(cid1, cid2) %>%
    # summarize(cid1_purity = mean(cid1_purity)) %>%
    filter(cid2 != "Other") %>%
    ggplot(aes(x = fct_relevel(cid2, cls_order), y = cid1_purity)) +
    geom_violin(scale = "width", aes(fill = cid2), show.legend = F) +
    scale_y_continuous(breaks = seq(0, 1, by = 0.5)) +
    geom_pointrange(
      stat = "summary", fatten = 0.8,
      fun.min = function(z) {
        quantile(z, 0.25)
      },
      fun.max = function(z) {
        quantile(z, 0.75)
      },
      fun = median
    ) +
    # geom_boxplot(width = 0.2, outliers = F, alpha = 0.2) +
    # geom_boxplot(aes(cell_type, score, fill = cell_type),
    #              width = 0.1, color = "white", outlier.shape = NA) +
    guides(
      x = guide_axis(angle = 45, cap = "both"),
      y = guide_axis(cap = "both")
    ) +
    scale_fill_manual(values = colors$cell_type) +
    labs(x = "Cell Type", y = "Specificity", title = l) +
    ggpubr::stat_compare_means(comparisons = list(c("Tumor Cell", "Epithelial")), tip.length = 0.05, label.y = 1.05, method = "wilcox", size = 2, vjust = -0.5) +
    coord_cartesian(clip = "off")
})
names(spec_pls) <- names(db$spec)

pcomb <- wrap_plots(spec_pls, ncol = 2) &
  theme(axis.title.y = element_text(hjust = 0.4))
ggsave(pcomb, file = here("plots", "all_specificity_plots.pdf"), width = 6, height = 6)
pcomb
```

```{r, fig.width = 2.85, fig.height = 3}
p1 <- spec_pls$cell_type + labs(title = NULL, y = "Patient\nspecificity\n(uncorrected)")

p2 <- spec_pls$timepoint_batch + labs(title = NULL, y = "Timepoint\nspecificity\n(corrected)")

pcomb <- p1 + p2 + plot_layout(ncol = 1, axes = "collect") & labs(x = NULL) & guides(x = guide_axis(angle = 45, cap = "both"))

pcomb
ggsave(pcomb, file = here("plots", "sample_timepoint_specificity_v2.pdf"), width = 2.85, height = 3)
```

```{r, fig.width = 3, fig.height = 2}
p1 <- spec_pls$cell_type_batch + labs(title = NULL, y = "Patient\nspecificity\n(corrected)")


pcomb <- p1 + labs(x = NULL) + guides(x = guide_axis(angle = 45, cap = "both"))

pcomb
ggsave(pcomb, file = here("plots", "specificity_patient_corrected_supp.pdf"), width = 3, height = 2)
```


### Density based tumor cells

#### Over cohort object

```{r, fig.width=7, fig.height=3}
library(ggtext)
df_raw <- FetchData(srt, vars = c("time_point", "is_tumor_cell", "cell_type", "umap_1", "umap_2")) %>%
  dplyr::rename("UMAP1" = "umap_1", "UMAP2" = "umap_2")

df <- df_raw %>%
  filter(is_tumor_cell)

devtools::load_all()
# Group by timepoint and calculate scaled density per group
df <- df %>%
  group_by(time_point, cell_type) %>%
  mutate(density = get_density(UMAP1, UMAP2, n = 500)) %>%
  mutate(density_scaled = standardize(density))



df$time_point_col <- factor(df$time_point,
  levels = levels(df$time_point),
  labels = glue("<b style='color:{colors$time_point[levels(df$time_point)]}'>{levels(df$time_point)}</b>")
)


# lab_cents <- get_label_centers(srt, group_var = "RNA_cluster_labels_coarse", reduced_dim = "RNA_umap_rpca_integrated")

p <- df %>%
  ungroup() %>%
  arrange((density_scaled)) %>%
  ggplot(aes(x = UMAP1, y = UMAP2)) +
  # scattermore::geom_scattermore(data = df_raw, color = "grey50", alpha = 0.5) +
  ggrastr::rasterize(geom_point(aes(color = density_scaled), size = 0.1), dpi = 600) +
  facet_wrap(~time_point_col) +
  scale_color_gradientn(colours = viridis::turbo(10)) +
  guides(color = guide_colorbar(keywidth = 0.5, keyheight = 3)) +
  # geom_label_repel(data = get_label_centers(srt, group_var = "RNA_cluster_labels_coarse", reduced_dim = "RNA_umap_rpca_integrated"), aes(x = x, y = y, label = RNA_cluster_labels_coarse), size = 3, label.padding = unit(0.1, "lines")) +
  # NoAxes() +
  # coord_cartesian(clip = "off") +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_markdown(size = 16),
    aspect.ratio = 1, panel.background = element_rect(fill = NA, color = "black"),
    plot.title = element_text(hjust = 0.5, size = 18),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  ) +
  labs(x = "UMAP1", y = "UMAP2", title = NULL, color = "Tumor Cell\nDensity")
ggsave(p, file = here("plots", "tumor_cell_density.pdf"), width = 7, height = 3)
p
```

#### Vertical version

```{r, fig.width=3, fig.height=7}
p <- df %>%
  ungroup() %>%
  arrange((density_scaled)) %>%
  ggplot(aes(x = UMAP1, y = UMAP2)) +
  ggrastr::rasterize(geom_point(aes(color = density_scaled), size = 0.1), dpi = 600) +
  facet_wrap(~time_point_col, ncol = 1) +
  scale_color_gradientn(colours = viridis::turbo(10), name = "Density") +
  guides(color = guide_colorbar(keywidth = 0.5, keyheight = 3)) +
  # geom_label_repel(data = get_label_centers(srt, group_var = "RNA_cluster_labels_coarse", reduced_dim = "RNA_umap_rpca_integrated"), aes(x = x, y = y, label = RNA_cluster_labels_coarse), size = 3, label.padding = unit(0.1, "lines")) +
  # NoAxes() +
  # coord_cartesian(clip = "off") +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_markdown(size = 16),
    aspect.ratio = 1, panel.background = element_rect(fill = NA, color = "black"),
    plot.title = element_text(hjust = 0.5, size = 18),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  ) +
  umap_theme() +
  NoAxes() +
  labs(x = "UMAP1", y = "UMAP2")
ggsave(p, file = here("plots", "tumor_cell_density_vertical.pdf"), width = 3, height = 7)
p
```

#### Over epithelial object

```{r, fig.width=7, fig.height=3}
library(ggtext)
df <- FetchData(srt_epi, vars = c("time_point", "is_tumor_cell", "cell_type", "umap_1", "umap_2", "TP53mut", "RB1mut")) %>%
  dplyr::rename("UMAP1" = "umap_1", "UMAP2" = "umap_2") %>%
  filter(is_tumor_cell)


# Group by timepoint and calculate scaled density per group
df <- df %>%
  group_by(time_point, cell_type) %>%
  mutate(density = get_density(UMAP1, UMAP2, n = 500)) %>%
  mutate(density_scaled = standardize(density))



df$time_point_col <- factor(df$time_point,
  levels = levels(df$time_point),
  labels = glue("<b style='color:{colors$time_point[levels(df$time_point)]}'>{levels(df$time_point)}</b>")
)


# lab_cents <- get_label_centers(srt, group_var = "RNA_cluster_labels_coarse", reduced_dim = "RNA_umap_rpca_integrated")

p <- df %>%
  ungroup() %>%
  arrange((density_scaled)) %>%
  ggplot(aes(x = UMAP1, y = UMAP2)) +
  ggrastr::rasterize(geom_point(aes(color = density_scaled), size = 0.1), dpi = 600) +
  facet_wrap(~time_point_col) +
  scale_color_gradientn(colours = viridis::turbo(10), name = "Density") +
  guides(color = guide_colorbar(keywidth = 0.5, keyheight = 3)) +
  # geom_label_repel(data = get_label_centers(srt, group_var = "RNA_cluster_labels_coarse", reduced_dim = "RNA_umap_rpca_integrated"), aes(x = x, y = y, label = RNA_cluster_labels_coarse), size = 3, label.padding = unit(0.1, "lines")) +
  # NoAxes() +
  # coord_cartesian(clip = "off") +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_markdown(size = 16),
    aspect.ratio = 1, panel.background = element_rect(fill = NA, color = "black"),
    plot.title = element_text(hjust = 0.5, size = 18),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  ) +
  labs(x = "UMAP1", y = "UMAP2", title = "Tumor Cell Density")
ggsave(p, file = here("plots", "tumor_cell_density_epi.pdf"), width = 7, height = 3)
p
```

##### TP53mut

```{r}
library(ggtext)
df <- FetchData(srt_epi, vars = c("time_point", "is_tumor_cell", "cell_type", "umap_1", "umap_2", "TP53mut", "RB1mut", "has_impact")) %>%
  dplyr::rename("UMAP1" = "umap_1", "UMAP2" = "umap_2") %>%
  filter(is_tumor_cell) %>%
  mutate(TP53mut = case_when(
    has_impact == F ~ "NA",
    TP53mut ~ "TP53mut",
    !TP53mut ~ "TP53wt",
    .default = "Other"
  ))
# Group by timepoint and calculate scaled density per group
df <- df %>%
  group_by(time_point, cell_type, TP53mut) %>%
  mutate(density = get_density(UMAP1, UMAP2, n = 500)) %>%
  mutate(density_scaled = standardize(density))



df$time_point_col <- factor(df$time_point,
  levels = levels(df$time_point),
  labels = glue("<b style='color:{colors$time_point[levels(df$time_point)]}'>{levels(df$time_point)}</b>")
)


# lab_cents <- get_label_centers(srt, group_var = "RNA_cluster_labels_coarse", reduced_dim = "RNA_umap_rpca_integrated")

p <- df %>%
  filter(TP53mut != "NA") %>%
  ungroup() %>%
  # mutate(TP53mut = ifelse(TP53mut, "TP53mut", "TP53wt")) %>%
  arrange((density_scaled)) %>%
  ggplot(aes(x = UMAP1, y = UMAP2)) +
  ggrastr::rasterise(geom_point(aes(color = density_scaled), size = 0.1), dpi = 600) +
  facet_grid(TP53mut ~ time_point_col) +
  scale_color_gradientn(colours = viridis::turbo(10), name = "Density") +
  guides(color = guide_colorbar(keywidth = 0.5, keyheight = 3)) +
  # geom_label_repel(data = get_label_centers(srt, group_var = "RNA_cluster_labels_coarse", reduced_dim = "RNA_umap_rpca_integrated"), aes(x = x, y = y, label = RNA_cluster_labels_coarse), size = 3, label.padding = unit(0.1, "lines")) +
  # NoAxes() +
  # coord_cartesian(clip = "off") +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_markdown(size = 16),
    aspect.ratio = 1, panel.background = element_rect(fill = NA, color = "black"),
    plot.title = element_text(hjust = 0.5, size = 18),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  ) +
  labs(x = "UMAP1", y = "UMAP2", title = "Tumor Cell Density")
ggsave(p, file = here("plots", "tumor_cell_density_TP53mut.pdf"), width = 7, height = 5)
p
```


```{r, fig.width = 5, fig.height = 3}
# TN only
p <- df %>%
  filter(TP53mut != "NA", time_point == "TN") %>%
  ungroup() %>%
  # mutate(TP53mut = ifelse(TP53mut, "TP53mut", "TP53wt")) %>%
  arrange((density_scaled)) %>%
  ggplot(aes(x = UMAP1, y = UMAP2)) +
  ggrastr::rasterise(geom_point(aes(color = density_scaled), size = 0.1), dpi = 600) +
  facet_grid(time_point_col ~ fct_rev(TP53mut)) +
  scale_color_gradientn(colours = viridis::turbo(10), name = "Density") +
  guides(color = guide_colorbar(keywidth = 0.5, keyheight = 3)) +
  # geom_label_repel(data = get_label_centers(srt, group_var = "RNA_cluster_labels_coarse", reduced_dim = "RNA_umap_rpca_integrated"), aes(x = x, y = y, label = RNA_cluster_labels_coarse), size = 3, label.padding = unit(0.1, "lines")) +
  # NoAxes() +
  # coord_cartesian(clip = "off") +
  theme(
    strip.background = element_blank(),
    strip.text.y = element_markdown(size = 16),
    aspect.ratio = 1, panel.background = element_rect(fill = NA, color = "black"),
    plot.title = element_text(hjust = 0.5, size = 18),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  ) +
  labs(x = "UMAP1", y = "UMAP2")
ggsave(p, file = here("plots", "TP53mut_tumor_cell_density_TNonly.pdf"), width = 5, height = 3)
p
```

##### Secondary mapk in PD

```{r}
library(ggtext)
second_mapk <- read.table(file = here("tables", "secondary_mapk.txt"), header = T, sep = "\t")

srt_epi$secondary_mapk <- second_mapk$secondary_mapk[match(srt_epi$sample_id_new, second_mapk$sample_id)]
srt_epi$MAPKalt <- ifelse(!is.na(srt_epi$secondary_mapk), "MAPKalt", "Unknown")

df <- FetchData(srt_epi, vars = c("time_point", "is_tumor_cell", "cell_type", "umap_1", "umap_2", "TP53mut", "RB1mut", "secondary_mapk", "MAPKalt")) %>%
  dplyr::rename("UMAP1" = "umap_1", "UMAP2" = "umap_2") %>%
  filter(is_tumor_cell) %>%
  mutate(secondary_mapk_tf = !is.na(secondary_mapk))

# Group by timepoint and calculate scaled density per group
df <- df %>%
  group_by(time_point, cell_type, MAPKalt) %>%
  mutate(density = get_density(UMAP1, UMAP2, n = 500)) %>%
  mutate(density_scaled = standardize(density))



df$time_point_col <- factor(df$time_point,
  levels = levels(df$time_point),
  labels = glue("<b style='color:{colors$time_point[levels(df$time_point)]}'>{levels(df$time_point)}</b>")
)


# lab_cents <- get_label_centers(srt, group_var = "RNA_cluster_labels_coarse", reduced_dim = "RNA_umap_rpca_integrated")

p <- df %>%
  ungroup() %>%
  arrange((density_scaled)) %>%
  ggplot(aes(x = UMAP1, y = UMAP2)) +
  ggrastr::rasterise(geom_point(aes(color = density_scaled), size = 0.1), dpi = 600) +
  facet_grid(MAPKalt ~ time_point_col) +
  scale_color_gradientn(colours = viridis::turbo(10), name = "Density") +
  guides(color = guide_colorbar(keywidth = 0.5, keyheight = 3)) +
  # geom_label_repel(data = get_label_centers(srt, group_var = "RNA_cluster_labels_coarse", reduced_dim = "RNA_umap_rpca_integrated"), aes(x = x, y = y, label = RNA_cluster_labels_coarse), size = 3, label.padding = unit(0.1, "lines")) +
  # NoAxes() +
  # coord_cartesian(clip = "off") +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_markdown(size = 16),
    aspect.ratio = 1, panel.background = element_rect(fill = NA, color = "black"),
    plot.title = element_text(hjust = 0.5, size = 18),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank()
  ) +
  labs(x = "UMAP1", y = "UMAP2", title = "Tumor Cell Density")
ggsave(p, file = here("plots", "tumor_cell_density_MAPKalt.pdf"), width = 7, height = 5)
p
```


### Mean Proportion over major cell types

```{r, fig.width = 4.5, fig.height = 2.5}
p <- srt@meta.data %>%
  add_count(sample_id_new, name = "sample_n") %>%
  add_count(sample_id_new, cell_type, name = "cell_type_n") %>%
  mutate(prop = cell_type_n / sample_n) %>%
  distinct(sample_id_new, cell_type, time_point, sample_n, cell_type_n, prop) %>%
  filter(cell_type != "Other") %>%
  ggplot(aes(x = cell_type, y = prop)) +
  geom_boxplot(aes(fill = time_point), outlier.size = 0.25, fatten = 1.5) +
  scale_fill_manual(values = colors$time_point) +
  labs(x = "Cell Type", y = "Proportion", fill = "Timepoint") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.2)) +
  guides(
    x = guide_axis(cap = "both", angle = 45),
    y = guide_axis(cap = "both")
  ) +
  ggpubr::geom_pwc(tip.length = 0, symnum.args = symnum_0.1_star, label = "p.adj.signif", aes(group = time_point), hide.ns = T, vjust = 0.5, method = "wilcox_test", step.increase = 0.05) +
  coord_cartesian(clip = "off") #+
# theme(legend.position = "top", plot.margin = margin(0, 1, 1, 1, "mm"), legend.title.position = "top", legend.title = element_text(hjust = 0.5), legend.box.margin = margin(), legend.box.spacing = unit(0, "mm"), legend.box.background = element_blank(), legend.background = element_blank())
p
ggsave(p, file = here("plots", "cohort_timepoint_cell_type_prop.pdf"), width = 4.5, height = 2.5)
```

### Counts per sample barplot

```{r, fig.width=10, fig.height = 4}
p1 <- srt@meta.data %>%
  ggplot(aes(x = sample_id_new)) +
  geom_bar(aes(fill = cell_type)) +
  facet_grid(. ~ time_point, scales = "free", space = "free") +
  guides(x = guide_axis(angle = 90)) +
  scale_fill_manual(values = colors$cell_type) +
  labs(fill = "Cell Type", y = "Count", x = "Sample") +
  guides(fill = guide_legend(keywidth = 0.4, keyheight = 0.2)) +
  theme_clean()

p2 <- srt@meta.data %>%
  ggplot(aes(x = sample_id_new)) +
  geom_bar(aes(fill = cell_type), position = "fill") +
  facet_grid(. ~ time_point, scales = "free", space = "free") +
  guides(
    x = guide_axis(angle = 90),
    y = guide_axis(cap = "none")
  ) +
  scale_fill_manual(values = colors$cell_type) +
  labs(fill = "Cell Type", y = "Proportion", x = "Sample") +
  guides(fill = guide_legend(keywidth = 0.4, keyheight = 0.2)) +
  theme_clean()

# Final formatting
p1 <- p1 + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) + labs(x = NULL)
p2 <- p2 + theme(strip.text.x = element_blank())

pcomb <- p1 + p2 + plot_layout(guides = "collect", axes = "collect", ncol = 1) & theme(plot.background = element_blank()) & scale_y_continuous(expand = c(0, 0))
pcomb
ggsave(pcomb, file = here("plots", "sample_cellcounts_barplot.pdf"), width = 10, height = 4)
```


### Tumor cellularity

```{r, fig.width = 9, fig.height = 2.5}
df <- srt@meta.data %>%
  add_count(sample_id_new, cell_type, name = "n_ct") %>%
  add_count(sample_id_new, name = "n_samp") %>%
  mutate(ct_prop = n_ct / n_samp) %>%
  distinct(sample_id_new, cell_type, ct_prop, n_ct, n_samp, procedure_type) %>%
  pivot_wider(id_cols = sample_id_new, names_from = cell_type, values_from = ct_prop, values_fill = 0) %>%
  pivot_longer(cols = !sample_id_new, names_to = "cell_type", values_to = "cell_type_prop") %>%
  left_join(db$metadata)

# How many samples w/no tumor
tum_prop <- df %>%
  filter(cell_type == "Tumor Cell") %>%
  pull(cell_type_prop)
# sum(tum_prop == 0)
# max(tum_prop)
p <- df %>%
  filter(cell_type == "Tumor Cell") %>%
  ggplot(aes(x = fct_reorder(sample_id_new, -cell_type_prop, mean), y = cell_type_prop)) +
  geom_col(aes(fill = time_point), show.legend = F) +
  scale_fill_manual(values = colors$time_point) +
  facet_grid(. ~ time_point, scales = "free", space = "free") +
  theme_clean() +
  theme(strip.background = element_blank()) +
  coord_cartesian(ylim = c(0, 1), clip = "off") +
  scale_y_continuous(expand = c(0, 0)) +
  guides(
    x = guide_axis(angle = 90),
    y = guide_axis(cap = "none")
  ) +
  labs(x = "Sample", y = "Tumor Cell Proportion")

p2 <- df %>%
  filter(cell_type == "Tumor Cell") %>%
  ggplot(aes(x = time_point, y = cell_type_prop)) +
  geom_boxplot(aes(fill = time_point), show.legend = F) +
  scale_fill_manual(values = colors$time_point) +
  theme_clean() +
  scale_y_continuous(expand = c(0, 0), breaks = seq(0, 1, by = 0.25)) +
  guides(
    x = guide_axis(angle = 45, cap = "none"),
    y = guide_axis(cap = "both")
  ) +
  labs(x = "Timepoint", y = "Tumor Cell Proportion") +
  coord_cartesian(clip = "off", ylim = c(0, 1)) +
  ggpubr::geom_pwc(tip.length = 0, label = "p.adj.signif", symnum.args = symnum_0.1_star, hide.ns = T, vjust = 0.2)
pcomb <- p + free(p2, type = "label", side = "b") + plot_layout(ncol = 2, widths = c(1, 0.1), axes = "collect") &
  theme(plot.background = element_blank())
pcomb
ggsave(pcomb, file = here("plots", "tumor_cell_proportion_barplot.pdf"), width = 9, height = 2.5)
```

Cellularity by tumor site

```{r, fig.width = 7, fig.height = 3}
p1 <- df %>%
  filter(cell_type == "Tumor Cell") %>%
  ggplot(aes(x = fct_relevel(site_of_tissue_simple, names(colors$site_of_tissue_simple)), y = cell_type_prop)) +
  geom_boxplot(aes(fill = site_of_tissue_simple), show.legend = F) +
  geom_quasirandom(alpha = 0.5) +
  scale_fill_manual(values = colors$site_of_tissue_simple) +
  theme_clean() +
  scale_y_continuous(expand = c(0, 0), breaks = seq(0, 1, by = 0.25)) +
  guides(
    x = guide_axis(angle = 0, cap = "none"),
    y = guide_axis(cap = "both")
  ) +
  labs(x = "Tissue Site", y = "Tumor Cell Proportion") +
  coord_cartesian(clip = "off", ylim = c(0, 1))

p2 <- df %>%
  filter(cell_type == "Tumor Cell") %>%
  ggplot(aes(x = fct_relevel(procedure_type, "Biopsy", "Fresh Frozen", "Resection"), y = cell_type_prop)) +
  geom_boxplot(aes(fill = procedure_type), show.legend = F) +
  geom_quasirandom(alpha = 0.5) +
  scale_fill_manual(values = dittoColors()) +
  theme_clean() +
  scale_y_continuous(expand = c(0, 0), breaks = seq(0, 1, by = 0.25)) +
  guides(
    x = guide_axis(angle = 0, cap = "none"),
    y = guide_axis(cap = "both")
  ) +
  labs(x = "Procedure Type", y = "Tumor Cell Proportion") +
  coord_cartesian(clip = "off", ylim = c(0, 1))

pcomb <- p1 + p2 + plot_layout(ncol = 2) &
  theme_classic() &
  ggpubr::stat_compare_means(label.x.npc = 0.1) &
  guides(x = guide_axis(angle = 45))
pcomb
ggsave(pcomb, file = here("plots", "tumor_cellularity_site_procedure.pdf"), width = 7, height = 3)
```


### InferCNV plots

```{r}
sce <- db$infercnv
sce$sample_id_new <- srt$sample_id_new[match(colnames(sce), srt$cell_id)]
sce$cell_type <- srt$cell_type[match(colnames(sce), srt$cell_id)]
sce$time_point <- srt$time_point[match(colnames(sce), srt$cell_id)]

sce <- sce[, sce$sample_id_new %in% unique(srt$sample_id_new) & sce$cell_type %in% c("Tumor Cell", "Epithelial")]

library(ComplexHeatmap)
# Sample 1000 tumor and 1000 normal cells from each patient


mat <- assay(sce) %>%
  t() %>%
  log2()

df_anno <- colData(sce) %>%
  as.data.frame() %>%
  select(cell_type, time_point, sample_id_new) %>%
  arrange(time_point, sample_id_new, cell_type)
# rownames_to_column("cell_id") %>%
# group_by(cell_type, sample_id_new) %>%
# slice_sample(n = 100) %>%
# as.data.frame() %>%
# column_to_rownames("cell_id")

df_anno$cell_type <- factor(df_anno$cell_type)

mat <- mat[rownames(df_anno), ]

# ct_cols <- dittoColors()[1:length(unique(df_anno$cell_type))]
# names(ct_cols) <- sort(unique(df_anno$cell_type))

anno_cols <- list(
  "time_point" = colors$time_point,
  "cell_type" = colors$cell_type[unique(df_anno$cell_type)],
  "sample_id_new" = colors$sample_id_new
)

left_anno <- HeatmapAnnotation(
  df = df_anno,
  annotation_label = c("Cell Type", "Timepoint", "Sample"),
  which = "row", show_legend = c(T, T, T), col = anno_cols,
  annotation_legend_param = list(
    "sample_id_new" = list(
      ncol = 3
    )
  )
)

# Split columns by chromosome
chrs <- as.vector(gsub("chr", "", GenomeInfoDb::seqnames(SummarizedExperiment::rowRanges(sce))))
col_split <- factor(chrs, levels = unique(gtools::mixedsort(chrs)))

ht <- Heatmap(
  matrix = mat,
  name = "Log2Ratio",
  left_annotation = left_anno,
  col = logr_col_fun(breaks = c(-0.05, -0.01, 0.01, 0.05), colors = c("blue", "white", "white", "red")),
  cluster_rows = F,
  cluster_columns = F,
  show_column_dend = F,
  show_row_dend = F,
  show_row_names = F,
  show_column_names = F,
  column_split = col_split,
  row_split = df_anno$cell_type,
  raster_by_magick = F, raster_device = "CairoPNG",
  border = T,
  raster_quality = 25
)

pdf(here("plots/cohort_cnv_heatmap.pdf"), width = 13, height = 8)
draw(ht)
dev.off()
```

## Figure 3 -- Epithelial Cells


```{r}
meta_programs_df <- read.table(file = "https://raw.githubusercontent.com/mjz1/meta_programs_tirosh/refs/heads/main/tirosh_mp_patched.txt", header = T, sep = "\t")

meta_programs <- split(meta_programs_df, meta_programs_df$cell_type)

meta_programs <- map(meta_programs, .f = function(x) {
  res <- split(x, x$meta_program)
  map(res, .f = function(y) y$Gene)
})

names(meta_programs) <- janitor::make_clean_names(names(meta_programs))
```


### Explore

#### Unknown PDC state

Mainly in four patients

```{r}
table(srt_epi@meta.data[srt_epi$cell_type_epi == "PD9/PD27 unk.", ]$sample_id_new) %>% sort(decreasing = T)
```

```{r}
epi_marker_f <- here("results", "epi_type_degs.txt")
if (FALSE) {
  epi_degs <- FindAllMarkers(srt_epi, group.by = "cell_type_epi", min.pct = 0.05, logfc.threshold = 0.25, only.pos = T, max.cells.per.ident = 1000)
  write.table(epi_degs, file = epi_marker_f, col.names = T, row.names = F, sep = "\t")
}
epi_degs <- read.table(epi_marker_f, header = T, sep = "\t")
```



```{r, fig.width = 18, fig.height = 4}
epi_degs_top <- epi_degs %>%
  group_by(cluster) %>%
  slice_head(n = 10)

p <- DotPlotter(srt_epi, features = split(epi_degs_top$gene, epi_degs_top$cluster), group.by = "cell_type_epi", rotate_y_strip = T)
p
```


### General Exploratory

```{r}
DimPlot_scCustom(srt_epi, group.by = "cell_type_epi", label = T, colors_use = colors$cell_type_epi, label.size = 3)
```


```{r, fig.width = 8, fig.height = 5}
p <- FeaturePlot_scCustom(srt_epi, features = c("SFTPB", "KRT5", "KRT17", "DSG3", "KRT6A", "SCGB3A2"), colors_use = viridis::rocket(20), order = T, raster = F, raster.dpi = c(256, 256), num_columns = 3, max.cutoff = "q99", pt.size = 0.01) &
  guides(color = guide_colorbar(keywidth = 0.5, keyheight = 3)) &
  umap_theme() &
  NoAxes() &
  # NoLegend() &
  theme(aspect.ratio = 1, plot.title = element_text(hjust = 0.5, face = "bold"))
p
```


```{r}
p <- VlnPlot_scCustom(srt_epi, features = c("SFTPB", "KRT5", "KRT17", "DSG3", "KRT6A", "SCGB3A2"), group.by = "cell_type_epi", pt.size = 0, plot_boxplot = T)
p
```


```{r}
x <- Plot_Density_Custom(seurat_object = srt_epi, features = c("TRAVAGLINI_LUNG_ALVEOLAR_EPITHELIAL_TYPE_1_CELL", "TRAVAGLINI_LUNG_ALVEOLAR_EPITHELIAL_TYPE_2_CELL", "TRAVAGLINI_LUNG_BASAL_CELL", "TRAVAGLINI_LUNG_NEUROENDOCRINE_CELL", "TRAVAGLINI_LUNG_PROLIFERATING_BASAL_CELL"), viridis_palette = "cividis", joint = F, combine = F)
x
```



```{r, fig.width = 10, fig.height = 8}
Plot_Density_Custom(seurat_object = srt_epi, features = c("SCGB3A1", "SCGB3A2"), joint = T, viridis_palette = "cividis")


x <- Plot_Density_Custom(seurat_object = srt_epi, features = c("KRT6A", "KRT13", "KRT14", "DSG3"), joint = T, viridis_palette = "cividis")


# AT0
# alveolar type 0 cells (n=1,440, expressing STFPB+, SCGB3A2+, SFTPChigh and SCGB3A1low) and the closely related preterminal bronchiole secretory cells (n=4,393, expressing SFTPB+, SCGB3A2+, SFTPClow and SCGB3A1high, together with alveolar type 0 cells called transitional club-AT2 cells)33,34
x <- Plot_Density_Custom(seurat_object = srt_epi, features = c("SFTPB", "SCGB3A2", "SFTPC"), joint = T, viridis_palette = "cividis")

x <- Plot_Density_Custom(seurat_object = srt_epi, features = c("SFTPB", "SCGB3A2", "SCGB3A1"), joint = T, viridis_palette = "cividis")
```


```{r, fig.width = 10, fig.height = 8}
# Hillock basal
x <- Plot_Density_Custom(seurat_object = srt_epi, features = c("KRT6A", "KRT13", "KRT14", "DSG3", "TP63"), viridis_palette = "cividis", joint = T)
x
```


```{r, fig.width = 10, fig.height = 8}
# Hillock luminal
x <- Plot_Density_Custom(seurat_object = srt_epi, features = c("KRT6A", "KRT13", "KRT14", "DSG3", "SERPINB2"), viridis_palette = "cividis", joint = T)
x
```


### Figure Plots

#### UMAP

```{r, fig.width = 6.75, fig.height = 4.25}
lab_cents <- scatools:::get_label_centers(obj = srt_epi, group_var = "cell_type_epi", reduced_dim = "umap")

df_leg <- srt_epi@meta.data %>%
  add_count(cell_type_epi, name = "epi_n") %>%
  distinct(cell_type_epi, epi_n) %>%
  mutate(label = glue("({prettyNum(epi_n, big.mark = ',')})"))

p1_leg <- df_leg %>%
  ggplot(aes(y = fct_rev(cell_type_epi), x = epi_n)) +
  geom_point(aes(x = -10000, color = cell_type_epi), size = 3.5) +
  geom_col(aes(fill = cell_type_epi), width = 0.5) +
  geom_text(aes(x = epi_n, label = label), size = 3, hjust = -0.1, color = "black") +
  scale_fill_manual(values = colors$cell_type_epi) +
  scale_color_manual(values = colors$cell_type_epi) +
  coord_cartesian(clip = "off") +
  # facet_grid(fct_relevel(cell_type_l2, "Myeloid", "TNK", "B/Plasma", "Stroma", "Other") ~ ., scales = "free", space = "free", switch = "y") +
  theme(
    strip.placement = "outside",
    strip.background.y = element_part_rect(side = "r", fill = NA),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    axis.text.x = element_blank(),
    strip.text.y.left = element_text(angle = 0, hjust = 1),
    plot.margin = margin(1, 4, 1, 1, unit = "lines")
  ) +
  labs(x = NULL, y = NULL) +
  scale_x_continuous(expand = c(0.025, 0)) +
  guides(x = guide_axis(cap = "both")) +
  NoLegend()

p1 <- DimPlot(srt_epi, reduction = "umap", group.by = "cell_type_epi", shuffle = T, raster.dpi = c(350, 350)) +
  umap_theme() +
  ggrepel::geom_text_repel(
    data = lab_cents, aes(x = x, y = y, label = cell_type_epi),
    min.segment.length = 0,
    segment.size = 0.75,
    fontface = "bold",
    force = 5,
    bg.colour = "white",
    max.iter = 1e9,
    bg.r = 0.1,
    position = ggpp::position_nudge_center(center_x = 2, center_y = 2, direction = "radial", x = 0, y = 0)
  ) +
  # scale_color_manual(values = colors$site_of_tissue_simple, breaks = names(colors$site_of_tissue_simple), labels = label_num_generator(srt$site_of_tissue_simple)) +
  #   ggrepel::geom_text_repel(color = "black", data = lab_cents, aes(x = x, y = y, label = cell_type_l5), size = 3, fontface = "bold", bg.colour = "white", bg.r = .1, force = 0.1, max.iter = 1e9, min.segment.length = Inf, max.time = 1, max.overlaps = Inf) +

  scale_color_manual(values = colors$cell_type_epi, labels = label_num_generator(srt_epi$cell_type_epi), breaks = levels(srt_epi$cell_type_epi)) +
  NoAxes() +
  NoLegend() +
  theme(aspect.ratio = 1) +
  labs(title = "Epithelial Cells")

layout <- c(
  area(t = 10, b = 100, l = 1, r = 100),
  area(t = 4, l = 88, b = 59, r = 100)
)
# pcomb <- (p1 + theme(plot.margin = margin())) + free(p1_leg + theme(aspect.ratio = 1.9, plot.margin = margin())) + plot_layout(design = layout)
pcomb <- (p1 + theme(plot.margin = margin())) + p1_leg + plot_layout(design = layout)


ggsave(pcomb, file = here("plots", "epi_umap.pdf"), width = 6.75, height = 4.25)
pcomb
```


```{r}
p2 <- DimPlot(srt_epi, reduction = "umap", group.by = "cell_type_epi_coarse", shuffle = T) +
  umap_theme() +
  # scale_color_manual(values = colors$site_of_tissue_simple, breaks = names(colors$site_of_tissue_simple), labels = label_num_generator(srt$site_of_tissue_simple)) +
  scale_color_manual(values = colors$cell_type_epi, labels = label_num_generator(srt_epi$cell_type_epi_coarse)) +

  NoAxes() +
  theme(aspect.ratio = 1) +
  labs(title = "Epithelial Cells")
p2
```


```{r}
ps <- FeaturePlot(srt_epi, reduction = "umap", features = c(markers$epi_markers_final$AT2), order = T, combine = T, max.cutoff = "q99") & scale_color_viridis_c(option = "cividis") & guides(color = guide_colorbar(keywidth = 0.4, keyheight = 3)) & NoAxes()
ps
```


#### Marker dotplots

##### Custom Genes


```{r}
epi_dotplot <- DotPlotter(srt_epi, features = markers$epi_markers_final, group.by = "cell_type_epi", cluster_groups = F, rotate_y_strip = T)
epi_dotplot +
  theme(axis.text.y = element_markdown())
ggsave(epi_dotplot, file = here("plots", "epi_marker_dotplot_unfilt.pdf"), width = 9, height = 4)
```

Tumor cells only

```{r}
epi_dotplot <- DotPlotter(srt_epi[, srt_epi$is_tumor_cell], features = markers$epi_markers_final, group.by = "cell_type_epi", cluster_groups = F, rotate_y_strip = T)
epi_dotplot +
  theme(axis.text.y = element_markdown())
ggsave(epi_dotplot, file = here("plots", "epi_marker_dotplot_tumor_only.pdf"), width = 9, height = 4)
```

##### HLCA marker set

```{r, fig.width = 14, fig.height = 4.5}
dp <- DotPlotter(srt_epi, features = db$markers$epi_markers_hlca, group.by = "cell_type_epi", cluster_groups = F)
dp_f <- dp + theme(strip.text.x = element_text(angle = 90, hjust = 0), panel.spacing = unit(1, "mm")) +
  guides(y.sec = guide_axis()) +
  labs(title = "Human Lung Cell Atlas Epithelial Marker List") +
  theme(plot.title = element_text(hjust = 0.5))
dp_f
ggsave(dp_f, file = here("plots", "hlca_marker_dotplot.pdf"), width = 14, height = 4.5)
```


#### Marker density plots

```{r, fig.width = 6, fig.height = 4}
glist <- c("SFTPB", "SFTPC", "SCGB3A2", "KRT13", "KRT6A", "DSG3")

x <- Plot_Density_Custom(seurat_object = srt_epi, features = glist, viridis_palette = "inferno", joint = F, combine = F) # %>%


names(x) <- glist

res <- map(names(x), .f = function(gene) {
  x[[gene]]$data %>%
    mutate(!!gene := standardize(feature))
})
names(res) <- names(x)

pls <- map(names(res), .f = function(gene) {
  p <- res[[gene]] %>%
    arrange(.data[[gene]]) %>%
    ggplot(aes(x = umap_1, y = umap_2)) +
    scattermore::geom_scattermore(aes(color = .data[[gene]]), pixels = c(256, 256)) +
    # geom_point(aes(color = feature), size = 0.5) +
    scale_color_viridis_c(option = "inferno") +
    labs(title = gene, color = "Density")
  # p <- ggrastr::rasterise(p, dpi = 300)
})

pcomb <- pls %>%
  wrap_plots(nrow = 2, guides = "collect") &
  guides(color = guide_colorbar(keywidth = 0.5, keyheight = 3)) &
  umap_theme() &
  NoAxes() &
  # NoLegend() &
  theme(aspect.ratio = 1, plot.title = element_text(hjust = 0.5, face = "bold")) &
  labs(color = "Expression\nDensity")

pcomb
ggsave(pcomb, filename = here("plots", "epi_exp_dens.pdf"), width = 6, height = 4)
```


#### SFTPB Expression

```{r, fig.width = 2.5, fig.height = 3}
# p_all <- VlnPlot_scCustom(srt_epi, colors_use = colors$cell_type_epi, features = "SFTPB", group.by = "cell_type_epi", plot_boxplot = T, sort = T, add.noise = F)
# p_all
df <- FetchData(srt_epi, vars = c("cell_type_epi", "SFTPB"))
# df <- VlnPlot_scCustom(srt_epi[,grepl("^PDC|^AT2", srt_epi$cell_type_epi)], colors_use = colors$cell_type_epi, features = "SFTPB", group.by = "cell_type_epi", plot_boxplot = T, sort = T, add.noise = F)$data
p <- df %>%
  filter(grepl("^PDC|^AT2", cell_type_epi)) %>%
  ggplot(aes(x = fct_reorder(cell_type_epi, -SFTPB, median), y = SFTPB)) +
  geom_violin(scale = "width", aes(fill = cell_type_epi)) +
  guides(x = guide_axis(angle = 45)) +
  geom_boxplot(width = 0.1, fill = "white", outliers = F) +
  # geom_pointrange(
  #   stat = "summary", fatten = 2,
  #   fun.min = function(z) {
  #     quantile(z, 0.25)
  #   },
  #   fun.max = function(z) {
  #     quantile(z, 0.75)
  #   },
  #   fun = median,
  # ) +
  scale_fill_manual(values = colors$cell_type_epi) +
  theme_classic(base_size = 16) +
  NoLegend() +
  labs(x = "Celltype", y = "Expression", title = "SFTPB") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
p
ggsave(p, file = here("plots", "sftpc_epi_expression.pdf"), width = 2.5, height = 3)
```


#### Relative cluster proportions

```{r, fig.width = 6.25, fig.height = 7.5}
p1 <- srt_epi@meta.data %>%
  ggplot(aes(x = cell_type_epi)) +
  geom_bar(aes(fill = sample_id_new), position = "fill") +
  scale_fill_manual(values = colors$sample_id_new) +
  labs(fill = "Patient", y = "Proportion") +
  guides(fill = guide_legend(keywidth = 0.5, keyheight = 0.5, ncol = 6))

p2 <- srt_epi@meta.data %>%
  ggplot(aes(x = cell_type_epi)) +
  geom_bar(aes(fill = site_of_tissue_simple), position = "fill") +
  scale_fill_manual(values = colors$site_of_tissue_simple) +
  labs(fill = "Tissue site", y = "Proportion") +
  guides(fill = guide_legend(keywidth = 0.5, keyheight = 0.5, ncol = 2))

p3 <- srt_epi@meta.data %>%
  ggplot(aes(x = cell_type_epi)) +
  geom_bar(aes(fill = time_point), position = "fill") +
  scale_fill_manual(values = colors$time_point) +
  guides(x = guide_axis(angle = 45)) +
  labs(fill = "Timepoint", y = "Proportion") +
  guides(fill = guide_legend(keywidth = 0.5, keyheight = 0.5))


# p4 <- srt_epi@meta.data %>%
#   ggplot(aes(x = cell_type_epi)) +
#   geom_bar(aes(fill = fct_rev(cell_type)), position = "fill") +
#   scale_fill_manual(values = colors$cell_type) +
#   guides(x = guide_axis(angle = 45))

p4 <- srt_epi@meta.data %>%
  # filter(!grepl("PD9", cell_type_epi)) %>%
  add_count(cell_type_epi, cell_type, name = "n1") %>%
  add_count(cell_type_epi, name = "n2") %>%
  mutate(prop = n1 / n2) %>%
  distinct(cell_type, cell_type_epi, prop) %>%
  filter(cell_type == "Tumor Cell") %>%
  ggplot(aes(x = cell_type_epi, y = prop)) +
  geom_col(aes(fill = fct_rev(cell_type)), show.legend = F) +
  geom_hline(yintercept = seq(0.25, 1, by = 0.25), alpha = 0.6, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
  scale_fill_manual(values = colors$cell_type) +
  labs(y = "Proportion\n(tumor cells)")

p5 <- srt_epi@meta.data %>%
  ggplot(aes(x = cell_type_epi)) +
  geom_bar(aes(fill = Phase), position = "fill") +
  scale_fill_manual(values = colors$Phase) +
  guides(x = guide_axis(angle = 45)) +
  labs(fill = "Cell Cycle Phase", y = "Proportion") +
  guides(fill = guide_legend(keywidth = 0.5, keyheight = 0.5))

gap_size <- 2

pcomb <- wrap_plots(
  # free(p1, type = "space", side = "r"),
  p1 + theme(plot.margin = margin(t = 4, r = 2, b = gap_size, l = 2, "mm")),
  free(p3 + theme(plot.margin = margin(t = gap_size, r = 2, b = gap_size, l = 2, "mm")), type = "space", side = "r"),
  free(p2 + theme(plot.margin = margin(t = gap_size, r = 2, b = gap_size, l = 2, "mm")), type = "space", side = "r"),
  # p2,
  free(p5 + theme(plot.margin = margin(t = gap_size, r = 2, b = gap_size, l = 2, "mm")), type = "space", side = "r"),
  free(p4 + theme(plot.margin = margin(t = gap_size, r = 2, b = 2, l = 4, "mm")), type = "space", side = "l"),
  ncol = 1, axes = "collect_x"
) &
  guides(
    x = guide_axis(angle = 65, cap = "none"),
    y = guide_axis(cap = "both")
  ) &
  scale_y_continuous(expand = c(0, 0)) &
  labs(x = "Epithelial cell state") &
  coord_cartesian(clip = "off")
ggsave(pcomb, file = here("plots", "epi_type_feature_proportions.pdf"), width = 6.75, height = 7.5)
pcomb
```


#### Epithelial type proportions

This version is proportion of each pts tumor cells only

```{r, fig.width = 5, fig.height = 3}
# my_comparisons <- list(c("Pre", "On"), c("Pre", "PD"), c("On", "PD"))
# my_comparisons <- list(c("Pre", "PD"))


use_col <- "cell_type_epi"

df <- srt_epi@meta.data %>%
  # filter(!grepl("PD9", cell_type_epi)) %>%
  # mutate(RNA_cluster_labels = fct_rev(RNA_cluster_labels)) %>%
  filter(cell_type == "Tumor Cell") %>%
  add_count(.data[[use_col]], time_point, sample_id_new, name = "epi_type_n") %>%
  add_count(sample_id_new, name = "sample_n") %>%
  mutate(tum_prop = epi_type_n / sample_n) %>%
  distinct(.data[[use_col]], epi_type_n, sample_n, tum_prop, sample_id, time_point)


p_df <- df %>%
  group_by(.data[[use_col]]) %>%
  rstatix::wilcox_test(tum_prop ~ time_point) %>%
  rstatix::adjust_pvalue(method = "BH") %>%
  rstatix::add_significance(
    symbols = symnum_0.1_star$symbols,
    cutpoints = symnum_0.1_star$cutpoints
  ) %>%
  filter(p.adj < 0.1) %>%
  rstatix::add_xy_position()



theme(legend.position = "none", strip.background = element_blank(), strip.text.y = element_text(size = 10), strip.clip = "off", strip.placement = "outside", panel.spacing = unit(0, "mm")) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.05))) +
  labs(x = NULL, y = NULL) +
  guides(x = guide_axis(angle = 45))

pl <- df %>%
  ggplot(aes(x = time_point, y = tum_prop)) +
  geom_boxplot(outliers = F, size = 0.5, aes(fill = time_point), width = 0.9) +
  geom_quasirandom(size = 0.75, alpha = 0.6, color = "black", aes(fill = time_point), stroke = 0.25, pch = 21, width = 0.3, show.legend = F) +
  guides(x = guide_axis(angle = 45)) +
  labs(y = "Tumor cell proportion", x = NULL) +
  scale_color_manual(values = colors$time_point) +
  scale_fill_manual(values = colors$time_point) +
  facet_grid(. ~ .data[[use_col]], scales = "free", space = "free", switch = "x") +
  ggprism::add_pvalue(p_df, tip.length = 0, label = "p.adj.signif", step.increase = -0.01, bracket.nudge.y = 0, bracket.size = 0.2, label.size = 2.5) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.25), expand = expansion(mult = c(0.05, 0.1))) +
  theme(
    axis.text.x = element_blank(),
    axis.title.y = element_text(hjust = 0.2),
    axis.ticks.x = element_blank(),
    strip.text.x = element_markdown(angle = 90, hjust = 1, size = 8),
    strip.background.x = element_part_rect(side = "t", linewidth = 0.5),
    strip.placement = "inside",
    panel.spacing = unit(1, "mm"),
    strip.clip = "off",
    legend.position = "inside",
    legend.position.inside = c(0.8, 0.9),
    legend.title.position = "top",
    legend.title = element_text(hjust = 0.5),
    legend.background = element_blank(),
    axis.line.x = element_blank()
  ) +
  labs(fill = "Timepoint") +
  guides(
    x = guide_axis(cap = "none"),
    y = guide_axis(cap = "upper"),
    fill = guide_legend(nrow = 1, keywidth = 0.75, keyheight = 0.75)
  ) +
  coord_cartesian(clip = "off")
ggsave(pl, file = here("plots", "epi_cluster_proportions.pdf"), width = 5, height = 3)
pl
```


#### Signature Heatmap

```{r, fig.width = 8, fig.height = 9}
emt_paths <- grep("emt|HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION", ignore.case = T, colnames(srt_epi@meta.data), value = T)
myc_paths <- c(grep("myc", ignore.case = T, colnames(srt_epi@meta.data), value = T))
stemness <- c("WONG_EMBRYONIC_STEM_CELL_CORE", "BENPORATH_PRC2_TARGETS")
metabolic_paths <- c("HALLMARK_GLYCOLYSIS", "HALLMARK_OXIDATIVE_PHOSPHORYLATION", "HALLMARK_MTORC1_SIGNALING")
histological_malignant <- c("LUAD", "LUSC", "SCLC")
histological <- c("TRAVAGLINI_LUNG_ALVEOLAR_EPITHELIAL_TYPE_1_CELL", "TRAVAGLINI_LUNG_ALVEOLAR_EPITHELIAL_TYPE_2_CELL", "alveolar_malignant", "TRAVAGLINI_LUNG_BASAL_CELL", "TRAVAGLINI_LUNG_CILIATED_CELL", "TRAVAGLINI_LUNG_CLUB_CELL", "TRAVAGLINI_LUNG_NEUROENDOCRINE_CELL", "TRAVAGLINI_LUNG_PROLIFERATING_BASAL_CELL", "TRAVAGLINI_LUNG_PROXIMAL_BASAL_CELL")
cell_cycle <- c(grep("cycle", colnames(srt_epi@meta.data), ignore.case = T, value = T))
mapk <- c("REACTOME_MAPK_FAMILY_SIGNALING_CASCADES", "REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES", "BIOCARTA_MAPK_PATHWAY", "KEGG_MAPK_SIGNALING_PATHWAY", "REACTOME_SIGNALING_BY_EGFR")


pth_list <- list(
  "EMT" = emt_paths,
  "MYC" = myc_paths,
  "Stem" = stemness,
  "Metabolic" = metabolic_paths,
  "Lung Cancer" = histological_malignant,
  "Histological" = histological,
  "Cell Cycle" = cell_cycle,
  "MAPK" = mapk
)

pth_df <- DotPlotter(srt_epi, features = pth_list, cluster_groups = F, cluster_feats = F, group.by = "cell_type_epi")$data
```


```{r, fig.width = 5, fig.height = 9}
# Adjust names and filter
pth_df <- pth_df %>%
  filter(!grepl("_epithelial$", features.plot)) %>%
  filter(!grepl("TGFB", features.plot)) %>%
  mutate(pathway = case_when(
    grepl("^TRAVAGLINI_LUNG_", features.plot) ~ str_to_title(gsub("_", " ", gsub("TRAVAGLINI_LUNG_", "", features.plot))),
    features.plot == "alveolar_malignant" ~ "Alveolar Signature",
    grepl("^HALLMARK", features.plot) ~ str_to_title(gsub("_", " ", gsub("HALLMARK_", "", features.plot))),
    features.plot == "emt_i_malignant" ~ "EMT (I)",
    features.plot == "emt_ii_malignant" ~ "EMT (II)",
    features.plot == "emt_iii_malignant" ~ "EMT (III)",
    features.plot == "emt_iv_malignant" ~ "EMT (IV)",
    features.plot == "emt_i_malignant" ~ "EMT (I)",
    features.plot == "emt_i_malignant" ~ "EMT (I)",
    features.plot == "S.Score" ~ "S Score",
    features.plot == "G2M.Score" ~ "G2M Score",
    features.plot == "cell_cycle_g1_s_malignant" ~ "G1S",
    features.plot == "cell_cycle_g2_m_malignant" ~ "G2M",
    features.plot == "cell_cycle_hmg_rich_malignant" ~ "HMG Rich",
    features.plot == "myc_malignant" ~ "MYC",
    features.plot == "WONG_EMBRYONIC_STEM_CELL_CORE" ~ "Embryonic Stem Cell (Wong)",
    features.plot == "BENPORATH_PRC2_TARGETS" ~ "PRC2 Targets (Ben-Porath)",
    features.plot == "REACTOME_MAPK_FAMILY_SIGNALING_CASCADES" ~ "MAPK (Reactome)",
    features.plot == "REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES" ~ "Nuclear MAPK (Reactome)",
    features.plot == "REACTOME_ONCOGENIC_MAPK_SIGNALING" ~ "Oncogenic MAPK (Reactome)",
    features.plot == "BIOCARTA_MAPK_PATHWAY" ~ "MAPK (Biocarta)",
    features.plot == "KEGG_MAPK_SIGNALING_PATHWAY" ~ "MAPK (KEGG)",
    features.plot == "REACTOME_SIGNALING_BY_EGFR" ~ "EGFR (Reactome)",
    .default = features.plot
  )) %>%
  # Fix additional
  mutate(pathway = case_when(
    pathway == "Mtorc1 Signaling" ~ "MTORC1",
    pathway == "Epithelial Mesenchymal Transition" ~ "EMT",
    pathway == "Alveolar Epithelial Type 2 Cell" ~ "AT2 Cell",
    pathway == "Alveolar Epithelial Type 1 Cell" ~ "AT1 Cell",
    pathway == "Oxidative Phosphorylation" ~ "OxPhos",
    grepl("Myc", pathway) ~ gsub("Myc", "MYC", pathway),
    .default = pathway
  )) %>%
  mutate(
    pathway_source = case_when(
      grepl("_malignant$", features.plot) ~ "Malignant Metaprograms",
      grepl("_epithelial$", features.plot) ~ "Epithelial Metaprograms",
      grepl("^TRAVAGLINI", features.plot) ~ "Travaglini",
      grepl("^HALLMARK_", features.plot) ~ "Hallmark Pathways",
      features.plot %in% c("S.Score", "G2M.Score") ~ "Seurat",
      .default = "Curated"
    )
  ) %>%
  # Remove extra myc pathways since they all show the same thing
  filter(
    !(grepl("MYC", features.plot, ignore.case = T) & pathway_source == "Curated")
  ) %>%
  arrange(desc(pathway_source), desc(pathway))

# constructive::construct(unique(pth_df$pathway))
pth_cat_order <- c(
  "Histological", "Lung Cancer", "MAPK", "EMT", "MYC",
  "Stem", "Metabolic", "Cell Cycle"
)

pth_order <- c(
  "G1S", "G2M", "G2M Score", "HMG Rich", "S Score",
  "EMT (I)", "EMT (II)", "EMT (III)", "EMT (IV)", "EMT",
  "LUAD", "LUSC", "SCLC",
  "AT1 Cell", "AT2 Cell", "Alveolar Signature", "Basal Cell", "Proximal Basal Cell",
  "Proliferating Basal Cell", "Club Cell", "Ciliated Cell",
  "Neuroendocrine Cell", "MYC", "MYC Targets V1", "MYC Targets V2",
  "Glycolysis", "Oxidative Phosphorylation", "MTORC1", "Embryonic Stem Cell (Wong)",
  "PRC2 Targets (Ben-Porath)",
  "MAPK (KEGG)", "MAPK (Biocarta)", "MAPK (Reactome)", "Nuclear MAPK (Reactome)", "EGFR (Reactome)"
)

pth_df <- pth_df %>%
  mutate(
    pathway = fct_relevel(pathway, rev(pth_order)),
    feature.groups = fct_relevel(feature.groups, pth_cat_order)
  )

p1 <- pth_df %>%
  ggplot(aes(x = "Pathway Source", y = pathway)) +
  geom_tile(aes(fill = pathway_source), color = "black") +
  scale_fill_pander() +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  theme(
    axis.line.x = element_blank(),
    axis.line.y = element_blank(), legend.position = "left", strip.background = element_blank(), strip.text = element_blank()
  ) +
  guides(
    x = guide_axis(angle = 90),
    fill = guide_legend(keywidth = 0.5, keyheight = 0.5)
  ) +
  facet_grid(feature.groups ~ ., scales = "free", space = "free") +
  labs(y = NULL, x = NULL, fill = "Pathway Source")


p2 <- pth_df %>%
  ggplot(aes(x = id, y = pathway)) +
  geom_tile(aes(fill = avg.exp.scaled), color = "black") +
  scale_fill_viridis_c(option = "inferno") +
  facet_grid(feature.groups ~ ., scales = "free", space = "free") +
  guides(
    x = guide_axis(angle = 90),
    x.sec = guide_axis(angle = 90),
    fill = guide_colorbar(keywidth = 0.4, keyheight = 3)
  ) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  theme(
    axis.line = element_blank(), panel.border = element_blank(),
    strip.background.y = element_blank(),
    strip.text.y.right = element_text(angle = 0, hjust = 0)
  ) +
  labs(y = NULL, x = NULL, fill = "Scaled\nExpression")

sig_heatplot <- (p1 + theme(plot.margin = margin(l = 0.25, r = 1, unit = "mm"))) + (p2 + theme(plot.margin = margin(r = 0.25, unit = "mm"), axis.text.y = element_blank(), axis.ticks.y = element_blank())) + plot_layout(guides = "collect", ncol = 2, widths = c(0.03, 1)) + plot_annotation(theme = theme(legend.position = "bottom")) &
  theme(panel.spacing = unit(1, "mm"), panel.border = element_rect(fill = NA, color = "black", linewidth = 0.5), legend.box.margin = margin()) &
  coord_cartesian(clip = "off")
ggsave(sig_heatplot, file = here("plots", "epi_types_signature_heatmap.pdf"), width = 5, height = 9)
sig_heatplot
```

#### Signature tumor boxplots

```{r, fig.width = 8, fig.height = 3}
pths <- map(pathways$pathway, .f = function(pth) {
  grep(pth, colnames(srt_epi@meta.data), value = T)
}) %>%
  unlist() %>%
  unique()

pls <- map(pths, .f = function(g) {
  sample_plots <- srt_epi@meta.data %>%
    filter(is_tumor_cell) %>%
    ggplot(aes(x = fct_reorder(sample_id_new, .data[[g]], mean), y = .data[[g]])) +
    geom_violin(scale = "width", aes(fill = time_point)) +
    geom_boxplot(fill = "white", width = 0.2, outliers = F) +
    # ggrastr::rasterize(geom_jitter(size = 0.05, alpha = 0.25), dpi = 300) +
    facet_grid(. ~ time_point, scales = "free", space = "free") +
    guides(x = guide_axis(angle = 90)) +
    labs(x = "Sample ID", y = glue("{g} signature score")) +
    theme(
      legend.position = "none",
      strip.background.x = element_part_rect(side = ""),
      panel.border = element_rect(fill = NA),
      axis.line = element_blank()
    ) +
    scale_fill_manual(values = colors$time_point)

  summarized_plots <- srt_epi@meta.data %>%
    group_by(sample_id_new, time_point, is_tumor_cell) %>%
    filter(is_tumor_cell) %>%
    dplyr::summarize(mean_exp = mean(.data[[g]])) %>%
    ggplot(aes(x = time_point, y = mean_exp, color = time_point)) +
    geom_boxplot(outliers = F) +
    geom_quasirandom(size = 0.5) +
    labs(x = "Time Point", y = glue("LUAD signature"), color = "Timepoint") +
    guides(
      x = guide_axis(angle = 0),
      y = guide_axis(cap = "upper")
    ) +
    ggpubr::geom_pwc(method = "t_test", tip.length = 0, label = "p.adj.signif") +
    theme(legend.position = "none") +
    coord_cartesian(clip = "off") +
    scale_color_manual(values = colors$time_point) +
    labs(x = NULL, y = glue("{g} signature score\n(per sample mean)"))
  summarized_plots
})
names(pls) <- pths

pdf(file = here("plots", "epi_tumor_summarized_sigs_boxplot.pdf"), width = 2.5, height = 4)
print(pls)
dev.off()
```


Single table so I can plot select pathways of interest

```{r, fig.width = 8, fig.height = 9}
select_paths <- unique(sig_heatplot$data$features.plot)
select_paths_name <- unique(sig_heatplot$data$pathway)

all_sig_summary <- srt_epi@meta.data %>%
  group_by(sample_id_new, time_point, is_tumor_cell) %>%
  filter(is_tumor_cell) %>%
  summarize(across(all_of(select_paths), ~ mean(.x, na.rm = T))) %>%
  pivot_longer(cols = all_of(select_paths), names_to = "features.plot", values_to = "Score") %>%
  group_by(features.plot) %>%
  mutate(Score_Scaled = scale(Score)[, 1]) %>%
  ungroup() %>%
  mutate(pathway = case_when(
    grepl("^TRAVAGLINI_LUNG_", features.plot) ~ str_to_title(gsub("_", " ", gsub("TRAVAGLINI_LUNG_", "", features.plot))),
    features.plot == "alveolar_malignant" ~ "Alveolar Signature",
    grepl("^HALLMARK", features.plot) ~ str_to_title(gsub("_", " ", gsub("HALLMARK_", "", features.plot))),
    features.plot == "emt_i_malignant" ~ "EMT (I)",
    features.plot == "emt_ii_malignant" ~ "EMT (II)",
    features.plot == "emt_iii_malignant" ~ "EMT (III)",
    features.plot == "emt_iv_malignant" ~ "EMT (IV)",
    features.plot == "emt_i_malignant" ~ "EMT (I)",
    features.plot == "emt_i_malignant" ~ "EMT (I)",
    features.plot == "S.Score" ~ "S Score",
    features.plot == "G2M.Score" ~ "G2M Score",
    features.plot == "cell_cycle_g1_s_malignant" ~ "G1S",
    features.plot == "cell_cycle_g2_m_malignant" ~ "G2M",
    features.plot == "cell_cycle_hmg_rich_malignant" ~ "HMG Rich",
    features.plot == "myc_malignant" ~ "MYC",
    features.plot == "WONG_EMBRYONIC_STEM_CELL_CORE" ~ "Embryonic Stem Cell (Wong)",
    features.plot == "BENPORATH_PRC2_TARGETS" ~ "PRC2 Targets (Ben-Porath)",
    features.plot == "REACTOME_MAPK_FAMILY_SIGNALING_CASCADES" ~ "MAPK (Reactome)",
    features.plot == "REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES" ~ "Nuclear MAPK (Reactome)",
    features.plot == "REACTOME_ONCOGENIC_MAPK_SIGNALING" ~ "Oncogenic MAPK (Reactome)",
    features.plot == "BIOCARTA_MAPK_PATHWAY" ~ "MAPK (Biocarta)",
    features.plot == "KEGG_MAPK_SIGNALING_PATHWAY" ~ "MAPK (KEGG)",
    features.plot == "REACTOME_SIGNALING_BY_EGFR" ~ "EGFR (Reactome)",
    .default = features.plot
  )) %>%
  # Fix additional
  mutate(pathway = case_when(
    pathway == "Mtorc1 Signaling" ~ "MTORC1",
    pathway == "Epithelial Mesenchymal Transition" ~ "EMT",
    pathway == "Alveolar Epithelial Type 2 Cell" ~ "AT2 Cell",
    pathway == "Alveolar Epithelial Type 1 Cell" ~ "AT1 Cell",
    pathway == "Oxidative Phosphorylation" ~ "OxPhos",
    grepl("Myc", pathway) ~ gsub("Myc", "MYC", pathway),
    .default = pathway
  ))

p <- all_sig_summary %>%
  filter(pathway %in% select_paths_name) %>%
  ggplot(aes(x = time_point, y = Score)) +
  geom_boxplot(outliers = F, size = 0.5, aes(fill = time_point), width = 0.9) +
  geom_quasirandom(size = 1, alpha = 0.6, color = "black", aes(fill = time_point), stroke = 0.25, pch = 21, width = 0.3) +
  facet_wrap(~pathway, scales = "free", strip.position = "left", ncol = 6) +
  scale_fill_manual(values = colors$time_point) +
  coord_cartesian(clip = "off") +
  theme(legend.position = "none", strip.background = element_blank(), strip.clip = "off", strip.placement = "outside") +
  ggpubr::geom_pwc(tip.length = 0, method = "t_test", label.size = 2, label = "p.adj.signif", p.adjust.method = "BH") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  labs(x = "Timepoint") +
  guides(x = guide_axis(angle = 45))
ggsave(p, file = here("plots", "epi_sig_tumor_specific_boxplots_select.pdf"), width = 8.5, height = 11)
p
```


##### For Main

```{r, fig.width = 8, fig.height = 2.5}
select_paths_main <- c("Alveolar Signature", "LUAD", "Nuclear MAPK (Reactome)", "MYC", "Proliferating Basal Cell", "EMT")
p <- all_sig_summary %>%
  filter(pathway %in% select_paths_main) %>%
  ggplot(aes(x = time_point, y = Score)) +
  # geom_violin(aes(fill = time_point), scale = "width") +
  geom_boxplot(outliers = F, size = 0.5, aes(fill = time_point), width = 0.9) +
  geom_quasirandom(size = 1, alpha = 0.6, color = "black", aes(fill = time_point), stroke = 0.25, pch = 21, width = 0.3) +
  facet_wrap(~ fct_relevel(pathway, select_paths_main), scales = "free", nrow = 1, strip.position = "left") +
  scale_color_manual(values = colors$time_point) +
  scale_fill_manual(values = colors$time_point) +
  coord_cartesian(clip = "off") +
  theme(legend.position = "none", strip.background = element_blank(), strip.text.y = element_text(size = 10), strip.clip = "off", strip.placement = "outside", panel.spacing = unit(0, "mm")) +
  ggpubr::geom_pwc(tip.length = 0, method = "t_test", label.size = 2, label = "p.adj.signif", p.adjust.method = "BH", step.increase = 0.075) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.05))) +
  labs(x = NULL, y = NULL) +
  guides(x = guide_axis(angle = 45))
ggsave(p, file = here("plots", "epi_tumor_boxplot_main.pdf"), width = 8, height = 2.6)
p
```



## Figure 4 -- MAPKalt

```{r}
srt_tum <- srt_epi[, srt_epi$is_tumor_cell]
```


### Pathway overlaps?

```{r}
plist <- split(pathways_v2$Gene, pathways_v2$pathway)

meta_programs_df <- read.table(file = "https://raw.githubusercontent.com/mjz1/meta_programs_tirosh/refs/heads/main/tirosh_mp_patched.txt", header = T, sep = "\t")

meta_programs <- split(meta_programs_df, meta_programs_df$cell_type)

meta_programs <- map(meta_programs, .f = function(x) {
  res <- split(x, x$meta_program)
  map(res, .f = function(y) y$Gene)
})

names(meta_programs) <- janitor::make_clean_names(names(meta_programs))

sel_paths <- c("LUAD_GIRARD", "REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES", "BIOCARTA_MAPK_PATHWAY", "KEGG_MAPK_SIGNALING_PATHWAY")

plist_sel <- list(
  "Alveolar" = meta_programs$malignant$alveolar,
  "LUAD" = plist$LUAD_GIRARD,
  "Nuclear MAPK (Reactome)" = plist$REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES,
  "MAPK (KEGG)" = plist$KEGG_MAPK_SIGNALING_PATHWAY,
  "MAPK (Biocarta)" = plist$BIOCARTA_MAPK_PATHWAY
)

plist_sel <- map(plist_sel, unique)

library(eulerr)
fit <- euler(plist_sel)
pdf(here("plots", "mapk_path_overlaps.pdf"), width = 5, height = 5)
plot(fit, quantities = TRUE, legend = T)
dev.off()


plist_sel <- list(
  "Alveolar" = meta_programs$malignant$alveolar,
  "LUAD" = plist$LUAD_GIRARD,
  "LUSC" = plist$LUSC_GIRARD,
  "SCLC" = plist$SCLC_CCLE,
  "Nuclear MAPK (Reactome)" = plist$REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES,
  "MAPK (KEGG)" = plist$KEGG_MAPK_SIGNALING_PATHWAY,
  "MAPK (Biocarta)" = plist$BIOCARTA_MAPK_PATHWAY
)

plist_sel <- map(plist_sel, unique)

library(eulerr)
fit <- euler(plist_sel)
pdf(here("plots", "mapk_path_overlaps_extended.pdf"), width = 7, height = 7)
plot(fit, quantities = TRUE, legend = F)
dev.off()
```


### MAPK correlations

Supervised pathway correlations

```{r, fig.width = 9, fig.height = 3}
paths <- c("Nuclear MAPK (Reactome)", "MAPK (Biocarta)", "MAPK (KEGG)")
all_sig_summary_wide <- all_sig_summary %>%
  pivot_wider(id_cols = c("sample_id_new", "time_point"), names_from = pathway, values_from = Score)



pls <- map(paths, .f = function(pth) {
  p <- all_sig_summary_wide %>%
    ggplot(aes(x = .data[[pth]], y = LUAD)) +
    stat_ellipse(
      data = all_sig_summary_wide[all_sig_summary_wide$time_point != "MRD", ],
      aes(fill = time_point),
      type = "t",
      geom = "polygon",
      alpha = 0.3, show.legend = F
    ) +
    geom_point(aes(fill = time_point), pch = 21, stroke = 0.5, size = 2.5) +
    geom_line(stat = "smooth", method = "lm", color = "#063970", se = F, linewidth = 1.25, alpha = 0.75) +
    ggpubr::stat_cor() +
    scale_fill_manual(values = colors$time_point, breaks = names(colors$time_point)) +
    theme_classic() +
    labs(y = "LUAD Signature", fill = "Timepoint", x = pth) +
    guides(fill = guide_legend(override.aes = list(size = 3.5)))
})

pcomb <- wrap_plots(pls, nrow = 1, guides = "collect", axes = "collect") & theme(panel.border = element_rect(fill = NA, color = "black"), axis.line = element_blank())
ggsave(pcomb, file = here("plots", "luad_mapk_select_cors.pdf"), width = 8, height = 2.75)
pcomb
```


```{r}
paths <- c("Nuclear MAPK (Reactome)", "MAPK (Biocarta)", "MAPK (KEGG)")
all_sig_summary_wide <- all_sig_summary %>%
  pivot_wider(id_cols = c("sample_id_new", "time_point"), names_from = pathway, values_from = Score)



pls <- map(paths, .f = function(pth) {
  p <- all_sig_summary_wide %>%
    ggplot(aes(x = .data[[pth]], y = .data[["Alveolar Signature"]])) +
    stat_ellipse(
      data = all_sig_summary_wide[all_sig_summary_wide$time_point != "MRD", ],
      aes(fill = time_point),
      type = "t",
      geom = "polygon",
      alpha = 0.3, show.legend = F
    ) +
    geom_point(aes(fill = time_point), pch = 21, stroke = 0.5, size = 2.5) +
    geom_line(stat = "smooth", method = "lm", color = "#063970", se = F, linewidth = 1.25, alpha = 0.75) +
    ggpubr::stat_cor() +
    scale_fill_manual(values = colors$time_point, breaks = names(colors$time_point)) +
    # scale_color_manual(values = colors$time_point) +
    theme_classic() +
    labs(y = "Alveolar Signature", fill = "Timepoint", x = pth) +
    guides(fill = guide_legend(override.aes = list(size = 3.5)))
})

pcomb <- wrap_plots(pls, nrow = 1, guides = "collect", axes = "collect") & theme(panel.border = element_rect(fill = NA, color = "black"), axis.line = element_blank())
ggsave(pcomb, file = here("plots", "alveolar_mapk_select_cors.pdf"), width = 8, height = 2.75)
pcomb
```

LUAD vs alveolar

```{r}
p <- all_sig_summary_wide %>%
  ggplot(aes(x = LUAD, y = .data[["Alveolar Signature"]])) +
  stat_ellipse(
    data = all_sig_summary_wide[all_sig_summary_wide$time_point != "MRD", ],
    aes(fill = time_point),
    type = "t",
    geom = "polygon",
    alpha = 0.3, show.legend = F
  ) +
  geom_point(aes(fill = time_point), pch = 21, stroke = 0.5, size = 2.5) +
  geom_line(stat = "smooth", method = "lm", color = "#063970", se = F, linewidth = 1.25, alpha = 0.75) +
  ggpubr::stat_cor() +
  scale_fill_manual(values = colors$time_point, breaks = names(colors$time_point)) +
  # scale_color_manual(values = colors$time_point) +
  theme_classic() +
  labs(y = "Alveolar Signature", fill = "Timepoint", x = "LUAD Signature") +
  guides(fill = guide_legend(override.aes = list(size = 3.5)))
p
ggsave(p, file = here("plots", "luad_vs_alveolar_tumorcells.pdf"), width = 4, height = 4)
```



```{r, fig.width = 9, fig.height = 3}
pcomb2 <- wrap_plots(pls, ncol = 1, guides = "collect") + plot_layout(axes = "collect") & theme(panel.border = element_rect(fill = NA, color = "black"), axis.line = element_blank()) # legend.position = "bottom", legend.direction = "vertical", legend.title.position = "left"
ggsave(pcomb2, file = here("plots", "luad_mapk_select_cors_vertical.pdf"), width = 3.9, height = 7)
# pcomb
```

Omnibus

```{r}
corres <- all_sig_summary_wide %>%
  select(all_of(colnames(all_sig_summary_wide)[3:ncol(all_sig_summary_wide)])) %>%
  as.matrix() %>%
  Hmisc::rcorr(type = "pearson")

library(ggcorrplot)
cormat <- corres$r

pmat <- corres$P

pdat <- ggcorrplot(
  corr = cormat,
  type = "lower",
  hc.order = T,
  p.mat = pmat,
  insig = "pch"
)$data %>%
  rstatix::add_significance(p.col = "pvalue") %>%
  rstatix::adjust_pvalue(p.col = "pvalue", method = "bonferroni") %>%
  rstatix::add_significance(p.col = "pvalue.adj") %>%
  mutate(sig = ifelse(pvalue.adj < 0.05, "*", "ns"))

p <- ggcorrplot(
  corr = cormat,
  type = "lower",
  hc.order = T,
  legend.title = bquote("Pearson's R")
)

ylab_anno <- data.frame(lab = levels(pdat$Var1), y = levels(pdat$Var1), x = (1:ncol(cormat)) - 1)
axis_tx_sz <- 7

p <- p +
  geom_text(data = pdat[pdat$pvalue.adj < 0.05, ], aes(label = sig), size = 3, nudge_y = -0.3) +
  # labs(title = str_split(nm, pattern = "__", simplify = T)[2]) +
  guides(x = guide_axis(angle = 90)) +

  geom_text(inherit.aes = F, data = ylab_anno, aes(x = x, y = y, label = lab), hjust = 1, size = axis_tx_sz * 0.33) +
  coord_cartesian(clip = "off") +
  theme(
    axis.text.y.left = element_blank(),
    axis.text.x.bottom = element_text(size = axis_tx_sz), plot.margin = margin(1, 1, 1, 10, unit = "lines"),
    panel.grid = element_blank()
  )
p
```


```{r}
cor_res <- all_sig_summary_wide %>%
  correlation::correlation(select = colnames(all_sig_summary_wide)[3:ncol(all_sig_summary_wide)])

p <- plot(cor_res) +
  ggraph::scale_edge_color_gradient2(low = scales::muted("blue"), high = scales::muted("red"))

p2 <- plot(summary(cor_res), show_data = "points") +
  guides(x = guide_axis(angle = 90))
```

### Within sample correlations

```{r}
library(easystats)
x <- srt_epi@meta.data %>%
  filter(is_tumor_cell) %>%
  mutate(sample_id_new = factor(sample_id_new)) %>%
  add_count(sample_id_new, name = "n_cells") %>%
  dplyr::group_by(sample_id_new) %>%
  correlation::correlation(select = c("LUAD", "LUSC", "SCLC"), select2 = c("BIOCARTA_MAPK_PATHWAY", "KEGG_MAPK_SIGNALING_PATHWAY", "REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES"), method = "spearman") %>%
  left_join(distinct(srt_epi@meta.data[, c("sample_id_new", "time_point")]), by = c("Group" = "sample_id_new")) %>%
  mutate(
    mapk_path = case_when(
      Parameter2 == "BIOCARTA_MAPK_PATHWAY" ~ "MAPK (Biocarta)",
      Parameter2 == "KEGG_MAPK_SIGNALING_PATHWAY" ~ "MAPK (KEGG)",
      Parameter2 == "REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES" ~ "Nuclear MAPK (Reactome)",
      .default = Parameter2
    )
  )
```


Plot for main

```{r, fig.width = 6, fig.height = 3.5}
p <- x %>%
  filter(Parameter2 == "REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES") %>%
  arrange(desc(p)) %>%
  ggplot(aes(x = Parameter1, y = rho)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  gghalves::geom_half_boxplot(outliers = F, aes(fill = Parameter1), errorbar.draw = T) +
  gghalves::geom_half_point_panel(aes(size = -log10(p), fill = Parameter1, alpha = n_Obs), pch = 21) +
  facet_nested("Within sample correlations" ~ mapk_path + time_point) +
  scale_fill_manual(values = colors$histology_predominant_short) +
  scale_alpha_continuous(trans = "log10", breaks = scales::breaks_log(), labels = scales::label_number(big.mark = ",")) +
  guides(
    fill = "none",
    x = guide_axis(angle = 0),
    size = guide_legend(override.aes = list(pch = 19)),
    alpha = guide_legend(override.aes = list(pch = 16, size = 3))
  ) +
  ggpubr::geom_pwc(tip.length = 0, label = "p.adj.signif", step.increase = 0.05, vjust = 0.5, hide.ns = T, bracket.nudge.y = 0.075) +
  theme_few() +
  theme(
    strip.clip = "off",
    plot.title = element_text(hjust = 0.5, size = 12), plot.title.position = "panel",
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(size = 8)
  ) +
  labs(x = "Lineage Signature", y = bquote("Spearman's" ~ rho), size = bquote("-log10"[Padj]), alpha = "No. cells")
ggsave(p, file = here("plots", "per_sample_mapk_cor_main.pdf"), width = 6, height = 3.5)
p
```


Plot for supp

```{r, fig.width = 11, fig.height = 3.5}
p <- x %>%
  filter(Parameter2 != "REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES") %>%
  arrange(desc(p)) %>%
  ggplot(aes(x = Parameter1, y = rho)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  gghalves::geom_half_boxplot(outliers = F, aes(fill = Parameter1), errorbar.draw = T) +
  gghalves::geom_half_point_panel(aes(size = -log10(p), fill = Parameter1, alpha = n_Obs), pch = 21) +
  facet_nested("Within sample correlations" ~ mapk_path + time_point) +
  scale_fill_manual(values = colors$histology_predominant_short) +
  scale_alpha_continuous(trans = "log10", breaks = scales::breaks_log(), labels = scales::label_number(big.mark = ",")) +
  guides(
    fill = "none",
    x = guide_axis(angle = 0),
    size = guide_legend(override.aes = list(pch = 19)),
    alpha = guide_legend(override.aes = list(pch = 16, size = 3))
  ) +
  ggpubr::geom_pwc(tip.length = 0, label = "p.adj.signif", step.increase = 0.05, vjust = 0.5, hide.ns = T, bracket.nudge.y = 0.075) +
  theme_few() +
  theme(
    strip.background.x = element_part_rect(side = "b", fill = NA, color = "black"),
    strip.clip = "off",
    plot.title = element_text(hjust = 0.5, size = 12), plot.title.position = "panel",
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(size = 8)
  ) +
  labs(x = "Lineage Signature", y = bquote("Spearman's" ~ rho), size = bquote("-log10"[Padj]), alpha = "No. cells")
ggsave(p, file = here("plots", "per_sample_mapk_cor_supp.pdf"), width = 11, height = 3.5)
p
```



```{r}
attributes(x)$modelframe %>%
  ggplot(aes(x = REACTOME_MAPK_TARGETS_NUCLEAR_EVENTS_MEDIATED_BY_MAP_KINASES, y = LUAD)) +
  geom_density2d_filled() +
  ggpubr::stat_cor(method = "spearman") +
  scattermore::geom_scattermore(pointsize = 1, pixels = c(256, 256)) +
  geom_smooth(method = "lm")
```


### MAPKalt LUAD

First just use IMPACT definitions

```{r, fig.width = 15, fig.height = 9}
library(gt)
supp <- read.table(here("metadata", "supplemental_mapk_mutation_table_edit.txt"), header = T, sep = "\t")
supp$useInferCNV <- grepl("inferCNV", supp$notes_other)

useinfer <- supp$sample_id[supp$useInferCNV]

samp_ord <- srt_tum@meta.data %>%
  filter(time_point == "PD", sample_id_new %in% srt_tum@meta.data$sample_id_new) %>%
  mutate(ord = fct_reorder(sample_id_new, LUAD, mean)) %>%
  pull(ord) %>%
  levels()

p0 <- db$metadata %>%
  filter(time_point == "PD", sample_id_new %in% srt_tum@meta.data$sample_id_new) %>%
  select(sample_id, sample_id_new, has_impact, EGFR_Primary, EGFR_Secondary, MAPKalt) %>%
  mutate(sample_id_new = fct_relevel(sample_id_new, samp_ord)) %>%
  mutate(useInfer = sample_id %in% useinfer) %>%
  arrange(desc(sample_id_new)) %>%
  gt::gt(rowname_col = "sample_id_new")

p1 <- srt_tum@meta.data %>%
  filter(time_point == "PD") %>%
  ggplot(aes(y = fct_relevel(sample_id_new, samp_ord), x = LUAD)) +
  geom_violin(scale = "width", aes(fill = time_point), show.legend = F) +
  geom_pointrange(
    stat = "summary", fatten = 0.1,
    fun.min = function(z) {
      quantile(z, 0.25)
    },
    fun.max = function(z) {
      quantile(z, 0.75)
    },
    fun = median, position = position_dodge(width = 0.8), orientation = "y"
  ) +
  # facet_grid(. ~ secondary_mapk_binned, scales = "free", space = "free") +
  # guides(x = guide_axis(angle = 90)) +
  labs(y = "Sample", x = "LUAD signature score") +
  scale_fill_manual(values = colors$time_point) +
  theme(
    axis.line = element_blank(),
    panel.border = element_rect(fill = NA, color = "black"),
    strip.background.x = element_blank()
  )

pcomb <- wrap_table(p0, space = "fixed") + p1
pcomb
```


Get inferCNV measures

```{r}
tumor_sce <- db$infercnv
tumor_sce <- tumor_sce[, colnames(srt_tum)]
tumor_sce <- scatools::overlap_genes(tumor_sce, ensDb = EnsDb.Hsapiens.v86::EnsDb.Hsapiens.v86, gene_biotype = "protein_coding")
impact_paths <- read.table(here("impact", "impact_mapping.txt"), header = T, sep = "\t")

tier1 <- c("EGFR", "MET", "ERBB2", "PIK3CA", "BRAF", "KRAS")
tier2 <- c("FGFR1", "FGFR2", "FGFR3", "FGFR4", "ALK", "RET", "ROS1", "RAC1", "MAPK1")


cna_df <- pbmcapply::pbmclapply(X = impact_paths$Gene[impact_paths$pathway == "rtk_ras"], mc.cores = 16, FUN = function(g) {
  try(df <- scatools::plot_gene_cna(sce = tumor_sce, gene = g, return_data = T)[, c("cell_id", g)])
})
names(cna_df) <- impact_paths$Gene[impact_paths$pathway == "rtk_ras"]
x <- map(cna_df, nrow)
genes_cnv <- names(cna_df[!(map(x, is.null) %>% unlist())])[-1]

cna_df <- cna_df[!(map(x, is.null) %>% unlist())] %>%
  purrr::reduce(full_join)

spl <- str_split(cna_df$cell_id, pattern = "_", simplify = T)
cna_df$sample_id <- spl[, 2]


cna_df_means <- cna_df %>%
  group_by(sample_id) %>%
  summarize(across(genes_cnv, ~ mean(.x, na.rm = T)))
```


```{r, fig.width=10, fig.height =5}
amp_thresh <- 1.05

cnv_df_long <- cna_df_means %>%
  pivot_longer(cols = !sample_id, values_to = "inferCNV", names_to = "Gene")
cnv_df_long$inferCNV_amp <- cnv_df_long$inferCNV > amp_thresh
cnv_df_long$inferCNV_amp_conf <- ifelse(cnv_df_long$inferCNV > 1.1, "high", "low")

cnv_df_long <- filter(cnv_df_long, inferCNV_amp) %>%
  unite(infercnv, Gene, inferCNV_amp_conf, sep = ": ", remove = F)

cnv_df_add <- cnv_df_long %>%
  filter(Gene %in% c("EGFR", "MET", "ERBB2", "FGFR1", "FGFR2", "FGFR3", "FGFR4", "PIK3CA", "BRAF", "KRAS")) %>%
  pivot_wider(id_cols = "sample_id", names_from = "inferCNV_amp_conf", values_from = "Gene", values_fn = function(x) paste0(x, collapse = ";"), values_fill = "", names_prefix = "infercnv_")

test <- left_join(db$metadata, cnv_df_add)
```


```{r}
p0 <- test %>%
  filter(time_point == "PD", sample_id_new %in% srt_tum@meta.data$sample_id_new) %>%
  select(sample_id, sample_id_new, has_impact, EGFR_Primary, EGFR_Secondary, MAPKalt, infercnv_low, infercnv_high) %>%
  mutate(sample_id_new = fct_relevel(sample_id_new, samp_ord)) %>%
  mutate(useInfer = sample_id %in% useinfer) %>%
  # arrange(desc(sample_id_new)) %>%
  gt::gt(rowname_col = "sample_id_new")
pcomb <- wrap_table(p0, space = "fixed") + p1
pcomb
```


```{r}
p0 <- test %>%
  filter(sample_id_new %in% srt_tum@meta.data$sample_id_new) %>%
  select(sample_id, sample_id_new, has_impact, EGFR_Primary, EGFR_Secondary, MAPKalt, infercnv_low, infercnv_high) %>%
  mutate(sample_id_new = fct_relevel(sample_id_new, samp_ord)) %>%
  mutate(useInfer = sample_id %in% useinfer) %>%
  # arrange(desc(sample_id_new)) %>%
  gt::gt(rowname_col = "sample_id_new")
```

Use this table to assign samples as secondary mapk or not

```{r}
p <- VlnPlot(srt_tum[, srt_tum$time_point == "PD"], group.by = "sample_id_new", sort = T, features = c(tier1, tier2), pt.size = 0, combine = F)
names(p) <- c(tier1, tier2)
```


TODO: Add for Pre and On

```{r}
secondary_mapk <- c(
  "PD1" = "EGFR_amp",
  "PD2" = "EGFR_amp",
  "PD3" = "EGFR_amp, ALK_fus",
  "PD4" = "MET_ampinf, BRAF_ampinf, EGFR_ampinf", # check focal amp signal -- these are whole chr events
  "PD5" = "EGFR_amp, KIT_amp",
  "PD6" = "NA",
  "PD7" = "EGFR_amp", # ROS1 high
  "PD8" = "MET_amp",
  "PD9" = "EGFR_amp, KRAS_amp",
  "PD10" = "MAP2K4_W175*", # Does this count? ROS1 high
  "PD11" = "ERBB2_amp, NF1_Q1235*", # ERBB2 not highly expressed, unclear if it belongs
  "PD12" = "MAP3K1_Q535*", # Does this count?
  "PD14" = "MET_ampinf", # Strange -- annotated as on-target from clinical table but not sure why -- different exon19 event
  "PD15" = "MET_amp",
  "PD16" = "EGFR_amp, MET_ampinf", # MET seems real inferCNV
  "PD17" = "ALK_fus, EGFR_ampinf",
  "PD18" = "NA", # No MET
  "PD20" = "NA", # No IMPACT
  "PD21" = "BRAF_fus",
  "PD22" = "EGFR_amp",
  "PD23" = "NA", # RET expression
  "PD24" = "NA",
  "PD25" = "NA", # Bit of KRAS
  "PD26" = "EGFR_C797S", # LUSC transformed; RAC1 high
  "PD27" = "EGFR_amp",
  "PD28" = "FGFR1_ampinf",
  "PD29" = "MAPK1_amp", # inferCNV agrees
  "PD30" = "MET_ampinf", # Need infercnv, met amp possibly
  "MRD1" = "NA",
  "MRD2" = "EGFR_amp",
  "MRD5" = "EGFR_amp",
  "MRD6" = "NA"
)

# Replace names

names(secondary_mapk) <- sample_id_map$sample_id_new[match(names(secondary_mapk), sample_id_map$sample_id_new_v1)]

mapk_df <- enframe(secondary_mapk, name = "sample_id", value = "secondary_mapk")
write.table(mapk_df, file = here("tables", "secondary_mapk.txt"), quote = F, sep = "\t", row.names = F, col.names = T)
no_mapk <- names(which(secondary_mapk == "NA"))
```


```{r}
srt_tum$secondary_mapk <- srt_tum@meta.data %>%
  mutate(secondary_mapk = case_when(
    time_point %in% c("TN", "MRD") ~ NA,
    sample_id_new %in% no_mapk ~ "Unknown",
    .default = "MAPKalt"
  )) %>%
  pull(secondary_mapk)

samp_ord <- srt_tum@meta.data %>%
  filter(time_point == "PD") %>%
  mutate(sample_id_new = fct_reorder(sample_id_new, -LUAD, median)) %>%
  pull(sample_id_new) %>%
  levels()


p1 <- srt_tum@meta.data %>%
  filter(time_point == "PD") %>%
  ggplot(aes(x = fct_relevel(sample_id_new, samp_ord), y = LUAD)) +
  # ggplot(aes(x = sample_id_new, y = LUAD)) +

  geom_violin(scale = "width", aes(fill = time_point), show.legend = F) +
  geom_pointrange(
    stat = "summary", fatten = 0.1,
    fun.min = function(z) {
      quantile(z, 0.25)
    },
    fun.max = function(z) {
      quantile(z, 0.75)
    },
    fun = median, position = position_dodge(width = 0.8)
  ) +
  facet_grid(. ~ secondary_mapk, scales = "free", space = "free") +
  guides(x = guide_axis(angle = 90)) +
  labs(x = "Sample", y = "LUAD score") +
  scale_fill_manual(values = colors$time_point) +
  theme_few() +
  theme(
    axis.line = element_blank(),
    panel.border = element_rect(fill = NA, color = "black"),
    strip.background.x = element_blank()
  )

p2 <- srt_tum@meta.data %>%
  filter(time_point == "PD") %>%
  group_by(sample_id, secondary_mapk, time_point) %>%
  summarize(LUAD = mean(LUAD)) %>%
  ggplot(aes(x = secondary_mapk, y = LUAD)) +
  geom_boxplot(outliers = F, aes(fill = time_point), show.legend = F) +
  geom_quasirandom(size = 0.3, pch = 1) +
  ggpubr::geom_pwc(tip.length = 0, label = "p.signif") +
  guides(x = guide_axis(angle = 45)) +
  scale_fill_manual(values = colors$time_point) +
  labs(x = NULL, y = "LUAD score\n(per sample means)") +
  coord_cartesian(clip = "off") +
  theme_classic()

pcomb <- p1 + p2 + plot_layout(nrow = 1, widths = c(1, 0.1), guides = "collect") & theme(axis.text.x = element_text(size = 6))
ggsave(pcomb, file = here("plots", "mapk_alt_luad_sig_relapsed_v2.pdf"), width = 5, height = 2.5)
pcomb
```


Create a plot with the mutations annotated in a bar on the side, and notable genes expression in a box heat encoding

```{r, fig.width = 6, fig.height = 4.25}
mapk_alt_df <- enframe(secondary_mapk, name = "sample_id_new", value = "variant_list") %>%
  separate_rows(variant_list, sep = ", ") %>%
  separate(variant_list, sep = "_", into = c("Gene", "Variant_Type"), remove = F) %>%
  mutate(
    infercnv = grepl("inf", Variant_Type),
    Variant_Type = gsub("inf", "", Variant_Type),
    Gene = case_when(
      grepl("NA", Gene) ~ NA,
      .default = Gene
    )
  ) %>%
  mutate(secondary_mapk = case_when(
    sample_id_new %in% no_mapk ~ "Unknown",
    .default = "MAPKalt"
  )) %>%
  tidyr::complete(nesting(sample_id_new, secondary_mapk), Gene, fill = list(Variant_Type = "NoAlt")) %>%
  mutate(Variant_Type = fct_recode(Variant_Type,
    "Amp" = "amp",
    "Fusion" = "fus"
  )) %>%
  filter(!grepl("MRD", sample_id_new))



var_order <- names(table(mapk_alt_df$Variant_Type) %>% sort(decreasing = T))
var_order <- c(var_order[-1], "NoAlt")

gene_ord <- mapk_alt_df[!is.na(mapk_alt_df$Gene) & mapk_alt_df$Variant_Type != "NoAlt", ]$Gene %>%
  table() %>%
  sort(decreasing = T)

p3 <- mapk_alt_df %>%
  filter(!is.na(Gene)) %>%
  ggplot(aes(y = fct_relevel(Gene, rev(names(gene_ord))), x = fct_relevel(sample_id_new, samp_ord))) +
  geom_tile(aes(fill = fct_relevel(Variant_Type, var_order))) +
  geom_point(
    data = mapk_alt_df[mapk_alt_df$infercnv & !is.na(mapk_alt_df$infercnv), ],
    aes(color = "inferCNV"), show.legend = F
  ) +
  guides(
    x = guide_axis(angle = 90),
    fill = guide_legend(order = 1, keywidth = 0.4, keyheight = 0.4, ncol = 2, theme = theme(legend.margin = margin()))
  ) +
  scale_fill_manual(
    values = c(
      "NoAlt" = "white",
      "Amp" = "#4477AA",
      "Fusion" = "#88CCEE",
      "C797S" = "#117733",
      "Q1235*" = "#DDCC77",
      "Q535*" = "#332288",
      "W175*" = "#999933"
    ),
    breaks = var_order[-length(var_order)]
  ) +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_color_manual(values = "black") +
  # scale_x_discrete(drop = FALSE) +
  facet_grid(. ~ secondary_mapk, scales = "free", space = "free") +
  labs(x = "Sample", y = "Gene", fill = "Variant", color = NULL)
p1 <- p1 + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.title.x = element_blank())

p3 <- p3 + theme_few() + theme(strip.text.x = element_blank(), strip.background.x = element_blank(), legend.background = element_blank(), plot.background = element_blank(), panel.grid = element_line(color = "grey50", linetype = "solid", linewidth = 1, colour = "black"))

pcomb <- (p1) + free(p2, type = "space", side = "b") + free(p3, type = "space") + plot_spacer() + plot_layout(ncol = 2, widths = c(1, 0.1)) +
  plot_annotation(theme = theme(plot.margin = margin(3, 8, 15, 12, unit = "mm")))
ggsave(pcomb, file = here("plots", "mapk_alt_luad_sig_relapsed_plus_muts_v2.pdf"), width = 6, height = 4.25)

pcomb
```


Psuedobulk exp heatmap

```{r, fig.width = 6, fig.height = 3}
# feats <- c(
#   "EGFR", "MET", "ALK", "BRAF", "ERBB2", "ERBB3", "FGFR1", "FGFR2", "FGFR3", "KIT", "KRAS",
#   "ASCL1", "NEUROD1", "VIM", "IGF1R", "CDC42", "IGF1R", "ROS1", "PIK3CA", "NOTCH1", "MYC", "AKT1", "AXL", "YAP1"
# )
feats <- c("EGFR", "MET", "ALK", "BRAF", "ERBB2", "FGFR1", "KIT", "KRAS")
mapk_exp_df <- bulk_exp(srt_tum, cell_type_col = "is_tumor_cell", sample_id_col = "sample_id_new", meta_cols = c("time_point", "secondary_mapk"), features = feats)

mapk_exp_df <- mapk_exp_df %>%
  filter(time_point == "PD") %>%
  mutate(across(all_of(feats), ~ standardize(.x)))


mapk_exp_df_long <- mapk_exp_df %>%
  pivot_longer(cols = feats, names_to = "Gene", values_to = "exp")

p_df <- mapk_exp_df_long %>%
  group_by(Gene) %>%
  rstatix::wilcox_test(exp ~ secondary_mapk, detailed = T) %>%
  rstatix::adjust_pvalue(method = "BH") %>%
  rstatix::add_significance(
    p.col = "p", cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1),
    symbols = c("***", "**", "*", "^", "")
  )

p4_b <- p_df %>%
  ggplot(aes(x = "W", y = fct_relevel(Gene, rev(names(gene_ord))))) +
  geom_tile(aes(fill = estimate)) +
  geom_text(aes(label = p.signif), vjust = 0.8) +
  theme_few() +
  scale_y_discrete(expand = c(0, 0)) +
  scale_x_discrete(expand = c(0, 0)) +
  guides(fill = guide_colorbar(keywidth = 0.4, keyheight = 3)) +
  theme(panel.border = element_blank(), axis.text = element_blank(), axis.ticks = element_blank()) +
  labs(fill = "W statistic", y = NULL, x = NULL) +
  scale_fill_gradient2(low = scales::muted("blue"), high = scales::muted("red"))

p4 <- mapk_exp_df_long %>%
  ggplot(aes(x = fct_relevel(sample_id_new, samp_ord), y = as.numeric(fct_relevel(Gene, rev(names(gene_ord)))))) +
  geom_tile(aes(fill = exp), color = "grey10") +
  scale_fill_viridis_c(option = "rocket") +
  guides(
    x = guide_axis(angle = 90),
    fill = guide_colorbar(keywidth = 0.4, keyheight = 3)
  ) +
  # scale_y_discrete(expand = c(0,0)) +
  scale_y_continuous(
    expand = c(0, 0), breaks = 1:n_distinct(mapk_exp_df_long$Gene),
    labels = rev(unique(mapk_exp_df_long$Gene)),
    sec.axis = sec_axis(~.,
      breaks = 1:n_distinct(mapk_exp_df_long$Gene),
      labels = rev(p_df$p.signif[match(feats, p_df$Gene)])
    )
  ) +
  scale_x_discrete(expand = c(0, 0)) +
  theme_few() +
  theme(
    strip.text.x = element_blank(), strip.background.x = element_blank(), panel.border = element_blank(),
    axis.ticks.y.right = element_blank(), axis.text.y.right = element_text(vjust = 0.8, hjust = 0)
  ) +
  facet_grid(. ~ secondary_mapk, scales = "free", space = "free") +
  labs(x = "Sample", y = "Gene", fill = "Normalized\nExpression")
p4
```



```{r, fig.width = 6, fig.height = 5.5}
pcomb <- (p1 + theme(plot.margin = margin())) + free(p2, type = "space", side = "b") + free((p3 + theme(plot.margin = margin(), axis.text.x = element_blank(), axis.ticks.x = element_blank(), axis.title.x = element_blank())), type = "space") + plot_spacer() + free(p4, type = "space") + plot_spacer() + plot_layout(ncol = 2, widths = c(1, 0.1), heights = c(1, 1, 0.6)) +
  plot_annotation(theme = theme(plot.margin = margin(3, 8, 15, 12, unit = "mm")))
ggsave(pcomb, file = here("plots", "mapk_alt_luad_sig_relapsed_plusmuts_plusexp_v2.pdf"), width = 6, height = 5.5)

pcomb
```

Do a model with MAPKalt vs not for the three histologies, then a model for each mutation's association with LUSC and SCLC

```{r}
# library(lme4)
library(tidyr)
library(broom.mixed)

dat <- srt_tum@meta.data %>%
  filter(time_point == "PD") %>%
  group_by(sample_id, secondary_mapk, time_point) %>%
  summarize(across(c(LUAD, LUSC, SCLC), mean))
dat$secondary_mapk <- factor(dat$secondary_mapk, levels = c("Unknown", "MAPKalt"))

mods <- map(c("LUAD", "LUSC", "SCLC"), .f = function(pth) {
  mod <- lm(glue::glue("{pth} ~ secondary_mapk"), dat) %>%
    broom::tidy(conf.int = TRUE) %>%
    mutate(Signature = pth, .before = 1) %>%
    filter(term == "secondary_mapkMAPKalt")
}) %>%
  list_rbind()

mods <- mods %>%
  rstatix::add_significance(
    p.col = "p.value",
    cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 0.1, 1),
    symbols = c("p<0.0001", "p<0.001", "p<0.01", "p<0.05", "p<0.1", "ns")
  )
```


```{r, fig.width = 2, fig.height = 1.5}
p <- mods %>%
  ggplot(., aes(x = estimate, y = fct_reorder(Signature, estimate, mean))) +
  geom_pointrange(aes(x = estimate, xmin = conf.low, xmax = conf.high), fatten = 0.4) +
  geom_vline(xintercept = 0, linetype = "11") +
  theme_few() +
  theme(
    axis.line = element_blank(),
    panel.border = element_rect(fill = NA, color = "black"),
    strip.background = element_blank(),
    plot.title = element_text(size = 10, hjust = 0.5),
    axis.title = element_text(size = 8)
  ) +
  labs(y = "Gene Signature", x = "Coefficient", title = "Secondary MAPK (PD)") +
  geom_text(data = {
    filter(mods, p.value != "ns")
  }, aes(y = Signature, x = estimate, label = p.value.signif), nudge_y = 0.325, size = 2.25)
ggsave(p, file = here("plots", "acquired_mapk_lm_v2.pdf"), width = 2, height = 1.5)
p
```


### Signature Heatmap

```{r}
feats <- c("LUAD", "LUSC", "SCLC")
sig_exp_df <- bulk_exp(srt_tum, cell_type_col = "is_tumor_cell", sample_id_col = "sample_id_new", meta_cols = c("time_point", "secondary_mapk", "histology_predominant_short"), features = feats)

sig_exp_df <- sig_exp_df %>%
  # filter(time_point == "PD") %>%
  mutate(across(all_of(feats), ~ scale(.x)))


samp_ord2 <- fct_reorder(sig_exp_df$sample_id_new, -sig_exp_df$LUAD, median) %>%
  levels()

sig_exp_df_long <- sig_exp_df %>%
  pivot_longer(cols = feats, names_to = "Gene", values_to = "exp")

p0 <- sig_exp_df %>%
  ggplot(aes(x = fct_relevel(sample_id_new, samp_ord2))) +
  geom_tile(aes(y = "MAPKalt", fill = secondary_mapk), show.legend = F, color = "black") +
  scale_fill_manual(
    values = c("MAPKalt" = "black", "Unknown" = "white"), breaks = c("MAPKalt"),
    labels = c("Secondary MAPK"), na.value = "white"
  ) +
  facet_grid(. ~ time_point, scales = "free", space = "free") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  theme_few() +
  labs(y = NULL, x = NULL, fill = NULL) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

p1 <- sig_exp_df_long %>%
  ggplot(aes(x = fct_relevel(sample_id_new, samp_ord2), y = fct_relevel(Gene, "SCLC", "LUSC", "LUAD"))) +
  geom_tile(aes(fill = exp), color = "black") +
  # scale_fill_viridis_c(option = "rocket", oob = scales::squish, limits = c(-2, 2)) +
  scale_fill_gradient2(low = scales::muted("blue"), mid = "white", high = scales::muted("red"), oob = scales::squish, limits = c(-2, 2)) +
  facet_grid(. ~ time_point, scales = "free", space = "free") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(fill = "Normalized\nExpression", x = "Sample", y = "Lineage\nSignature") +
  guides(
    x = guide_axis(angle = 90),
    fill = guide_colorbar(keywidth = 0.4, keyheight = 3)
  ) +
  theme_few() +
  theme(strip.text.x = element_blank())
```


```{r, fig.width = 9, fig.height = 2}
pcomb <- p0 + free(p1, type = "label", side = "l") + plot_layout(ncol = 1, heights = c(0.1, 1)) & theme(plot.margin = margin())
pcomb
# ggsave(pcomb, file = here("plots", "sig_heatmap.pdf"), width = 9, height = 2)
```


#### Vertical

```{r, fig.width = 4.5, fig.height = 10}
library(ggpattern)

sig_exp_df$secondary_mapk[sig_exp_df$sample_id_new %in% db$metadata$sample_id_new[!db$metadata$has_impact]] <- NA
p0 <- sig_exp_df %>%
  mutate(secondary_mapk = ifelse(time_point %in% c("TN", "MRD"), "NA", secondary_mapk)) %>%
  ggplot(aes(x = fct_relevel(sample_id_new, rev(samp_ord2)))) +
  geom_tile(aes(y = "MAPKalt", fill = secondary_mapk), show.legend = T, color = "black") +
  # geom_point(data = sig_exp_df[sig_exp_df$time_point != "PD", ], aes(y = "MAPKalt"), pch = 4, size = 5) +
  # geom_tile_pattern(aes(y = "MAPKalt", fill = "white", pattern = "stripe", color = "black"), show.legend = F) +
  scale_fill_manual(
    name = "Secondary MAPK\nAlteration",
    values = c("MAPKalt" = "black", "Unknown" = "white", "NA" = "grey60"), breaks = c("MAPKalt", "Unknown", "NA"),
    labels = c("Detected", "Undetected", "Not applicable")
  ) +
  guides(fill = guide_legend(keywidth = 0.4, keyheight = 0.5, order = 1)) +
  ggnewscale::new_scale_fill() +
  geom_tile(aes(y = "Clinical Histology", fill = histology_predominant_short), color = "black") +
  scale_fill_manual(
    name = "Clinical Histology", values = colors$histology_predominant_short,
    breaks = names(colors$histology_predominant_short)
  ) +
  guides(fill = guide_legend(keywidth = 0.4, keyheight = 0.5, order = 2)) +
  facet_grid(time_point ~ ., scales = "free", space = "free") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0), limits = c("MAPKalt", "Clinical Histology")) +
  guides(x = guide_axis(angle = 90)) +
  theme_few() +
  labs(y = NULL, x = NULL, fill = NULL) +
  theme(
    axis.text.y = element_blank(), axis.ticks.y = element_blank(),
    strip.text.y.right = element_text(angle = 0, hjust = 0)
  ) +
  coord_flip()

p1 <- sig_exp_df_long %>%
  ggplot(aes(x = fct_relevel(sample_id_new, rev(samp_ord2)), y = fct_relevel(Gene, "LUAD", "LUSC", "SCLC"))) +
  geom_tile(aes(fill = exp), color = "black") +
  # scale_fill_viridis_c(option = "rocket", oob = scales::squish, limits = c(-2, 2)) +
  scale_fill_gradient2(low = scales::muted("blue"), mid = "white", high = scales::muted("red"), oob = scales::squish, limits = c(-2, 2)) +
  facet_grid(time_point ~ ., scales = "free", space = "free") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(fill = "Normalized\nExpression", x = "Samples ranked by LUAD signature", y = "Lineage\nSignature") +
  guides(
    x = guide_axis(angle = 90),
    fill = guide_colorbar(keywidth = 0.4, keyheight = 3)
  ) +
  theme_few() +
  theme(strip.text.y = element_blank()) +
  coord_flip()


pcomb <- free((p1 + theme(plot.margin = margin(4, 0, 0, 30, "mm"))), type = "label", side = "b") + (p0 + theme(plot.margin = margin(4, 30, 0, 0, "mm"))) + plot_layout(ncol = 2, widths = c(1, 0.66), guides = "collect") + plot_annotation(theme = theme(legend.position = "bottom", legend.box.margin = margin(), legend.margin = margin()))

pcomb
ggsave(pcomb, file = here("plots", "sig_heatmap_vertical.pdf"), width = 4.5, height = 10)
```


### Ternary diagram

```{r, fig.width = 5.4, fig.height = 3.4}
if (FALSE) {

  sig_exp_df <- bulk_exp(srt_tum, cell_type_col = "is_tumor_cell", sample_id_col = "sample_id_new", meta_cols = c("time_point", "secondary_mapk"), features = c("LUAD", "LUSC", "SCLC"))

  devtools::load_all()
  sig_exp_df <- sig_exp_df %>%
    # filter(time_point == "PD") %>%
    mutate(across(all_of(c("LUAD", "LUSC", "SCLC")), ~ standardize(.x)))

  sig_exp_df$secondary_mapk[is.na(sig_exp_df$secondary_mapk)] <- "Unknown"
  # sig_exp_df$secondary_mapk[sig_exp_df$time == "Pre"] <- "N/A"

  library(ggtern)

  p1 <- sig_exp_df %>%
    mutate(acquired_mapk = ifelse(secondary_mapk == "MAPKalt", "MAPKalt", "Unknown")) %>%
    ggtern(aes(y = LUAD, x = LUSC, z = SCLC)) +
    geom_mask() +
    geom_point(aes(fill = time_point, shape = acquired_mapk), size = 3, show.legend = T) +
    scale_fill_manual(values = colors$time_point) +
    scale_shape_manual(values = c(21, 24)) +
    theme_classic() +
    Tlab("LUAD") + Llab("LUSC") + Rlab("SCLC") +
    theme_nolabels() +
    # theme_nomask() +
    guides(
      fill = guide_legend(override.aes = list(shape = 21)),
      shape = guide_legend(override.aes = list(fill = "black"))
    ) +
    labs(fill = "Timepoint", shape = "Acquired MAPK") +
    theme_noticks() +
    theme(panel.background = element_blank(), panel.border = element_blank()) +
    # facet_grid(.~time_point, switch = "x") +
    theme(
      # axis.title = element_markdown(color = "blue"), # This is where to do it
      strip.background = element_part_rect(side = ""),
      legend.position = "right",
      strip.text.x.bottom = element_text(size = 12, face = "bold")
    )
  ggsave(p1, file = here("plots", "sample_ternary_mapk.pdf"), width = 5.4, height = 3.4)
  p1
  detach("package:ggtern", unload = TRUE)
}
```
